---
title: "3mer_motifs"
author: "Andy Beck"
date: "2025-03-03"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Libraries and Data

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(ggpubfigs)
library(reactable)
```

### Directories

```{r}
base_dir <- "output"
motif_dir <- paste0(base_dir, "/3mer_counts")
```

### Load Results

```{r}
load_motif <- function(motif_dir, pop, subtype){
  f_name <- paste0(motif_dir, "/", pop, "_", subtype, ".csv")
  df <- read_csv(f_name, show_col_types = FALSE) %>%
    mutate(p_s = singletons / sum(singletons),
           p_c = controls / sum(controls))
  return(df)
}
```

### Functions for motif counting

Here we'll generate functions for getting motifs for singletons and their matched controls on the fly. For most of the document we will rely on pre-generated results.

```{r}
library(reticulate)
#conda_python("base", conda = "/net/snowwhite/home/beckandy/mambaforge/bin/mamba")
#conda_create("my_project_env", conda = "/net/snowwhite/home/beckandy/mambaforge/bin/mamba", python_version = 3.12)
#py_install(packages = c("pandas","ray"))
conda_python("my_project_env", conda = "/net/snowwhite/home/beckandy/mambaforge/bin/mamba")
source_python("scratch/3mer_enrichment_R.py")

#get_motif_table("AT_CG", "ALL", "output", as.integer(-2:2))

motif_df <- function(subtype, population, rel_vals = c(-1:1)){
  rel_vals <- sort(rel_vals)
  df <- try(get_motif_table(subtype, population, "output", rel_vals = as.integer(rel_vals)))
  if (inherits(df, 'try-error')){
    py_eval('ray.shutdown()')
  }
  return(df)
}

ray_start <- function(){
  py_eval('ray.init(num_cpus=22)')
}

ray_shutdown <- function(){
  py_eval('ray.shutdown()')
}

ray_start()

#motif_df("AT_CG", "ALL", -1:1)

#motif_df("AT_CG", "AMR", rel_vals = c(-1,0,1))
```

### Plotting and Other Functions

```{r}
subtype_print_names <- function(st){
  if(str_starts(st, "AT")){
    return(paste0("A → ", str_sub(st, 4, 4)))
  } else if(str_starts(st, "GC")){
    return(paste0("C → ", str_sub(st, 5, 5)))
  } else{
    return(paste0("CpG → ", str_sub(st, 9, 9), "pG"))
  }
}

motif_prop_plot <- function(df, subtype, motif = as.character("(-1, 1)")){
  p <- df  %>% 
    select(motif, p_s, p_c) %>%
    rename(Singletons = p_s, Controls = p_c) %>%
    pivot_longer(-motif, names_to = "category", values_to = "proportion") %>%
    ggplot(aes(x = proportion, y = motif, shape = category, color = category)) +
    geom_point(size=3) +
    xlab("Class Proportion") +
    ylab("Motif") +
    theme_minimal() +
    scale_color_manual(values = ggpubfigs::friendly_pals$wong_eight) +
    labs(color = "Category", shape = "Category") + 
    theme(legend.position = "inside", legend.position.inside = c(0.85,0.85),
          legend.background = element_rect(color = "white", fill = "white"),
          legend.box.background = element_rect(fill = alpha('#e5e5e5', .6), 
                                               linewidth = 1, 
                                               linetype = "solid", color = "#333333")) +
    ggtitle(subtype_print_names(subtype), motif)
  return(p)
}
```


## Analyses

### ALL Populations

```{r}
population <- "ALL"
```

#### A > G

```{r}
subtype <- "AT_GC"

df <- load_motif(motif_dir, population, subtype)

df %>%
  reactable(pagination = F)
```

Let's assess how similar/dissimilar these distributions are between the singletons and their matched controls:

First: do the rankings of motifs match?

```{r}
motif_rank_singletons <- df %>%
  arrange(desc(p_s)) %>% 
  pull(motif)

motif_rank_controls <- df %>%
  arrange(desc(p_c)) %>% 
  pull(motif)

motif_rank_singletons
motif_rank_controls
motif_rank_singletons == motif_rank_controls
```

Not really much at all!

```{r}
df %>%
  ggplot(aes(x = p_s, y = p_c)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = 2) +
  xlab("Proportion Singletons") +
  ylab("Proportion Controls") +
  ggtitle(subtype_print_names(subtype))

motif_prop_plot(df, subtype)
```

```{r}
rel_val <- sum(df$p_s * log(df$p_s / df$p_c))

n_singletons <- sum(df$singletons)
exp_s <- n_singletons * df$p_c
chi_sq <- sum(((df$singletons - exp_s)^2) / exp_s)
```

I don't think we're going to do any better than (-1, +1), but let's see what we get with two other motifs: (-2, +1) and (-2, -1):

```{r}
df2 <- motif_df("AT_GC", "ALL", -2:-1) %>%
  mutate(p_s = singletons / sum(singletons),
         p_c = controls / sum(controls))

df3 <- motif_df("AT_GC", "ALL", c(-2,1)) %>%
  mutate(p_s = singletons / sum(singletons),
         p_c = controls / sum(controls))

sum(df2$p_s * log(df2$p_s / df2$p_c))
sum(df3$p_s * log(df3$p_s / df3$p_c))

df2$exp_s <- sum(df2$singletons) * df2$p_c
df3$exp_s <- sum(df3$singletons) * df3$p_c

sum(((df2$singletons - df2$exp_s)^2) / df2$exp_s)
sum(((df3$singletons - df3$exp_s)^2) / df3$exp_s)

motif_prop_plot(df3, subtype, "(-2, 1)")
```

Based on this, it would seem that a 5-mer motif might be necessary for this subtype.

#### A > C

```{r}
subtype <- "AT_CG"

df <- load_motif(motif_dir, population, subtype)

df %>%
  reactable(pagination = F)
```

Let's assess how similar/dissimilar these distributions are between the singletons and their matched controls:

First: do the rankings of motifs match?

```{r}
motif_rank_singletons <- df %>%
  arrange(desc(p_s)) %>% 
  pull(motif)

motif_rank_controls <- df %>%
  arrange(desc(p_c)) %>% 
  pull(motif)

motif_rank_singletons
motif_rank_controls
motif_rank_singletons == motif_rank_controls
```

The answer is: no! Let's visualize this in a scatterplot:

```{r}
df %>%
  ggplot(aes(x = p_s, y = p_c)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = 2)

motif_prop_plot(df, subtype)
```


Next, let's look at a few metrics of density similarity/dissimilarity:

```{r}
# compute relative entropy
re_val <- sum(df$p_s * log(df$p_s / df$p_c))

print(re_val)
```

Relative to the values we observe for, say, our single position models, the relative entropy we observe here is quite high; however, I'm still unsure how to really say anything about the strength of this difference between the distributions. One thought is how about we draw random re-samples of the singletons using their observed proportion and then get a "null distribution" of relative entropy values from these re-samples?

```{r}
n_singletons <- df$singletons %>% sum()

s_1 <- rmultinom(1, n_singletons, df$p_s)
s_1 <- s_1 / sum(s_1)

sum(df$p_s * log(df$p_s / s_1))
```

Maybe this isn't quite what I'm looking for; maybe we want to draw the same number of singletons, but from the proportions from the control distribution?

```{r}
re_vals <- c()

for(i in 1:1000){
  s_1 <- rmultinom(1, n_singletons, df$p_c)
  s_1 <- s_1 / sum(s_1)

  re_vals <- c(re_vals, sum(s_1 * log(s_1 / df$p_c)))
}

data.frame(re = re_vals) %>%
  ggplot(aes(x = re)) +
  geom_density() +
  xlab("Relative Entropy") +
  theme_classic()
```

Ok, so empirically, our relative entropy value of `r re_val` is far larger than what we'd expect if our singletons were a random draw from a distribution with the same proportions as observed in our control sample. Great! 

What might be more interesting to look at is the relative entropy statistics that we might have observed for motifs other than the central 3-mer. For example, what if we defined the motif by the nucleotides at the -2 and -1 positions? Do we see a larger relative entropy value for such a motif?

```{r}
df2 <- motif_df("AT_CG", "ALL", -2:0) %>%
  mutate(p_s = singletons / sum(singletons),
         p_c = controls / sum(controls))

df3 <- motif_df("AT_CG", "ALL", c(-1, 2)) %>%
  mutate(p_s = singletons / sum(singletons),
         p_c = controls / sum(controls))

sum(df2$p_s * log(df2$p_s / df2$p_c))
sum(df3$p_s * log(df3$p_s / df3$p_c))

motif_prop_plot(df2, subtype, "(-2, -1)")
motif_prop_plot(df3, subtype, "(-1, 2)")
```

```{r}
df$exp_s <- sum(df$singletons) * df$p_c
df2$exp_s <- sum(df2$singletons) * df2$p_c
df3$exp_s <- sum(df3$singletons) * df3$p_c

sum(((df$singletons - df$exp_s)^2) / df$exp_s)
sum(((df2$singletons - df2$exp_s)^2) / df2$exp_s)
sum(((df3$singletons - df3$exp_s)^2) / df3$exp_s)
```


Here we observe a large re value relative to our "null" draws from above, and it's close to what we observed for the -1,1 motif

* The top single position effects were at: -1, -3, 2, +1
* The top two-way interaction was at (1,2)
    * (-1, 1) was the second highest 2-position relative entropy

## Cleanup

```{r}
ray_shutdown()
```

