---
title: "control_control"
author: "Andy Beck"
date: "2025-02-05"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

## Setup and Functions

### Libraries 

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(reactable)
```

### Directories and Other Variables

```{r}
base_dir <- "output/single_pos_cc" # template: {base_dir}/{pop}/{subtype}.csv or {subtype}_v2.csv

subtypes <- c("AT_CG", "AT_GC", "AT_TA",
            "GC_AT", "GC_TA", "GC_CG",
            "cpg_GC_AT", "cpg_GC_TA", "cpg_GC_CG")
```

### Functions

```{r}
load_res <- function(base_dir, population, subtype){
  f_name <- paste0(base_dir, "/", population, "/", subtype, ".csv")
  return(read_csv(f_name, show_col_types = FALSE) %>%
           mutate(re = dev / (2 * (singletons + controls) ),
                  sig = dev > qchisq(0.95, 3)))
}

subtype_print_names <- function(st){
  if(str_starts(st, "AT")){
    return(paste0("A → ", str_sub(st, 4, 4)))
  } else if(str_starts(st, "GC")){
    return(paste0("C → ", str_sub(st, 5, 5)))
  } else{
    return(paste0("CpG → ", str_sub(st, 9, 9), "pG"))
  }
}

plot_res <- function(pop, subtype, base_dir){
  df <- load_res(base_dir, pop, subtype)
  
  p <- df %>%
    ggplot(aes(x = offset, y = re, colour = sig)) +
    geom_point() +
    theme_bw() +
    xlab(paste0("Relative Position: ", subtype_print_names(subtype))) +
    ylab("Relative Entropy") +
    theme(text = element_text(family = "Helvetica", size = 12)) +
    guides(color = guide_legend(title = "Significant"))
  return(p)
}
```

## Closest versus Furthest

These results are based on comparing the distibutions of singletons flanking the controls sampled nearest to the singleton to the distributions of nucleotides flanking the furthest sampled control.

```{r}
pop <- "ALL"
subtype <- "AT_CG"
df <- load_res(base_dir, pop, subtype)

df %>%
  ggplot(aes(x = offset, y = re, colour = sig)) +
  geom_point() +
  theme_bw() +
  xlab(paste0("Relative Position: ", subtype_print_names(subtype))) +
  ylab("Relative Entropy") +
  theme(text = element_text(family = "Helvetica", size = 12),
        legend.position = "inside",
        legend.position.inside = c(0.85, 0.85)) +
  guides(color = guide_legend(title = "Significant"))

```

Well, this is most certainly an interesting result! Let's see what the results look like when we use randomly sampled controls:

```{r}
pop <- "ALL"
subtype <- "AT_GC_v2"
df <- load_res(base_dir, pop, subtype)

df %>%
  ggplot(aes(x = offset, y = re, colour = sig)) +
  geom_point() +
  theme_bw() +
  xlab(paste0("Relative Position: ", subtype_print_names(subtype))) +
  ylab("Relative Entropy") +
  theme(text = element_text(family = "Helvetica", size = 12),
        legend.position = "inside",
        legend.position.inside = c(0.85, 0.85)) +
  guides(color = guide_legend(title = "Significant"))

df %>%
  ggplot(aes(x = dev)) +
  geom_density() +
  stat_function(fun = dchisq, args = c(3), n = 1000, colour = "red") + 
  geom_vline(xintercept = 1, colour = "red", linetype = 2) + 
  geom_vline(xintercept = qchisq(0.95, 3), colour = "blue", linetype = 3) +
  xlab(paste0("Deviance: ", subtype_print_names(substr(subtype,1,nchar(subtype)-3) ))) +
  annotate("text", 
           x = 11, 
           y = 0.05, 
           label = paste0("FPR: ", sum(df$dev > qchisq(0.95,3)) / length(df$dev)) )
```

This looks much more like what we're expecting! But anyways, let's make these figures for all population/subtype pairs.

### ALL

```{r results='asis', echo=FALSE}

pop <- "ALL"
for(st in subtypes){
  cat(paste0('\n#### ', st, "\n"))
  cat('#### Near v Far \n')
  print(plot_res(pop, st, base_dir))
  cat('\n#### Random Controls\n')
  print(plot_res(pop, paste0(st, "_v2"), base_dir))
  df <- load_res(base_dir, pop, paste0(st, "_v2"))
  p <- df %>%
    ggplot(aes(x = dev)) +
    geom_density() +
    stat_function(fun = dchisq, args = c(3), n = 1000, colour = "red") + 
    geom_vline(xintercept = 1, colour = "red", linetype = 2) +
    xlab(paste0("Deviance: ", subtype_print_names( st ) )) +
    annotate("text", 
           x = 11, 
           y = 0.05, 
           label = paste0("FPR: ", sum(df$dev > qchisq(0.95,3)) / length(df$dev)) )
  print(p)
}

```

### AFR

```{r results='asis', echo=FALSE}
pop <- "AFR"
for(st in subtypes){
  cat(paste0('\n#### ', st, "\n"))
  cat('#### Near v Far \n')
  print(plot_res(pop, st, base_dir))
  cat('\n#### Random Controls\n')
  print(plot_res(pop, paste0(st, "_v2"), base_dir))
  df <- load_res(base_dir, pop, paste0(st, "_v2"))
  p <- df %>%
    ggplot(aes(x = dev)) +
    geom_density() +
    stat_function(fun = dchisq, args = c(3), n = 1000, colour = "red") + 
    geom_vline(xintercept = 1, colour = "red", linetype = 2) +
    xlab(paste0("Deviance: ", subtype_print_names( st ) )) +
    annotate("text", 
           x = 11, 
           y = 0.05, 
           label = paste0("FPR: ", sum(df$dev > qchisq(0.95,3)) / length(df$dev)) )
  print(p)
}
```

### AMR

```{r results='asis', echo=FALSE}
pop <- "AMR"
for(st in subtypes){
  cat(paste0('\n#### ', st, "\n"))
  cat('#### Near v Far \n')
  print(plot_res(pop, st, base_dir))
  cat('\n#### Random Controls\n')
  print(plot_res(pop, paste0(st, "_v2"), base_dir))
  df <- load_res(base_dir, pop, paste0(st, "_v2"))
  p <- df %>%
    ggplot(aes(x = dev)) +
    geom_density() +
    stat_function(fun = dchisq, args = c(3), n = 1000, colour = "red") + 
    geom_vline(xintercept = 1, colour = "red", linetype = 2) +
    xlab(paste0("Deviance: ", subtype_print_names( st ) )) +
    annotate("text", 
           x = 11, 
           y = 0.05, 
           label = paste0("FPR: ", sum(df$dev > qchisq(0.95,3)) / length(df$dev)) )
  print(p)
}
```

### EAS

```{r results='asis', echo=FALSE}
pop <- "EAS"
for(st in subtypes){
  cat(paste0('\n#### ', st, "\n"))
  cat('#### Near v Far \n')
  print(plot_res(pop, st, base_dir))
  cat('\n#### Random Controls\n')
  print(plot_res(pop, paste0(st, "_v2"), base_dir))
  df <- load_res(base_dir, pop, paste0(st, "_v2"))
  p <- df %>%
    ggplot(aes(x = dev)) +
    geom_density() +
    stat_function(fun = dchisq, args = c(3), n = 1000, colour = "red") + 
    geom_vline(xintercept = 1, colour = "red", linetype = 2) +
    xlab(paste0("Deviance: ", subtype_print_names( st ) )) +
    annotate("text", 
           x = 11, 
           y = 0.05, 
           label = paste0("FPR: ", sum(df$dev > qchisq(0.95,3)) / length(df$dev)) )
  print(p)
}
```

### EUR

```{r results='asis', echo=FALSE}
pop <- "EUR"
for(st in subtypes){
  cat(paste0('\n#### ', st, "\n"))
  cat('#### Near v Far \n')
  print(plot_res(pop, st, base_dir))
  cat('\n#### Random Controls\n')
  print(plot_res(pop, paste0(st, "_v2"), base_dir))
  df <- load_res(base_dir, pop, paste0(st, "_v2"))
  p <- df %>%
    ggplot(aes(x = dev)) +
    geom_density() +
    stat_function(fun = dchisq, args = c(3), n = 1000, colour = "red") + 
    geom_vline(xintercept = 1, colour = "red", linetype = 2) +
    xlab(paste0("Deviance: ", subtype_print_names( st ) )) +
    annotate("text", 
           x = 11, 
           y = 0.05, 
           label = paste0("FPR: ", sum(df$dev > qchisq(0.95,3)) / length(df$dev)) )
  print(p)
}
```

### SAS

```{r results='asis', echo=FALSE}
pop <- "SAS"
for(st in subtypes){
  cat(paste0('\n#### ', st, "\n"))
  cat('#### Near v Far \n')
  print(plot_res(pop, st, base_dir))
  cat('\n#### Random Controls\n')
  print(plot_res(pop, paste0(st, "_v2"), base_dir))
  df <- load_res(base_dir, pop, paste0(st, "_v2"))
  p <- df %>%
    ggplot(aes(x = dev)) +
    geom_density() +
    stat_function(fun = dchisq, args = c(3), n = 1000, colour = "red") + 
    geom_vline(xintercept = 1, colour = "red", linetype = 2) +
    xlab(paste0("Deviance: ", subtype_print_names( st ) )) +
    annotate("text", 
           x = 11, 
           y = 0.05, 
           label = paste0("FPR: ", sum(df$dev > qchisq(0.95,3)) / length(df$dev)) )
  print(p)
}
```


