---
title: "interaction_parameters"
author: "theandyb"
date: "2025-05-13"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(broom)
library(reticulate)
library(conflicted)

conda_python("my_project_env", conda = "/net/snowwhite/home/beckandy/mambaforge/bin/mamba")
source_python("scratch/3mer_enrichment_R.py")
```

## Data Functions

```{r}
#' Load counts of singletons and controls stratified by nucleotides at given flanking positions
#' @param subtype mutation subtype
#' @param population 1kGP super population (ALL for all all five)
#' @param rel_values vector of relative positions to stratify singleton and control counts by
motif_df <- function(subtype, population, rel_vals = c(-1:1)){
  rel_vals <- sort(rel_vals)
  df <- try(get_motif_table(subtype, population, "output", rel_vals = as.integer(rel_vals)))
  if (inherits(df, 'try-error')){
    py_eval('ray.shutdown()')
  }
  return(df)
}

#' Function to start underlying ray instance
#' If not run before motif_df, an instance using all cores will be started, which might not be desirable
ray_start <- function(){
  py_eval('ray.init(num_cpus=22)')
}

#' Function to stop underlying ray instance
#' Not sure what happens if this isn't run when R session stops ¯\_(ツ)_/¯
ray_shutdown <- function(){
  py_eval('ray.shutdown()')
}

```

## Single Subtype Example

Let's look at our strongest example: the +1 relative position for the A>G subtypes:

```{r}
ray_start()

df_all <- motif_df("AT_GC", "ALL", c(0,1))
df_afr <- motif_df("AT_GC", "AFR", c(0,1))
df_amr <- motif_df("AT_GC", "AMR", c(0,1))
df_eas <- motif_df("AT_GC", "EAS", c(0,1))
df_eur <- motif_df("AT_GC", "EUR", c(0,1))
df_sas <- motif_df("AT_GC", "SAS", c(0,1))

ray_shutdown()
```

Let's reformat these data frames for log linear modeling!

```{r}
single_pos_mod <- function(df, nuc_pos){
  df %>%
    mutate(nuc = str_sub(motif, nuc_pos, nuc_pos)) %>%
    select(nuc, singletons, controls) %>%
    pivot_longer(-nuc, names_to = "status", values_to = "n")
}

fit_sat_mod <- function(df, nuc_pos){
  df_mod <- single_pos_mod(df, nuc_pos)
  mod_obj <- glm(n ~ status*nuc, data = df_mod, family = poisson)
  return(mod_obj)
}

```

### Parameter Comparisons

```{r}
mod_all <- fit_sat_mod(df_all, 2)
mod_afr <- fit_sat_mod(df_afr, 2)
mod_amr <- fit_sat_mod(df_amr, 2)
mod_eas <- fit_sat_mod(df_eas, 2)
mod_eur <- fit_sat_mod(df_eur, 2)
mod_sas <- fit_sat_mod(df_sas, 2)

# tidy dataframes
df_mod_all <- broom::tidy(mod_all) %>% rename(all = estimate) %>% select(term, all)
df_mod_afr <- broom::tidy(mod_afr) %>% rename(afr = estimate) %>% select(term, afr)
df_mod_amr <- broom::tidy(mod_amr) %>% rename(amr = estimate) %>% select(term, amr)
df_mod_eas <- broom::tidy(mod_eas) %>% rename(eas = estimate) %>% select(term, eas)
df_mod_eur <- broom::tidy(mod_eur) %>% rename(eur = estimate) %>% select(term, eur)
df_mod_sas <- broom::tidy(mod_sas) %>% rename(sas = estimate) %>% select(term, sas)

# merge together

df_mod <- inner_join(df_mod_all, df_mod_afr, by = "term") %>%
  inner_join(df_mod_amr) %>%
  inner_join(df_mod_eas) %>%
  inner_join(df_mod_eur) %>%
  inner_join(df_mod_sas)
```

### Group Parameter

Let's compare models with the same interaction among populations versus models that allow for population specific interactions

```{r}
df_comb <- single_pos_mod(df_all, 2) %>% mutate(pop = "ALL") %>%
  bind_rows({single_pos_mod(df_afr, 2) %>% mutate(pop = "AFR")}) %>%
  bind_rows({single_pos_mod(df_amr, 2) %>% mutate(pop = "AMR")}) %>%
  bind_rows({single_pos_mod(df_eas, 2) %>% mutate(pop = "EAS")}) %>%
  bind_rows({single_pos_mod(df_eur, 2) %>% mutate(pop = "EUR")}) %>%
  bind_rows({single_pos_mod(df_sas, 2) %>% mutate(pop = "SAS")})

mod_full <- glm(n ~ nuc * status * pop, data = df_comb, family = poisson)
mod_red <- glm(n ~ nuc * status + pop * nuc + pop * status, data = df_comb, family = poisson)

anova(mod_red, mod_full)
```

#### Group Parameter Function

```{r}
group_anova <- function(subtype, rp){
  ray_start()
  
  df_all <- motif_df(subtype, "ALL", c(0,rp))
  df_afr <- motif_df(subtype, "AFR", c(0,rp))
  df_amr <- motif_df(subtype, "AMR", c(0,rp))
  df_eas <- motif_df(subtype, "EAS", c(0,rp))
  df_eur <- motif_df(subtype, "EUR", c(0,rp))
  df_sas <- motif_df(subtype, "SAS", c(0,rp))
  
  ray_shutdown()
  
  df_comb <- single_pos_mod(df_all, 2) %>% mutate(pop = "ALL") %>%
    bind_rows({single_pos_mod(df_afr, 2) %>% mutate(pop = "AFR")}) %>%
    bind_rows({single_pos_mod(df_amr, 2) %>% mutate(pop = "AMR")}) %>%
    bind_rows({single_pos_mod(df_eas, 2) %>% mutate(pop = "EAS")}) %>%
    bind_rows({single_pos_mod(df_eur, 2) %>% mutate(pop = "EUR")}) %>%
    bind_rows({single_pos_mod(df_sas, 2) %>% mutate(pop = "SAS")})
  
  mod_full <- glm(n ~ nuc * status * pop, data = df_comb, family = poisson)
  mod_red <- glm(n ~ nuc * status + pop * nuc + pop * status, data = df_comb, family = poisson)
  
  return(anova(mod_red, mod_full))
}
```

```{r}
group_anova("AT_GC", 100)
group_anova("AT_GC", 10)
```

## Use already generated residual files

```{r}
read_resid <- function(pop, subtype, rp){
  f_name <- paste0("output/single_pos/resid/", pop, "/", subtype, "_rp_", rp, ".csv")
  read_csv(f_name, show_col_types = F) %>%
    select(nuc, status, n) %>%
    mutate(pop = pop)
}

mod_df <- function(subtype, rp, pops = c("ALL", "AFR", "AMR", "EAS", "EUR", "SAS")){
  
}
```

