---
title: "logistic_models"
author: "Andy Beck"
date: "2025-03-14"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(broom)
library(reticulate)
library(rcompanion) #nagelkerke function
#library(speedglm)

conda_python("my_project_env", conda = "/net/snowwhite/home/beckandy/mambaforge/bin/mamba")
source_python("scratch/3mer_enrichment_R.py")
```


## Introduction

Here in this document we'll explore one particular alternative model structure for the examination of the influence of local sequence context on rates of germline mutation: logistic regression models. But first, some functions to enable loading of data "on the fly":

## Data Functions

```{r}
#' Load counts of singletons and controls stratified by nucleotides at given flanking positions
#' @param subtype mutation subtype
#' @param population 1kGP super population (ALL for all all five)
#' @param rel_values vector of relative positions to stratify singleton and control counts by
motif_df <- function(subtype, population, rel_vals = c(-1:1)){
  rel_vals <- sort(rel_vals)
  df <- try(get_motif_table(subtype, population, "output", rel_vals = as.integer(rel_vals)))
  if (inherits(df, 'try-error')){
    py_eval('ray.shutdown()')
  }
  return(df)
}

#' Function to start underlying ray instance
#' If not run before motif_df, an instance using all cores will be started, which might not be desirable
ray_start <- function(){
  py_eval('ray.init(num_cpus=22)')
}

#' Function to stop underlying ray instance
#' Not sure what happens if this isn't run when R session stops ¯\_(ツ)_/¯
ray_shutdown <- function(){
  py_eval('ray.shutdown()')
}

```

## Conditional Logistic Regression Models

```{r}
library(survival)
```

```{r}
df <- read_csv("scratch/EUR_AG_1.csv") %>%
  filter(nuc %in% c("A", "C", "G", "T"))

mod_clr <- clogit(status ~ nuc + strata(id), df)
```


## Single Position Effects

For our first example, let's compare the results of two models for the influence of the nucleotide at the +1 relative position for the A > G mutation subtype. 

Model 1:

$$
\textrm{logit}(\Pr(X_i \textrm{singleton} | \textrm{nucleotide}_i)) = \beta_0 + \sum_{j \in C,G,T} \beta_j I(\textrm{nucleotide}_i = j)
$$

And model 2:

$$
\log(n_i^s) = \lambda_0 + \lambda_{\textrm{singleton}}I(s=\textrm{singleton}) + \sum_{j \in C,G,T} \lambda_j I(i=j)
$$

Both models are for data which we can represent in the following table:

```{r}
population <- "ALL"
subtype <- "AT_GC"

ray_start()
df <- motif_df(subtype, population, rel_vals = as.vector(c(1)))
ray_shutdown()

df %>% knitr::kable()
```

Let's fit the two models and see what we obtain with regards to parameter estimates and goodness of fit statistics:

```{r}
mod_logit <- glm(cbind(singletons, controls) ~ motif, data = df, family = binomial())

mod_logit2 <- df %>%
  mutate(controls2 = round(controls / 5)) %>%
  glm(cbind(singletons, controls2) ~ motif, data = ., family = binomial())

mod_loglin <- df %>%
  pivot_longer(-motif, names_to = "status", values_to = "n") %>%
  glm(n ~ motif + status, data=., family = poisson())
mod_loglin_sat <- df %>%
  pivot_longer(-motif, names_to = "status", values_to = "n") %>%
  glm(n ~ motif * status, data=., family = poisson())


mod_logit %>%
  tidy() %>%
  knitr::kable()

mod_logit %>%
  glance() %>%
  knitr::kable()

mod_logit2 %>%
  tidy() %>%
  knitr::kable()

mod_logit2 %>%
  glance() %>%
  knitr::kable()

mod_loglin %>%
  tidy() %>%
  knitr::kable()

mod_loglin %>%
  glance() %>%
  knitr::kable()

mod_loglin_sat %>%
  tidy() %>%
  knitr::kable()

mod_loglin_sat %>%
  glance() %>%
  knitr::kable()
```

Let's see if anything interesting happens if we "lengthen" our data.frame and give each "observation" (singleton, control) it's own row (divide controls by 1/5 to make even in number):

```{r}
# df <- df %>%
#   mutate(controls2 = round(controls / 5))
# 
# df_long <- data.frame(status = numeric(0), motif = character(0))
# 
# for(i in 1:length(df$motif)){
#   df_singletons <- data.frame(status = rep(1, df$singletons[i]), motif = df$motif[i])
#   df_controls <- data.frame(status = rep(0, df$controls2[i]), motif = df$motif[i])
#   df_long <- bind_rows(df_long, df_singletons, df_controls)
# }
# 
# df_long[] = lapply(df_long, sample)
# mod_logit3 <- speedglm(status ~ motif, data = df_long, family = binomial())
```

### Extended Window

Let's grab 7-mer motifs surrounding singletons and matched controls for the A>G subtype and play around with models using nucleotides in this extended window. Note that now single-position logistic regression models are no longer fully saturated (?)

```{r}
population <- "EUR"
subtype <- "AT_GC"

ray_start()
df <- motif_df(subtype, population, rel_vals = c(-3:-1, 1:3))
ray_shutdown()

# df$controls <- round(df$controls / 5)

df <- df %>%
  separate(motif, sep=1:6, into = paste0("n", 1:6))

df %>% head(10) %>% knitr::kable()
```

Contrast each individual position using Nagelkerke's R2:

```{r}
mod_n1 <- glm(cbind(singletons, controls) ~ n1, data = df, family = binomial)
mod_n2 <- glm(cbind(singletons, controls) ~ n2, data = df, family = binomial)
mod_n3 <- glm(cbind(singletons, controls) ~ n3, data = df, family = binomial)
mod_n4 <- glm(cbind(singletons, controls) ~ n4, data = df, family = binomial)
mod_n5 <- glm(cbind(singletons, controls) ~ n5, data = df, family = binomial)
mod_n6 <- glm(cbind(singletons, controls) ~ n6, data = df, family = binomial)

df_res <- data.frame(pos = c(-3:-1, 1:3),
                     r2 = c(nagelkerke(mod_n1)$Pseudo[3],
                            nagelkerke(mod_n2)$Pseudo[3],
                            nagelkerke(mod_n3)$Pseudo[3],
                            nagelkerke(mod_n4)$Pseudo[3],
                            nagelkerke(mod_n5)$Pseudo[3],
                            nagelkerke(mod_n6)$Pseudo[3]),
                     aic = c(mod_n1$aic, mod_n2$aic, mod_n3$aic, mod_n4$aic, mod_n5$aic, mod_n6$aic))

df_res %>% knitr::kable()
```

And same for loglinear models:

```{r}
df_long <- df %>% pivot_longer(singletons:controls, names_to = "status", values_to = "n")

mod_n1 <- glm(n ~ status + n1, data = df_long, family = poisson)
mod_n2 <- glm(n ~ status + n2, data = df_long, family = poisson)
mod_n3 <- glm(n ~ status + n3, data = df_long, family = poisson)
mod_n4 <- glm(n ~ status + n4, data = df_long, family = poisson)
mod_n5 <- glm(n ~ status + n5, data = df_long, family = poisson)
mod_n6 <- glm(n ~ status + n6, data = df_long, family = poisson)

df_res <- data.frame(pos = c(-3:-1, 1:3),
                     r2 = c(nagelkerke(mod_n1)$Pseudo[3],
                            nagelkerke(mod_n2)$Pseudo[3],
                            nagelkerke(mod_n3)$Pseudo[3],
                            nagelkerke(mod_n4)$Pseudo[3],
                            nagelkerke(mod_n5)$Pseudo[3],
                            nagelkerke(mod_n6)$Pseudo[3]),
                     aic = c(mod_n1$aic, mod_n2$aic, mod_n3$aic, mod_n4$aic, mod_n5$aic, mod_n6$aic),
                     deviance = c(mod_n1$deviance, mod_n2$deviance, mod_n3$deviance, mod_n4$deviance, mod_n5$deviance, mod_n6$deviance))

df_res %>% knitr::kable()
```


```{r}
df_long %>% 
  select(n4, status, n) %>%
  group_by(status, n4) %>%
  summarize(n = sum(n)) %>%
  glm(n ~ status + n4, family = poisson, data = .) %>%
  deviance()

df_long %>% 
  select(n5, status, n) %>%
  group_by(status, n5) %>%
  summarize(n = sum(n)) %>%
  glm(n ~ status + n5, family = poisson, data = .) %>%
  deviance()
```

```{r}
ray_start()
df_1 <- motif_df(subtype, population, rel_vals = 1)
df_2 <- motif_df(subtype, population, rel_vals = 2)
ray_shutdown()

df_1 %>%
  pivot_longer(singletons:controls, names_to = "status", values_to = "n") %>%
  glm(n ~ motif + status, data = ., family = poisson()) %>%
  deviance()

df_2 %>%
  pivot_longer(singletons:controls, names_to = "status", values_to = "n") %>%
  glm(n ~ motif + status, data = ., family = poisson()) %>%
  deviance()
```

