---
title: "pop_comparison"
author: "Andy Beck"
date: "2025-01-20"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

In this document we are primariliy interested in assessing the consistency of results across the five 1kGP super-populations. Namely, for the single position models:

1. Are the patterns of influence seen across positions shared across populations?
2. At individual positions, are the same nucleotides the most influential across populations?

The questions we ask for the two position models are similar.

## Setup and Data Functions

### Libraries

```{r}
library(tidyverse)
library(ggpubfigs) # functions for color blind friendly palettes
library(reactable) # table output for html
library(knitr) # PDF table output
library(corrr)
```

### Single Position Functions

Load position-level results:

```{r}
sp_deviance_df_load <- function(res_dir, pop, subtype){
  f_name <- paste0(res_dir, "/", pop, "/", subtype, ".csv")
  df <- read_csv(f_name, show_col_types = FALSE) %>%
    rowwise() %>%
    mutate(re = dev / (2 * (singletons+controls)),
           is_sig = dev > qchisq(0.05, df = 3, lower.tail = F))
  return(df)
}

sp_re_all_pop <- function(res_dir, subtype){
  df_ALL <- sp_deviance_df_load(res_dir, "ALL", subtype) %>%
    select(offset, re, is_sig) %>%
    mutate(pop="ALL")
  
  df_AFR <- sp_deviance_df_load(res_dir, "AFR", subtype) %>%
    select(offset, re, is_sig) %>%
    mutate(pop="AFR")
  
  df_AMR <- sp_deviance_df_load(res_dir, "AMR", subtype) %>%
    select(offset, re, is_sig) %>%
    mutate(pop="AMR")
  
  df_EAS <- sp_deviance_df_load(res_dir, "EAS", subtype) %>%
    select(offset, re, is_sig) %>%
    mutate(pop="EAS")
  
  df_EUR <- sp_deviance_df_load(res_dir, "EUR", subtype) %>%
    select(offset, re, is_sig) %>%
    mutate(pop="EUR")
  
  df_SAS <- sp_deviance_df_load(res_dir, "SAS", subtype) %>%
    select(offset, re, is_sig) %>%
    mutate(pop="SAS")
  
  #return(bind_rows(df_ALL, df_AFR, df_AMR, df_EAS, df_EUR, df_SAS) %>%
  #  pivot_wider(id_cols = offset, names_from = pop, values_from = re))
  return(bind_rows(df_ALL, df_AFR, df_AMR, df_EAS, df_EUR, df_SAS))
}
```

Load position-level residuals:

```{r}
sp_resid_df_load <- function(res_dir, pop, subtype, rp){
  f_name <- paste0(res_dir, "/resid/", pop, "/", subtype, "_rp_", rp, ".csv")
  df <- read_csv(f_name, show_col_types = FALSE)
  df$rp <- rp
  return(df)
}

sp_resid_df_load_all <- function(res_dir, pop, subtype, rp = c(-500:-1, 1:500)){
  results <- vector(mode = "list", length = length(rp))
  for(i in 1:length(rp)){
    if(rp[i] == 1 & str_starts(subtype, "cpg")) next
    results[[i]] <- sp_resid_df_load(res_dir, pop, subtype, rp[i])
  }
  df <- bind_rows(results)
  return(df)
}

sp_resid_all_pop <- function(res_dir, subtype, rp = c(-500:-1, 1:500)){
  df_ALL <- sp_resid_df_load_all(res_dir, "ALL", subtype, rp) %>%
    select(rp, nuc, status, res) %>%
    mutate(pop = "ALL")
  
  df_AFR <- sp_resid_df_load_all(res_dir, "AFR", subtype, rp) %>%
    select(rp, nuc, status, res) %>%
    mutate(pop = "AFR")
  
  df_AMR <- sp_resid_df_load_all(res_dir, "AMR", subtype, rp) %>%
    select(rp, nuc, status, res) %>%
    mutate(pop = "AMR")
  
  df_EAS <- sp_resid_df_load_all(res_dir, "EAS", subtype, rp) %>%
    select(rp, nuc, status, res) %>%
    mutate(pop = "EAS")
  
  df_EUR <- sp_resid_df_load_all(res_dir, "EUR", subtype, rp) %>%
    select(rp, nuc, status, res) %>%
    mutate(pop = "EUR")
  
  df_SAS <- sp_resid_df_load_all(res_dir, "SAS", subtype, rp) %>%
    select(rp, nuc, status, res) %>%
    mutate(pop = "SAS")
  
  return(bind_rows(df_ALL, df_AFR, df_AMR, df_EAS, df_EUR, df_SAS))
}
```

### Two position functions

```{r}
tp_deviance_df_load <- function(base_dir, pop, subtype){
  f_name <- paste0(base_dir, "/", pop, "/", subtype, ".csv")
  df <- read_csv(f_name, show_col_types = F, 
                 col_names = c("dev","singletons","controls","rp1","rp2")) %>%
    rowwise() %>%
    mutate(re = dev / (2 * (singletons + controls))) %>%
    ungroup()
  return(df)
}

tp_resid_df_load <- function(base_dir, pop, subtype, p, q){
  f_name <- paste0(base_dir, "/resid/", pop, "/", subtype, "_p", p, "_q", q, ".csv")
  df <- read_csv(f_name, show_col_types = F)
  return(df)
}
```

### Paths

```{r}
base_dir = "output"
single_pos_dir = paste0(base_dir, "/single_pos")
two_pos_dir = paste0(base_dir, "/two_pos")
```

## Single Position Results

### A > C

```{r}
subtype <- "AT_CG"

single_pos_analysis <- function(subtype, rp = c(-10:-1, 1:10)){
  df_res <- sp_re_all_pop(single_pos_dir, subtype)
  df_resid <- sp_resid_all_pop(single_pos_dir, subtype, rp = rp)
  
  p <- df_res %>%
    filter(abs(offset) < 20) %>%
    mutate(pop = factor(pop, levels = c("ALL", "AFR", "AMR", "EAS", "EUR", "SAS"))) %>%
    ggplot(aes(x = offset, y = re, colour = pop)) +
    geom_point() +
    geom_line() +
    scale_colour_manual(values = friendly_pal("vibrant_seven")) +
    theme_classic() +
    labs(colour = "Population") +
    xlab("Relative Position") +
    ylab("Relative Entropy")
  print(p)
  
  p <- df_res %>%
    filter(abs(offset) < 20) %>%
    mutate(pop = factor(pop, levels = c("ALL", "AFR", "AMR", "EAS", "EUR", "SAS"))) %>%
    ggplot(aes(x = offset, y = re, colour = pop)) +
    geom_point() +
    geom_line() +
    scale_colour_manual(values = friendly_pal("vibrant_seven")) +
    theme_classic() +
    labs(colour = "Population") +
    xlab("Relative Position") +
    ylab("Relative Entropy") +
    scale_y_log10()
  print(p)
  
  df_res %>%
    select(-is_sig) %>%
    pivot_wider(id_cols = offset, names_from = pop, values_from = re) %>%
    correlate(method = "pearson", diagonal = 1) %>%
    focus(-offset, mirror = T) %>%
    shave() %>%
    fashion() %>%
    print()
  
  df_res %>%
    select(-is_sig) %>%
    pivot_wider(id_cols = offset, names_from = pop, values_from = re) %>%
    correlate(method = "spearman", diagonal = 1) %>%
    focus(-offset, mirror = T) %>%
    shave() %>%
    fashion() %>%
    print()
  
  df_wide <- df_res %>%
    select(-is_sig) %>%
    pivot_wider(id_cols = offset, names_from = pop, values_from = re)
  
  nms <- combn(names(df_wide %>% select(-offset, -ALL)), 2, simplify = T) %>% 
    t() %>% 
    as.data.frame() %>% rename(x = V1, y = V2)
  
  plts <- map2(.x = nms$x, 
               .y = nms$y,
            ~{ggplot(data = df_wide, 
                     aes(x = .data[[.x]], 
                                y = .data[[.y]])) +
                geom_point()}
            )
  
  print(gridExtra::grid.arrange(grobs = plts))
  
  for(poppy in c("ALL", "AFR", "AMR", "EAS", "EUR", "SAS")){
    p <- df_res %>%
      filter(pop == poppy) %>%
      ggplot(aes(x = offset, y = re, colour = is_sig)) +
      geom_point() +
      scale_colour_manual(values = friendly_pal("vibrant_seven")) +
      theme_classic() +
      labs(colour = "Significant") +
      xlab("Relative Position") +
      ylab("Relative Entropy") +
      ggtitle(paste0("Population: ", poppy, "; Subtype: ", subtype)) +
      scale_y_log10()
    print(p)
  }
  
  df_resid %>%
    pivot_wider(id_cols = rp:status, names_from = pop, values_from = res) %>%
    select(ALL:SAS) %>%
    correlate(method = "pearson", diagonal = 1) %>%
    shave() %>%
    fashion(decimals = 3) %>%
    print()
  
  df_resid %>%
    pivot_wider(id_cols = rp:status, names_from = pop, values_from = res) %>%
    select(ALL:SAS) %>%
    correlate(method = "spearman", diagonal = 1) %>%
    shave() %>%
    fashion(decimals = 3) %>%
    print()
}

single_pos_analysis(subtype)
```

### A > G

```{r}
subtype <- "AT_GC"
single_pos_analysis(subtype)
```

### A > T

```{r}
subtype <- "AT_TA"
single_pos_analysis(subtype)
```

### C > T

```{r}
subtype <- "GC_AT"
single_pos_analysis(subtype)
```

### C > A

```{r}
subtype <- "GC_TA"
single_pos_analysis(subtype)
```

### C > G

```{r}
subtype <- "GC_CG"
single_pos_analysis(subtype)
```

### CpG > TpG

```{r}
subtype <- "cpg_GC_AT"
single_pos_analysis(subtype, rp = c(-10:-1,2:10))
```

### CpG > ApG

```{r}
subtype <- "cpg_GC_TA"
single_pos_analysis(subtype, rp = c(-10:-1,2:10))
```

### CpG > GpG

```{r}
subtype <- "cpg_GC_CG"
single_pos_analysis(subtype, rp = c(-10:-1,2:10))
```
