---
title: "pop_comparison"
author: "Andy Beck"
date: "2025-01-20"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

In this document we are primarily interested in assessing the consistency of results across the five 1kGP super-populations. Namely, for the single position models:

1. Are the patterns of influence seen across positions shared across populations?
2. At individual positions, are the same nucleotides the most influential across populations?

The questions we ask for the two position models are similar.

Addendum:

* What differences might be interesting to suss out?
* What if we utilize our model framework to identify differences between the populations?
* Are "windows of significance consistent across populations?

## Setup and Data Functions

### Libraries

```{r}
library(tidyverse)
library(ggpubfigs) # functions for color blind friendly palettes
library(reactable) # table output for html
library(knitr) # PDF table output
library(corrr)

subtype_print_names <- function(st){
  if(str_starts(st, "AT")){
    return(paste0("A → ", str_sub(st, 4, 4)))
  } else if(str_starts(st, "GC")){
    return(paste0("C → ", str_sub(st, 5, 5)))
  } else{
    return(paste0("CpG → ", str_sub(st, 9, 9), "pG"))
  }
}
```

### Single Position Functions

Load position-level results:

```{r}
sp_deviance_df_load <- function(res_dir, pop, subtype){
  f_name <- paste0(res_dir, "/", pop, "/", subtype, ".csv")
  df <- read_csv(f_name, show_col_types = FALSE) %>%
    rowwise() %>%
    mutate(re = dev / (2 * (singletons+controls)),
           is_sig = dev > qchisq(0.05, df = 3, lower.tail = F))
  return(df)
}

sp_re_single_pop <- function(res_dir, subtype, pop){
  df <- sp_deviance_df_load(res_dir, pop, subtype) %>%
    mutate(pop=pop)
}

sp_re_all_pop <- function(res_dir, subtype){
  df_ALL <- sp_re_single_pop(res_dir, subtype, "ALL")
  
  df_AFR <- sp_re_single_pop(res_dir, subtype, "AFR")
  
  df_AMR <- sp_re_single_pop(res_dir, subtype, "AMR")
  
  df_EAS <- sp_re_single_pop(res_dir, subtype, "EAS")
  
  df_EUR <- sp_re_single_pop(res_dir, subtype, "EUR")
  
  df_SAS <- sp_re_single_pop(res_dir, subtype, "SAS")
  
  #return(bind_rows(df_ALL, df_AFR, df_AMR, df_EAS, df_EUR, df_SAS) %>%
  #  pivot_wider(id_cols = offset, names_from = pop, values_from = re))
  return(bind_rows(df_ALL, df_AFR, df_AMR, df_EAS, df_EUR, df_SAS))
}

get_pos_cc <- function(control_control, pop, subtype){
  f_name <- paste0(control_control, "/", pop, "/", subtype, "_v2.csv")
  df <- read_csv(f_name, show_col_types = FALSE) %>%
           mutate(re = dev / (2 * (singletons + controls) ))
  return(df)
}

get_pos_cc_pops <- function(control_control, subtype){
  pops <- c("ALL", "AFR", "AMR", "EAS", "EUR", "SAS")
  results <- vector("list", length(pops))
  names(results) <- pops
  for(p in pops){
    results[[p]] <- get_pos_cc(control_control, p, subtype)
    results[[p]]$pop <- p
  }
  return(bind_rows(results))
}

plot_re_cc_autosig <- function(df_res, df_cc, p, win_size = 10, sig_frac = 2/10){
  df_cc_sub <- df_cc %>% filter(pop == p) %>%
    mutate(is_sig = "Control")
  df_res_sub <- df_res %>% filter(pop == p)
  q_95_c <- quantile(df_cc_sub$re, 0.95)
  
  frac_sig_df <- data.frame(offset = min(df_res_sub$offset):(max(df_res_sub$offset) - win_size), frac_sig = 0)
  frac_sig_df <- frac_sig_df %>%
    filter(offset %in% df_res_sub$offset)
  
  for(i in 1:length(frac_sig_df$offset)){
    win_low = frac_sig_df$offset[i]
    win_high = win_low + win_size
    tmp_vec <- df_res_sub %>%
      filter(offset >= win_low, offset <= win_high) %>%
      mutate(is_sig = re > q_95_c) %>%
      pull(is_sig)
    frac_sig_df$frac_sig[i] <- sum(tmp_vec) / length(tmp_vec)
  }
  # frac_sig_df %>%
  #   ggplot(aes(x = offset, y = frac_sig)) +
  #   geom_point()
  left_boundary <- frac_sig_df %>%
    filter(frac_sig >= sig_frac) %>%
    pull(offset) %>%
    min()
  right_boundary <- frac_sig_df %>%
    filter(frac_sig >= sig_frac) %>%
    pull(offset) %>%
    max()
  p <- df_res_sub %>%
    mutate(is_sig = as.character(is_sig)) %>%
    bind_rows(df_cc_sub) %>%
    ggplot(aes(x = offset, y = re, colour = is_sig, alpha = is_sig)) +
    geom_point() +
    scale_alpha_manual(values = c(0.2, 1, 1)) + guides(alpha="none") + 
    scale_color_manual(values = friendly_pals$wong_eight[c(3,1:2)]) +
    theme_classic() +
    theme(text = element_text(family = "Arial"),
        legend.position = "inside",
        legend.position.inside = c(.85, .8)) +
    labs(colour = "Significant") +
    xlab("Relative Position") +
    ylab("Relative Entropy (log10)") +
    ggtitle(paste0("Population: ", p, "; Subtype: ", subtype)) +
    scale_y_log10() +
    geom_hline(yintercept = q_95_c, colour = friendly_pals$wong_eight[3], linetype = 2) +
    geom_vline(xintercept = left_boundary, colour = friendly_pals$wong_eight[2]) +
    geom_vline(xintercept = right_boundary, colour = friendly_pals$wong_eight[2]) +
    annotate("text", x = left_boundary - 110, y = min(df_res_sub$re), label = left_boundary) + 
    annotate("text", x = right_boundary + 100, y = min(df_res_sub$re), label = right_boundary) 
  return(p)
}
```

Load position-level residuals:

```{r}
sp_resid_df_load <- function(res_dir, pop, subtype, rp){
  f_name <- paste0(res_dir, "/resid/", pop, "/", subtype, "_rp_", rp, ".csv")
  df <- read_csv(f_name, show_col_types = FALSE)
  df$rp <- rp
  return(df)
}

sp_resid_df_load_all <- function(res_dir, pop, subtype, rp = c(-500:-1, 1:500)){
  results <- vector(mode = "list", length = length(rp))
  for(i in 1:length(rp)){
    if(rp[i] == 1 & str_starts(subtype, "cpg")) next
    results[[i]] <- sp_resid_df_load(res_dir, pop, subtype, rp[i])
  }
  df <- bind_rows(results)
  return(df)
}

sp_resid_all_pop <- function(res_dir, subtype, rp = c(-500:-1, 1:500)){
  df_ALL <- sp_resid_df_load_all(res_dir, "ALL", subtype, rp) %>%
    select(rp, nuc, status, res) %>%
    mutate(pop = "ALL")
  
  df_AFR <- sp_resid_df_load_all(res_dir, "AFR", subtype, rp) %>%
    select(rp, nuc, status, res) %>%
    mutate(pop = "AFR")
  
  df_AMR <- sp_resid_df_load_all(res_dir, "AMR", subtype, rp) %>%
    select(rp, nuc, status, res) %>%
    mutate(pop = "AMR")
  
  df_EAS <- sp_resid_df_load_all(res_dir, "EAS", subtype, rp) %>%
    select(rp, nuc, status, res) %>%
    mutate(pop = "EAS")
  
  df_EUR <- sp_resid_df_load_all(res_dir, "EUR", subtype, rp) %>%
    select(rp, nuc, status, res) %>%
    mutate(pop = "EUR")
  
  df_SAS <- sp_resid_df_load_all(res_dir, "SAS", subtype, rp) %>%
    select(rp, nuc, status, res) %>%
    mutate(pop = "SAS")
  
  return(bind_rows(df_ALL, df_AFR, df_AMR, df_EAS, df_EUR, df_SAS))
}
```

Load population-population models:

```{r}
load_pop_pop_resid <- function(pop_pop_dir, pop1, pop2, subtype, rp){
  f_name <- paste0(pop_pop_dir, "/", pop1, "_", pop2, "/", subtype, "_rp_", rp, ".csv")
  df <- read_csv(f_name, show_col_types = FALSE)
  return(df)
}

pop_pop_dev <- function(pop_pop_dir, pop1, pop2, subtype, rp){
  df <- load_pop_pop_resid(pop_pop_dir, pop1, pop2, subtype, rp)
  dev_stat <- sum((df$res)^2)
  n_singletons <- df %>%
    filter(status == "singletons") %>%
    pull(n) %>%
    sum()
  n_controls <- df %>%
    filter(status == "singletons") %>%
    pull(n) %>%
    sum()
  re_val <- dev_stat / (2*(n_singletons + n_controls))
  is_sig <- dev_stat > qchisq(0.95, 3)
  return(data.frame(rp = rp, dev = dev_stat, re = re_val, singletons = n_singletons, controls = n_controls, is_sig = is_sig))
}

load_pop_pop_df <- function(pop_pop_dir, pop1, pop2, subtype, rps = c(-20:-1,1:20)){
  results <- vector(mode = "list", length = length(rps))
  for(i in 1:length(rps)){
    results[[i]] <- pop_pop_dev(pop_pop_dir, pop1, pop2, subtype, rps[i])
  }
  return(bind_rows(results))
}
```


### Two position functions

```{r}
tp_deviance_df_load <- function(base_dir, pop, subtype){
  f_name <- paste0(base_dir, "/", pop, "/", subtype, ".csv")
  df <- read_csv(f_name, show_col_types = F, 
                 col_names = c("dev","singletons","controls","rp1","rp2")) %>%
    rowwise() %>%
    mutate(re = dev / (2 * (singletons + controls))) %>%
    ungroup()
  return(df)
}

tp_resid_df_load <- function(base_dir, pop, subtype, p, q){
  f_name <- paste0(base_dir, "/resid/", pop, "/", subtype, "_p", p, "_q", q, ".csv")
  df <- read_csv(f_name, show_col_types = F)
  return(df)
}
```

### Paths

```{r}
base_dir = "output"
single_pos_dir = paste0(base_dir, "/single_pos")
two_pos_dir = paste0(base_dir, "/two_pos")
control_control <- paste0(base_dir, "/single_pos_cc")
pop_pop_dir <- paste0(base_dir, "/single_pos_pp")
```

## Single Position Results

### A > C

```{r}
subtype <- "AT_CG"

residual_analysis <- function(df_resid, subtype){
  grouped_resid_df <- df_resid %>%
    filter(status == "singletons", pop != "ALL") %>%
    group_by(rp, nuc) %>%
    arrange(pop, rp, nuc)
  
  populations <- unique(grouped_resid_df$pop)
  
  corr_matrix <- expand.grid(pop1 = populations, pop2 = populations) %>%
    filter(pop1 != pop2) %>% 
    rowwise() %>%
    mutate(cor_val = cor(
      grouped_resid_df$res[grouped_resid_df$pop == pop1],
      grouped_resid_df$res[grouped_resid_df$pop == pop2],
      method = "pearson"
    )) %>%
    ungroup()
  
  p1 <- corr_matrix %>%
    filter(pop1 != "AFR", pop2 != "SAS") %>%
    rowwise() %>%
    mutate(pop1 = as.character(pop1), pop2 = as.character(pop2)) %>%
    mutate(cor_val = ifelse(pop2 > pop1, NA, cor_val)) %>%
    ggplot(aes(x = pop1, y = pop2, fill = cor_val)) +
    geom_tile() +
    scale_fill_distiller(palette = "Reds", na.value = "white", direction = 1, limits = c(0,1)) +
    theme_classic() +
    xlab("") +
    ylab("") +
    ggtitle(paste0(subtype_print_names(subtype), "; Pearson Correlation"))
  
  corr_matrix <- expand.grid(pop1 = populations, pop2 = populations) %>%
    filter(pop1 != pop2) %>% 
    rowwise() %>%
    mutate(cor_val = cor(
      grouped_resid_df$res[grouped_resid_df$pop == pop1],
      grouped_resid_df$res[grouped_resid_df$pop == pop2],
      method = "spearman"
    )) %>%
    ungroup()
  
  p2 <- corr_matrix %>%
    filter(pop1 != "AFR", pop2 != "SAS") %>%
    rowwise() %>%
    mutate(pop1 = as.character(pop1), pop2 = as.character(pop2)) %>%
    mutate(cor_val = ifelse(pop2 > pop1, NA, cor_val)) %>%
    ggplot(aes(x = pop1, y = pop2, fill = cor_val)) +
    geom_tile() +
    scale_fill_distiller(palette = "Reds", na.value = "white", direction = 1, limits = c(0,1)) +
    theme_classic() +
    xlab("") +
    ylab("") +
    ggtitle(paste0(subtype_print_names(subtype), "; Spearman Correlation"))
  
  corr_at_pos <- data.frame(rp = unique(grouped_resid_df$rp), 
                            AFR_AMR = 0,
                            AFR_EAS = 0,
                            AFR_EUR = 0,
                            AFR_SAS = 0,
                            AMR_EAS = 0,
                            AMR_EUR = 0,
                            AMR_SAS = 0,
                            EAS_EUR = 0,
                            EAS_SAS = 0,
                            EUR_SAS = 0)
  for(i in 1:length(corr_at_pos$rp)){
    df2 <- grouped_resid_df %>%
      filter(rp == corr_at_pos$rp[i]) %>%
      ungroup() %>%
      select(nuc, pop, res) %>%
      pivot_wider(names_from = pop, values_from = res)
    
    corr_at_pos$AFR_AMR[i] <- cor(df2$AFR, df2$AMR, method = "pearson")
    corr_at_pos$AFR_EAS[i] <- cor(df2$AFR, df2$EAS, method = "pearson")
    corr_at_pos$AFR_EUR[i] <- cor(df2$AFR, df2$EUR, method = "pearson")
    corr_at_pos$AFR_SAS[i] <- cor(df2$AFR, df2$SAS, method = "pearson")
    corr_at_pos$AMR_EAS[i] <- cor(df2$AMR, df2$EAS, method = "pearson")
    corr_at_pos$AMR_EUR[i] <- cor(df2$AMR, df2$EUR, method = "pearson")
    corr_at_pos$AMR_SAS[i] <- cor(df2$AMR, df2$SAS, method = "pearson")
    corr_at_pos$EAS_EUR[i] <- cor(df2$EAS, df2$EUR, method = "pearson")
    corr_at_pos$EAS_SAS[i] <- cor(df2$EAS, df2$SAS, method = "pearson")
    corr_at_pos$EUR_SAS[i] <- cor(df2$EUR, df2$SAS, method = "pearson")
  }
  
  p3 <- corr_at_pos %>% 
    pivot_longer(-rp, names_to = "pops", values_to = "cor") %>%
    separate(pops, into = c("pop1", "pop2"), sep = "_") %>%
    ggplot(aes(x = rp, y = cor)) + 
    geom_point() +
    ylim(c(0,1)) + 
    facet_grid(rows = vars(pop2), cols = vars(pop1)) +
    ggtitle(paste0(subtype_print_names(subtype), "; Pearson Correlation"))
  
  for(i in 1:length(corr_at_pos$rp)){
    df2 <- grouped_resid_df %>%
      filter(rp == corr_at_pos$rp[i]) %>%
      ungroup() %>%
      select(nuc, pop, res) %>%
      pivot_wider(names_from = pop, values_from = res)
    
    corr_at_pos$AFR_AMR[i] <- cor(df2$AFR, df2$AMR, method = "spearman")
    corr_at_pos$AFR_EAS[i] <- cor(df2$AFR, df2$EAS, method = "spearman")
    corr_at_pos$AFR_EUR[i] <- cor(df2$AFR, df2$EUR, method = "spearman")
    corr_at_pos$AFR_SAS[i] <- cor(df2$AFR, df2$SAS, method = "spearman")
    corr_at_pos$AMR_EAS[i] <- cor(df2$AMR, df2$EAS, method = "spearman")
    corr_at_pos$AMR_EUR[i] <- cor(df2$AMR, df2$EUR, method = "spearman")
    corr_at_pos$AMR_SAS[i] <- cor(df2$AMR, df2$SAS, method = "spearman")
    corr_at_pos$EAS_EUR[i] <- cor(df2$EAS, df2$EUR, method = "spearman")
    corr_at_pos$EAS_SAS[i] <- cor(df2$EAS, df2$SAS, method = "spearman")
    corr_at_pos$EUR_SAS[i] <- cor(df2$EUR, df2$SAS, method = "spearman")
  }
  
  p4 <- corr_at_pos %>% 
    pivot_longer(-rp, names_to = "pops", values_to = "cor") %>%
    separate(pops, into = c("pop1", "pop2"), sep = "_") %>%
    ggplot(aes(x = rp, y = cor)) + 
    geom_point() +
    ylim(c(0,1)) + 
    facet_grid(rows = vars(pop2), cols = vars(pop1)) +
    ggtitle(paste0(subtype_print_names(subtype), "; Spearman Correlation"))
  
  return(list(p1 = p1, p2 = p2, p3 = p3, p4 = p4))
}

single_pos_analysis <- function(subtype, rp = c(-10:-1, 1:10)){
  df_res <- sp_re_all_pop(single_pos_dir, subtype)
  df_resid <- sp_resid_all_pop(single_pos_dir, subtype, rp = rp)
  df_cc <- get_pos_cc_pops(control_control, subtype)
  
  p <- df_res %>%
    filter(abs(offset) < 20) %>%
    mutate(pop = factor(pop, levels = c("ALL", "AFR", "AMR", "EAS", "EUR", "SAS"))) %>%
    ggplot(aes(x = offset, y = dev, colour = pop)) +
    geom_point() +
    geom_line() +
    scale_colour_manual(values = friendly_pal("vibrant_seven")) +
    theme_classic() +
    labs(colour = "Population") +
    xlab("Relative Position") +
    ylab("Deviance")
  print(p)
  
  p <- p +
    ylab("Deviance (log 10)") + 
    scale_y_log10()
  print(p)
  
  p <- df_res %>%
    filter(abs(offset) < 20) %>%
    mutate(pop = factor(pop, levels = c("ALL", "AFR", "AMR", "EAS", "EUR", "SAS"))) %>%
    ggplot(aes(x = offset, y = re, colour = pop)) +
    geom_point() +
    geom_line() +
    scale_colour_manual(values = friendly_pal("vibrant_seven")) +
    theme_classic() +
    labs(colour = "Population") +
    xlab("Relative Position") +
    ylab("Relative Entropy")
  print(p)
  
  p <- p +
    ylab("Relative Entropy (log10)") +
    scale_y_log10()
  print(p)
  
  p <- df_res %>%
    mutate(pop = factor(pop, levels = c("ALL", "AFR", "AMR", "EAS", "EUR", "SAS"))) %>%
    ggplot(aes(x = offset, y = re, colour = pop)) +
    geom_point(alpha = 0.5) +
    scale_colour_manual(values = friendly_pal("vibrant_seven")) +
    theme_classic() +
    labs(colour = "Population") +
    xlab("Relative Position") +
    ylab("Relative Entropy")
  print(p)
  
  p <- p +
    ylab("Relative Entropy (log10)") +
    scale_y_log10()
  print(p)
  
  cat("Correlation of RE statistics across all flanking positions +/- 1000 bp:\n")
  df_res %>%
    select(-is_sig) %>%
    pivot_wider(id_cols = offset, names_from = pop, values_from = re) %>%
    correlate(method = "pearson", diagonal = 1) %>%
    focus(-offset, mirror = T) %>%
    shave() %>%
    fashion() %>%
    print()
  
  df_res %>%
    select(-is_sig) %>%
    pivot_wider(id_cols = offset, names_from = pop, values_from = re) %>%
    correlate(method = "spearman", diagonal = 1) %>%
    focus(-offset, mirror = T) %>%
    shave() %>%
    fashion() %>%
    print()
  
  cat("Correlation of RE statistics across all flanking positions +/- 20 bp:\n")
  df_res %>%
    filter(abs(offset) < 20) %>%
    select(-is_sig) %>%
    pivot_wider(id_cols = offset, names_from = pop, values_from = re) %>%
    correlate(method = "pearson", diagonal = 1) %>%
    focus(-offset, mirror = T) %>%
    shave() %>%
    fashion() %>%
    print()
  
  df_res %>%
    filter(abs(offset) < 20) %>%
    select(-is_sig) %>%
    pivot_wider(id_cols = offset, names_from = pop, values_from = re) %>%
    correlate(method = "spearman", diagonal = 1) %>%
    focus(-offset, mirror = T) %>%
    shave() %>%
    fashion() %>%
    print()
  
  df_res %>%
    select(-is_sig) %>%
    pivot_wider(id_cols = offset, names_from = pop, values_from = re) %>%
    correlate(method = "spearman", diagonal = 1) %>%
    focus(-offset, mirror = T) %>%
    shave() %>%
    fashion() %>%
    print()
  
  df_wide <- df_res %>%
    select(-is_sig) %>%
    pivot_wider(id_cols = offset, names_from = pop, values_from = re)
  
  nms <- combn(names(df_wide %>% select(-offset, -ALL)), 2, simplify = T) %>% 
    t() %>% 
    as.data.frame() %>% rename(x = V1, y = V2)
  
  plts <- map2(.x = nms$x, 
               .y = nms$y,
            ~{ggplot(data = df_wide, 
                     aes(x = .data[[.x]], 
                                y = .data[[.y]])) +
                geom_point()}
            )
  
  print(gridExtra::grid.arrange(grobs = plts))
  
  for(poppy in c("ALL", "AFR", "AMR", "EAS", "EUR", "SAS")){
    print(plot_re_cc_autosig(df_res, df_cc, poppy, win_size = 10, sig_frac = 2 / 10))
  }
  
  df_resid %>%
    pivot_wider(id_cols = rp:status, names_from = pop, values_from = res) %>%
    select(ALL:SAS) %>%
    correlate(method = "pearson", diagonal = 1) %>%
    shave() %>%
    fashion(decimals = 3) %>%
    print()
  
  df_resid %>%
    pivot_wider(id_cols = rp:status, names_from = pop, values_from = res) %>%
    select(ALL:SAS) %>%
    correlate(method = "spearman", diagonal = 1) %>%
    shave() %>%
    fashion(decimals = 3) %>%
    print()
  
  resid_plots <- residual_analysis(df_resid, subtype)

  print(resid_plots$p1)
  print(resid_plots$p2)
  print(resid_plots$p3)
  print(resid_plots$p4)
}

single_pos_analysis(subtype)
```

```{r}
df_res <- sp_re_all_pop(single_pos_dir, subtype)
df_resid <- sp_resid_all_pop(single_pos_dir, subtype, rp = c(-10:-1, 1:10))
df_cc <- get_pos_cc_pops(control_control, subtype)


# resid_plots <- residual_analysis(df_resid, subtype)
# 
# print(resid_plots$p1)
# print(resid_plots$p2)
# print(resid_plots$p3)
# print(resid_plots$p4)

# for(poppy in c("ALL", "AFR", "AMR", "EAS", "EUR", "SAS")){
#   print(poppy)
#   print(plot_re_cc_autosig(df_res, df_cc, poppy))
# }

df_res %>%
  filter(abs(offset) <= 20) %>%
  group_by(pop) %>%
  mutate(rank_re = dense_rank(desc(re))) %>%
  mutate(pop = factor(pop, levels = c("ALL", "AFR", "AMR", "EAS", "EUR", "SAS"))) %>%
  filter(pop != "ALL") %>%
  ggplot(aes(x = rank_re, y = offset, colour = pop)) +
  geom_point(alpha = 0.5) +
  geom_line() +
  scale_colour_manual(values = friendly_pal("vibrant_seven")) +
  theme_classic() +
  labs(colour = "Population") +
  ylab("Relative Position") +
  xlab("Rank")
  
```


For the A > C subtype, we observe a high correlation of relative entropy statistics at the +/- 20

### A > G

```{r}
subtype <- "AT_GC"
single_pos_analysis(subtype)
```

### A > T

```{r}
subtype <- "AT_TA"
single_pos_analysis(subtype)
```

### C > T

```{r}
subtype <- "GC_AT"
single_pos_analysis(subtype)
```

### C > A

```{r}
subtype <- "GC_TA"
single_pos_analysis(subtype)
```

### C > G

```{r}
subtype <- "GC_CG"
single_pos_analysis(subtype)
```

### CpG > TpG

```{r}
subtype <- "cpg_GC_AT"
single_pos_analysis(subtype, rp = c(-10:-1,2:10))
```

### CpG > ApG

```{r}
subtype <- "cpg_GC_TA"
single_pos_analysis(subtype, rp = c(-10:-1,2:10))
```

### CpG > GpG

```{r}
subtype <- "cpg_GC_CG"
single_pos_analysis(subtype, rp = c(-10:-1,2:10))
```

## Single Position Pop-Pop Model

In the following section, we utilize our log-linear model framework to assess consistency of single position model results by fitting models of the following form (in R's model notation):

```
glm(count ~ nuc*status + nuc*population + status*population, family = poisson())
```

Here, in contrast to our single population models, we allow for the distributions of nucleotides at flanking positions to differ between the singletons and the controls via the inclusion of a nucleotide-by-status interaction. The devinace statistic for this model evaluates the fit of this model to a fully-saturated model, in which in addition to the above terms is a three-way interaction between nucleotide, population, and status; under the null hypothesis that these parameters are 0, we assume that any nucleotide by status interaction is shared across the populations. Note that we fit these models for pairs of populations, so the degree of freedom for this test statistic is three (three two-way interactions between nucleotide and status, one population parameter to interact with).

```{r}
pop_combs <- combn(c("AFR", "AMR", "EAS", "EUR", "SAS"), 2) %>% t() 
```


### A > C

```{r}
subtype <- "AT_CG"

pop_pop_analysis <- function(pop_pop_dir, pop1, pop2, subtype, rps = c(-20:-1, 1:20)){
  df <- load_pop_pop_df(pop_pop_dir, pop1, pop2, subtype, rps)

  p <- df %>%
    ggplot(aes(x = rp, y = dev, colour = is_sig)) +
    geom_point() +
    scale_color_manual(values = friendly_pals$wong_eight[c(1:2)]) +
    theme_classic() +
    theme(text = element_text(family = "Arial"),
          legend.position = "inside",
          legend.position.inside = c(.85, .8)) +
    labs(colour = "Significant") +
    xlab("Relative Position") +
    ylab("Deviance") + 
    ggtitle(paste0(subtype_print_names(subtype), ": ", pop1, ", ", pop2))
  return(p)
}

for(i in 1:dim(pop_combs)[1]){
  pop1 <- pop_combs[i,1]
  pop2 <- pop_combs[i,2]
  p <- pop_pop_analysis(pop_pop_dir, pop1, pop2, subtype, rps = c(-10:-1, 1:10)) +
    geom_hline(yintercept = qchisq(0.95, 3), color = "red", linetype = 2) 
  print(p)
}
```


### A > G

```{r}
subtype <- "AT_GC"

for(i in 1:dim(pop_combs)[1]){
  pop1 <- pop_combs[i,1]
  pop2 <- pop_combs[i,2]
  p <- pop_pop_analysis(pop_pop_dir, pop1, pop2, subtype, rps = c(-10:-1, 1:10)) +
    geom_hline(yintercept = qchisq(0.95, 3), color = "red", linetype = 2) 
  print(p)
}
```

### A > T

```{r}
subtype <- "AT_TA"

for(i in 1:dim(pop_combs)[1]){
  pop1 <- pop_combs[i,1]
  pop2 <- pop_combs[i,2]
  p <- pop_pop_analysis(pop_pop_dir, pop1, pop2, subtype, rps = c(-10:-1, 1:10)) +
    geom_hline(yintercept = qchisq(0.95, 3), color = "red", linetype = 2) 
  print(p)
}
```

### C > T

```{r}
subtype <- "GC_AT"

for(i in 1:dim(pop_combs)[1]){
  pop1 <- pop_combs[i,1]
  pop2 <- pop_combs[i,2]
  p <- pop_pop_analysis(pop_pop_dir, pop1, pop2, subtype, rps = c(-10:-1, 1:10)) +
    geom_hline(yintercept = qchisq(0.95, 3), color = "red", linetype = 2) 
  print(p)
}
```

### C > A

```{r}
subtype <- "GC_TA"

for(i in 1:dim(pop_combs)[1]){
  pop1 <- pop_combs[i,1]
  pop2 <- pop_combs[i,2]
  p <- pop_pop_analysis(pop_pop_dir, pop1, pop2, subtype, rps = c(-10:-1, 1:10)) +
    geom_hline(yintercept = qchisq(0.95, 3), color = "red", linetype = 2) 
  print(p)
}
```

### C > G

```{r}
subtype <- "GC_CG"

for(i in 1:dim(pop_combs)[1]){
  pop1 <- pop_combs[i,1]
  pop2 <- pop_combs[i,2]
  p <- pop_pop_analysis(pop_pop_dir, pop1, pop2, subtype, rps = c(-10:-1, 1:10)) +
    geom_hline(yintercept = qchisq(0.95, 3), color = "red", linetype = 2) 
  print(p)
}
```

### CpG > TpG

```{r}
subtype <- "GC_AT"

for(i in 1:dim(pop_combs)[1]){
  pop1 <- pop_combs[i,1]
  pop2 <- pop_combs[i,2]
  p <- pop_pop_analysis(pop_pop_dir, pop1, pop2, subtype, rps = c(-10:-1, 2:10)) +
    geom_hline(yintercept = qchisq(0.95, 3), color = "red", linetype = 2) 
  print(p)
}
```

### CpG > ApG

```{r}
subtype <- "GC_TA"

for(i in 1:dim(pop_combs)[1]){
  pop1 <- pop_combs[i,1]
  pop2 <- pop_combs[i,2]
  p <- pop_pop_analysis(pop_pop_dir, pop1, pop2, subtype, rps = c(-10:-1, 2:10)) +
    geom_hline(yintercept = qchisq(0.95, 3), color = "red", linetype = 2) 
  print(p)
}
```

### CpG > GpG

```{r}
subtype <- "GC_CG"

for(i in 1:dim(pop_combs)[1]){
  pop1 <- pop_combs[i,1]
  pop2 <- pop_combs[i,2]
  p <- pop_pop_analysis(pop_pop_dir, pop1, pop2, subtype, rps = c(-10:-1, 2:10)) +
    geom_hline(yintercept = qchisq(0.95, 3), color = "red", linetype = 2) 
  print(p)
}
```

