---
title: "pop_comparison"
author: "Andy Beck"
date: "2025-01-20"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

In this document we are primarily interested in assessing the consistency of results across the five 1kGP super-populations. Namely, for the single position models:

1. Are the patterns of influence seen across positions shared across populations?
2. At individual positions, are the same nucleotides the most influential across populations?

The questions we ask for the two position models are similar.

Addendum:

* What differences might be interesting to suss out?
* What if we utilize our model framework to identify differences between the populations?
* Are "windows of significance consistent across populations?

## Setup and Data Functions

### Libraries

```{r}
library(tidyverse)
library(ggpubfigs) # functions for color blind friendly palettes
library(reactable) # table output for html
library(knitr) # PDF table output
library(corrr)
library(gtsummary)

subtype_print_names <- function(st){
  if(str_starts(st, "AT")){
    return(paste0("A → ", str_sub(st, 4, 4)))
  } else if(str_starts(st, "GC")){
    return(paste0("C → ", str_sub(st, 5, 5)))
  } else{
    return(paste0("CpG → ", str_sub(st, 9, 9), "pG"))
  }
}

subtypes <- c("AT_CG", "AT_GC", "AT_TA",
              "GC_AT", "GC_TA", "GC_CG",
              "cpg_GC_AT", "cpg_GC_TA", "cpg_GC_CG")
```

### Single Position Functions

Load position-level results:

```{r}
sp_deviance_df_load <- function(res_dir, pop, subtype){
  f_name <- paste0(res_dir, "/", pop, "/", subtype, ".csv")
  df <- read_csv(f_name, show_col_types = FALSE) %>%
    rowwise() %>%
    mutate(re = dev / (2 * (singletons+controls)),
           is_sig = dev > qchisq(0.05, df = 3, lower.tail = F))
  return(df)
}

sp_re_single_pop <- function(res_dir, subtype, pop){
  df <- sp_deviance_df_load(res_dir, pop, subtype) %>%
    mutate(pop=pop)
}

sp_re_all_pop <- function(res_dir, subtype){
  df_ALL <- sp_re_single_pop(res_dir, subtype, "ALL")
  
  df_AFR <- sp_re_single_pop(res_dir, subtype, "AFR")
  
  df_AMR <- sp_re_single_pop(res_dir, subtype, "AMR")
  
  df_EAS <- sp_re_single_pop(res_dir, subtype, "EAS")
  
  df_EUR <- sp_re_single_pop(res_dir, subtype, "EUR")
  
  df_SAS <- sp_re_single_pop(res_dir, subtype, "SAS")
  
  #return(bind_rows(df_ALL, df_AFR, df_AMR, df_EAS, df_EUR, df_SAS) %>%
  #  pivot_wider(id_cols = offset, names_from = pop, values_from = re))
  return(bind_rows(df_ALL, df_AFR, df_AMR, df_EAS, df_EUR, df_SAS))
}

get_pos_cc <- function(control_control, pop, subtype){
  f_name <- paste0(control_control, "/", pop, "/", subtype, "_v2.csv")
  df <- read_csv(f_name, show_col_types = FALSE) %>%
           mutate(re = dev / (2 * (singletons + controls) ))
  return(df)
}

get_pos_cc_pops <- function(control_control, subtype){
  pops <- c("ALL", "AFR", "AMR", "EAS", "EUR", "SAS")
  results <- vector("list", length(pops))
  names(results) <- pops
  for(p in pops){
    results[[p]] <- get_pos_cc(control_control, p, subtype)
    results[[p]]$pop <- p
  }
  return(bind_rows(results))
}

window_frac_sig <- function(df_res, df_cc, population, win_size = 10){
  df_cc_sub <- df_cc %>% filter(pop == population) %>%
    mutate(is_sig = "Control")
  df_res_sub <- df_res %>% filter(pop == population)
  q_95_c <- quantile(df_cc_sub$re, 0.95)
  
  frac_sig_df <- data.frame(offset = min(df_res_sub$offset):(max(df_res_sub$offset) - win_size), frac_sig = 0)
  frac_sig_df <- frac_sig_df %>%
    filter(offset %in% df_res_sub$offset)
  
  for(i in 1:length(frac_sig_df$offset)){
    win_low = frac_sig_df$offset[i]
    win_high = win_low + win_size
    tmp_vec <- df_res_sub %>%
      filter(offset >= win_low, offset <= win_high) %>%
      mutate(is_sig = re > q_95_c) %>%
      pull(is_sig)
    frac_sig_df$frac_sig[i] <- sum(tmp_vec) / length(tmp_vec)
  }
  return(frac_sig_df)
}

plot_re_cc_autosig <- function(df_res, df_cc, population, win_size = 10, sig_frac = 8/10){
  df_cc_sub <- df_cc %>% filter(pop == population) %>%
    mutate(is_sig = "Control")
  df_res_sub <- df_res %>% filter(pop == population)
  q_95_c <- quantile(df_cc_sub$re, 0.95)
  
  frac_sig_df <- window_frac_sig(df_res, df_cc, population, win_size) 
  
  frac_sig_df %>%
    ggplot(aes(x = offset, y = frac_sig)) +
    geom_point()
  
  # frac_sig_df <- frac_sig_df %>%
  #   mutate(frac_sig_lead = lead(frac_sig),
  #          frac_sig_lag = lag(frac_sig))
  
  left_boundary <- frac_sig_df %>%
    filter(frac_sig >= sig_frac) %>% #, frac_sig_lead >= sig_frac, frac_sig_lag >= sig_frac) %>%
    pull(offset) %>%
    min()
  right_boundary <- frac_sig_df %>%
    filter(frac_sig >= sig_frac) %>% #, frac_sig_lead >= sig_frac, frac_sig_lag >= sig_frac) %>%
    pull(offset) %>%
    max()
  
  p <- df_res_sub %>%
    mutate(is_sig = as.character(is_sig)) %>%
    bind_rows(df_cc_sub) %>%
    ggplot(aes(x = offset, y = re, colour = is_sig, alpha = is_sig)) +
    geom_point() +
    scale_alpha_manual(values = c(0.2, 1, 1)) + guides(alpha="none") + 
    scale_color_manual(values = friendly_pals$wong_eight[c(3,1:2)]) +
    theme_classic() +
    theme(text = element_text(family = "Arial"),
        legend.position = "inside",
        legend.position.inside = c(.85, .8)) +
    labs(colour = "Significant") +
    xlab("Relative Position") +
    ylab("Relative Entropy (log10)") +
    scale_y_log10() +
    geom_vline(xintercept = left_boundary, colour = friendly_pals$wong_eight[2]) +
    geom_vline(xintercept = right_boundary, colour = friendly_pals$wong_eight[2]) +
    annotate("text", x = left_boundary - 110, y = min(df_res_sub$re), label = left_boundary) + 
    annotate("text", x = right_boundary + 100, y = min(df_res_sub$re), label = right_boundary) +
    ggtitle(paste0("Population: ", population, "; Subtype: ", subtype_print_names(subtype)))
  return(p)
}

get_df_windows <- function(df_res, df_cc, population, thresh_vals = (1:10)/10){
  df_cc_sub <- df_cc %>% filter(pop == population)
  df_res_sub <- df_res %>% filter(pop == population)

  q_95_c <- quantile(df_cc_sub$re, 0.95)
  frac_sig_df <- window_frac_sig(df_res, df_cc, population)
  df_windows <- data.frame(frac_thresh = thresh_vals, 
                           left_boundary = -1000, 
                           right_boundary = 1000,
                           frac_in = 0,
                           frac_out = 0)
  
  for(i in 1:length(thresh_vals)){
    sig_frac <- thresh_vals[i]
    left_boundary <- frac_sig_df %>%
      filter(frac_sig >= sig_frac) %>% #, frac_sig_lead >= sig_frac, frac_sig_lag >= sig_frac) %>%
      pull(offset) %>%
      min()
    right_boundary <- frac_sig_df %>%
      filter(frac_sig >= sig_frac) %>% #, frac_sig_lead >= sig_frac, frac_sig_lag >= sig_frac) %>%
      pull(offset) %>%
      max()
    
    sig_vec <- df_res_sub %>%
      filter(offset < left_boundary | offset > right_boundary) %>%
      mutate(is_sig = re > q_95_c) %>%
      pull(is_sig)
    df_windows$frac_out[i] <- sum(sig_vec) / length(sig_vec)
    
    sig_vec <- df_res_sub %>%
      filter(offset >= left_boundary,  offset <= right_boundary) %>%
      mutate(is_sig = re > q_95_c) %>%
      pull(is_sig)
    df_windows$frac_in[i] <- sum(sig_vec) / length(sig_vec)
    
    
    df_windows$left_boundary[i] <- left_boundary
    df_windows$right_boundary[i] <- right_boundary
  }
  return(df_windows)
}
```

Load position-level residuals:

```{r}
sp_resid_df_load <- function(res_dir, pop, subtype, rp){
  f_name <- paste0(res_dir, "/resid/", pop, "/", subtype, "_rp_", rp, ".csv")
  df <- read_csv(f_name, show_col_types = FALSE)
  df$rp <- rp
  return(df)
}

sp_resid_df_load_all <- function(res_dir, pop, subtype, rp = c(-500:-1, 1:500)){
  results <- vector(mode = "list", length = length(rp))
  for(i in 1:length(rp)){
    if(rp[i] == 1 & str_starts(subtype, "cpg")) next
    results[[i]] <- sp_resid_df_load(res_dir, pop, subtype, rp[i])
  }
  df <- bind_rows(results)
  return(df)
}

sp_resid_all_pop <- function(res_dir, subtype, rp = c(-500:-1, 1:500)){
  df_ALL <- sp_resid_df_load_all(res_dir, "ALL", subtype, rp) %>%
    select(rp, nuc, status, res) %>%
    mutate(pop = "ALL")
  
  df_AFR <- sp_resid_df_load_all(res_dir, "AFR", subtype, rp) %>%
    select(rp, nuc, status, res) %>%
    mutate(pop = "AFR")
  
  df_AMR <- sp_resid_df_load_all(res_dir, "AMR", subtype, rp) %>%
    select(rp, nuc, status, res) %>%
    mutate(pop = "AMR")
  
  df_EAS <- sp_resid_df_load_all(res_dir, "EAS", subtype, rp) %>%
    select(rp, nuc, status, res) %>%
    mutate(pop = "EAS")
  
  df_EUR <- sp_resid_df_load_all(res_dir, "EUR", subtype, rp) %>%
    select(rp, nuc, status, res) %>%
    mutate(pop = "EUR")
  
  df_SAS <- sp_resid_df_load_all(res_dir, "SAS", subtype, rp) %>%
    select(rp, nuc, status, res) %>%
    mutate(pop = "SAS")
  
  return(bind_rows(df_ALL, df_AFR, df_AMR, df_EAS, df_EUR, df_SAS))
}
```

Load population-population models:

```{r}
load_pop_pop_resid <- function(pop_pop_dir, pop1, pop2, subtype, rp){
  f_name <- paste0(pop_pop_dir, "/", pop1, "_", pop2, "/", subtype, "_rp_", rp, ".csv")
  df <- read_csv(f_name, show_col_types = FALSE)
  return(df)
}

pop_pop_dev <- function(pop_pop_dir, pop1, pop2, subtype, rp){
  df <- load_pop_pop_resid(pop_pop_dir, pop1, pop2, subtype, rp)
  dev_stat <- sum((df$res)^2)
  n_singletons <- df %>%
    filter(status == "singletons") %>%
    pull(n) %>%
    sum()
  n_controls <- df %>%
    filter(status == "singletons") %>%
    pull(n) %>%
    sum()
  re_val <- dev_stat / (2*(n_singletons + n_controls))
  is_sig <- dev_stat > qchisq(0.95, 3)
  return(data.frame(rp = rp, dev = dev_stat, re = re_val, singletons = n_singletons, controls = n_controls, is_sig = is_sig))
}

load_pop_pop_df <- function(pop_pop_dir, pop1, pop2, subtype, rps = c(-20:-1,1:20)){
  results <- vector(mode = "list", length = length(rps))
  for(i in 1:length(rps)){
    results[[i]] <- pop_pop_dev(pop_pop_dir, pop1, pop2, subtype, rps[i])
  }
  return(bind_rows(results))
}
```

Get all pairs of population-population correlation values for single position models:

```{r}
single_pos_resid_table <- function(single_pos_dir, restriced_range = 20){
  subtypes <- c("AT_CG", "AT_GC", "AT_TA",
                "GC_AT", "GC_TA", "GC_CG",
                "cpg_GC_AT", "cpg_GC_TA", "cpg_GC_CG")
  
  cor_types <- c("pearson", "spearman")
  
  pops <- c("AFR", "AMR", "EAS", "EUR", "SAS")
  pop_combs <- combn(pops,2)
  pops_df <- data.frame(pop1 = pop_combs[1,],
                        pop2 = pop_combs[2,])
  
  final <- data.frame(subtype = character(0), 
                      range = character(0),
                      res_type = character(0),
                      pop1 = character(0),
                      pop2 = character(0),
                      value = numeric(0))
  
  for(st in subtypes){
    df_res <- sp_re_all_pop(single_pos_dir, st)
    
    for(i in 1:length(pops_df$pop1)){
      pop1 <- pops_df$pop1[i]
      pop2 <- pops_df$pop2[i]
      df_res2 <- df_res %>%
        filter(pop == pop1 | pop == pop2) %>%
        select(offset, re, pop)
      
      for(ct in cor_types){
        # All positions
        value <- df_res2 %>%
          pivot_wider(id_cols = "offset", names_from = "pop", values_from = re) %>%
          {cor(.[,2], .[,3], method = ct)} %>%
          as.numeric()
        final <- bind_rows(final, 
                           data.frame(subtype = st, 
                                      range = "all", 
                                      res_type = ct, 
                                      pop1 = pop1, 
                                      pop2=pop2, 
                                      value = value))
        # +/- 20
        value <- df_res2 %>%
          filter(abs(offset) <= restriced_range) %>% 
          pivot_wider(id_cols = "offset", names_from = "pop", values_from = re) %>%
          {cor(.[,2], .[,3], method = ct)} %>%
          as.numeric()
        final <- bind_rows(final, 
                           data.frame(subtype = st, 
                                      range = "restricted", 
                                      res_type = ct, 
                                      pop1 = pop1, 
                                      pop2=pop2, 
                                      value = value))
      }
    }
  }
  return(final)
}
```


### Two position functions

```{r}
tp_deviance_df_load <- function(base_dir, pop, subtype){
  f_name <- paste0(base_dir, "/", pop, "/", subtype, ".csv")
  df <- read_csv(f_name, show_col_types = F, 
                 col_names = c("dev","singletons","controls","rp1","rp2")) %>%
    rowwise() %>%
    mutate(re = dev / (2 * (singletons + controls))) %>%
    ungroup()
  return(df)
}

tp_resid_df_load <- function(base_dir, pop, subtype, p, q){
  f_name <- paste0(base_dir, "/resid/", pop, "/", subtype, "_p", p, "_q", q, ".csv")
  df <- read_csv(f_name, show_col_types = F)
  return(df)
}
```

### Paths

```{r}
base_dir = "output"
single_pos_dir = paste0(base_dir, "/single_pos")
two_pos_dir = paste0(base_dir, "/two_pos")
control_control <- paste0(base_dir, "/single_pos_cc")
pop_pop_dir <- paste0(base_dir, "/single_pos_pp")
```

### Single Position Subtype Analysis Functions

```{r}
residual_analysis <- function(df_resid, subtype, no_plots = F){
  grouped_resid_df <- df_resid %>%
    filter(status == "singletons", pop != "ALL") %>%
    group_by(rp, nuc) %>%
    arrange(pop, rp, nuc)
  
  populations <- unique(grouped_resid_df$pop)
  
  
  corr_at_pos <- data.frame(rp = unique(grouped_resid_df$rp), 
                            AFR_AMR = 0,
                            AFR_EAS = 0,
                            AFR_EUR = 0,
                            AFR_SAS = 0,
                            AMR_EAS = 0,
                            AMR_EUR = 0,
                            AMR_SAS = 0,
                            EAS_EUR = 0,
                            EAS_SAS = 0,
                            EUR_SAS = 0)
  for(i in 1:length(corr_at_pos$rp)){
    df2 <- grouped_resid_df %>%
      filter(rp == corr_at_pos$rp[i]) %>%
      ungroup() %>%
      select(nuc, pop, res) %>%
      pivot_wider(names_from = pop, values_from = res)
    
    corr_at_pos$AFR_AMR[i] <- cor(df2$AFR, df2$AMR, method = "pearson")
    corr_at_pos$AFR_EAS[i] <- cor(df2$AFR, df2$EAS, method = "pearson")
    corr_at_pos$AFR_EUR[i] <- cor(df2$AFR, df2$EUR, method = "pearson")
    corr_at_pos$AFR_SAS[i] <- cor(df2$AFR, df2$SAS, method = "pearson")
    corr_at_pos$AMR_EAS[i] <- cor(df2$AMR, df2$EAS, method = "pearson")
    corr_at_pos$AMR_EUR[i] <- cor(df2$AMR, df2$EUR, method = "pearson")
    corr_at_pos$AMR_SAS[i] <- cor(df2$AMR, df2$SAS, method = "pearson")
    corr_at_pos$EAS_EUR[i] <- cor(df2$EAS, df2$EUR, method = "pearson")
    corr_at_pos$EAS_SAS[i] <- cor(df2$EAS, df2$SAS, method = "pearson")
    corr_at_pos$EUR_SAS[i] <- cor(df2$EUR, df2$SAS, method = "pearson")
  }
  
  df_pearson <- corr_at_pos
  
  for(i in 1:length(corr_at_pos$rp)){
    df2 <- grouped_resid_df %>%
      filter(rp == corr_at_pos$rp[i]) %>%
      ungroup() %>%
      select(nuc, pop, res) %>%
      pivot_wider(names_from = pop, values_from = res)
    
    corr_at_pos$AFR_AMR[i] <- cor(df2$AFR, df2$AMR, method = "spearman")
    corr_at_pos$AFR_EAS[i] <- cor(df2$AFR, df2$EAS, method = "spearman")
    corr_at_pos$AFR_EUR[i] <- cor(df2$AFR, df2$EUR, method = "spearman")
    corr_at_pos$AFR_SAS[i] <- cor(df2$AFR, df2$SAS, method = "spearman")
    corr_at_pos$AMR_EAS[i] <- cor(df2$AMR, df2$EAS, method = "spearman")
    corr_at_pos$AMR_EUR[i] <- cor(df2$AMR, df2$EUR, method = "spearman")
    corr_at_pos$AMR_SAS[i] <- cor(df2$AMR, df2$SAS, method = "spearman")
    corr_at_pos$EAS_EUR[i] <- cor(df2$EAS, df2$EUR, method = "spearman")
    corr_at_pos$EAS_SAS[i] <- cor(df2$EAS, df2$SAS, method = "spearman")
    corr_at_pos$EUR_SAS[i] <- cor(df2$EUR, df2$SAS, method = "spearman")
  }
  
  if(no_plots == F){
    corr_matrix <- expand.grid(pop1 = populations, pop2 = populations) %>%
      filter(pop1 != pop2) %>% 
      rowwise() %>%
      mutate(cor_val = cor(
        grouped_resid_df$res[grouped_resid_df$pop == pop1],
        grouped_resid_df$res[grouped_resid_df$pop == pop2],
        method = "pearson"
      )) %>%
      ungroup()
    
    p1 <- corr_matrix %>%
      filter(pop1 != "AFR", pop2 != "SAS") %>%
      rowwise() %>%
      mutate(pop1 = as.character(pop1), pop2 = as.character(pop2)) %>%
      mutate(cor_val = ifelse(pop2 > pop1, NA, cor_val)) %>%
      ggplot(aes(x = pop1, y = pop2, fill = cor_val)) +
      geom_tile() +
      scale_fill_distiller(palette = "Reds", na.value = "white", direction = 1, limits = c(0,1)) +
      theme_classic() +
      xlab("") +
      ylab("") +
      ggtitle(paste0(subtype_print_names(subtype), "; Pearson Correlation"))
    
    corr_matrix <- expand.grid(pop1 = populations, pop2 = populations) %>%
      filter(pop1 != pop2) %>% 
      rowwise() %>%
      mutate(cor_val = cor(
        grouped_resid_df$res[grouped_resid_df$pop == pop1],
        grouped_resid_df$res[grouped_resid_df$pop == pop2],
        method = "spearman"
      )) %>%
      ungroup()
    
    p2 <- corr_matrix %>%
      filter(pop1 != "AFR", pop2 != "SAS") %>%
      rowwise() %>%
      mutate(pop1 = as.character(pop1), pop2 = as.character(pop2)) %>%
      mutate(cor_val = ifelse(pop2 > pop1, NA, cor_val)) %>%
      ggplot(aes(x = pop1, y = pop2, fill = cor_val)) +
      geom_tile() +
      scale_fill_distiller(palette = "Reds", na.value = "white", direction = 1, limits = c(0,1)) +
      theme_classic() +
      xlab("") +
      ylab("") +
      ggtitle(paste0(subtype_print_names(subtype), "; Spearman Correlation"))
    
    p3 <- df_pearson %>% 
      pivot_longer(-rp, names_to = "pops", values_to = "cor") %>%
      separate(pops, into = c("pop1", "pop2"), sep = "_") %>%
      ggplot(aes(x = rp, y = cor)) + 
      geom_point() +
      ylim(c(0,1)) + 
      facet_grid(rows = vars(pop2), cols = vars(pop1)) +
      ggtitle(paste0(subtype_print_names(subtype), "; Pearson Correlation"))
    
    p4 <- corr_at_pos %>% 
      pivot_longer(-rp, names_to = "pops", values_to = "cor") %>%
      separate(pops, into = c("pop1", "pop2"), sep = "_") %>%
      ggplot(aes(x = rp, y = cor)) + 
      geom_point() +
      ylim(c(0,1)) + 
      facet_grid(rows = vars(pop2), cols = vars(pop1)) +
      ggtitle(paste0(subtype_print_names(subtype), "; Spearman Correlation"))
    
    return(list(p1 = p1, p2 = p2, p3 = p3, p4 = p4, df_spearman = corr_at_pos, df_pearson = df_pearson))
  }
  else {
    return(list(df_spearman = corr_at_pos, df_pearson = df_pearson))
  }
  
}

single_pos_analysis <- function(subtype, rp = c(-10:-1, 1:10), sig_frac = 5/10){
  df_res <- sp_re_all_pop(single_pos_dir, subtype)
  df_resid <- sp_resid_all_pop(single_pos_dir, subtype, rp = rp)
  df_cc <- get_pos_cc_pops(control_control, subtype)
  
  p <- df_res %>%
    filter(abs(offset) < 20) %>%
    mutate(pop = factor(pop, levels = c("ALL", "AFR", "AMR", "EAS", "EUR", "SAS"))) %>%
    ggplot(aes(x = offset, y = dev, colour = pop)) +
    geom_point() +
    geom_line() +
    scale_colour_manual(values = friendly_pal("vibrant_seven")) +
    theme_classic() +
    labs(colour = "Population") +
    xlab("Relative Position") +
    ylab("Deviance")
  print(p)
  
  p <- p +
    ylab("Deviance (log 10)") + 
    scale_y_log10()
  print(p)
  
  p <- df_res %>%
    filter(abs(offset) < 20) %>%
    mutate(pop = factor(pop, levels = c("ALL", "AFR", "AMR", "EAS", "EUR", "SAS"))) %>%
    ggplot(aes(x = offset, y = re, colour = pop)) +
    geom_point() +
    geom_line() +
    scale_colour_manual(values = friendly_pal("vibrant_seven")) +
    theme_classic() +
    labs(colour = "Population") +
    xlab("Relative Position") +
    ylab("Relative Entropy")
  print(p)
  
  p <- p +
    ylab("Relative Entropy (log10)") +
    scale_y_log10()
  print(p)
  
  p <- df_res %>%
    mutate(pop = factor(pop, levels = c("ALL", "AFR", "AMR", "EAS", "EUR", "SAS"))) %>%
    ggplot(aes(x = offset, y = re, colour = pop)) +
    geom_point(alpha = 0.5) +
    scale_colour_manual(values = friendly_pal("vibrant_seven")) +
    theme_classic() +
    labs(colour = "Population") +
    xlab("Relative Position") +
    ylab("Relative Entropy")
  print(p)
  
  p <- p +
    ylab("Relative Entropy (log10)") +
    scale_y_log10()
  print(p)
  
  cat("Correlation of RE statistics across all flanking positions +/- 1000 bp:\n")
  df_res %>%
    select(-is_sig) %>%
    pivot_wider(id_cols = offset, names_from = pop, values_from = re) %>%
    correlate(method = "pearson", diagonal = 1) %>%
    focus(-offset, mirror = T) %>%
    shave() %>%
    fashion() %>%
    print()
  
  df_res %>%
    select(-is_sig) %>%
    pivot_wider(id_cols = offset, names_from = pop, values_from = re) %>%
    correlate(method = "spearman", diagonal = 1) %>%
    focus(-offset, mirror = T) %>%
    shave() %>%
    fashion() %>%
    print()
  
  cat("Correlation of RE statistics across all flanking positions +/- 20 bp:\n")
  df_res %>%
    filter(abs(offset) < 20) %>%
    select(-is_sig) %>%
    pivot_wider(id_cols = offset, names_from = pop, values_from = re) %>%
    correlate(method = "pearson", diagonal = 1) %>%
    focus(-offset, mirror = T) %>%
    shave() %>%
    fashion() %>%
    print()
  
  df_res %>%
    filter(abs(offset) < 20) %>%
    select(-is_sig) %>%
    pivot_wider(id_cols = offset, names_from = pop, values_from = re) %>%
    correlate(method = "spearman", diagonal = 1) %>%
    focus(-offset, mirror = T) %>%
    shave() %>%
    fashion() %>%
    print()
  
  df_wide <- df_res %>%
    select(-is_sig) %>%
    pivot_wider(id_cols = offset, names_from = pop, values_from = re)
  
  nms <- combn(names(df_wide %>% select(-offset, -ALL)), 2, simplify = T) %>% 
    t() %>% 
    as.data.frame() %>% rename(x = V1, y = V2)
  
  plts <- map2(.x = nms$x, 
               .y = nms$y,
            ~{ggplot(data = df_wide, 
                     aes(x = .data[[.x]], 
                                y = .data[[.y]])) +
                geom_point()}
            )
  
  print(gridExtra::grid.arrange(grobs = plts))
  
  # for(poppy in c("ALL", "AFR", "AMR", "EAS", "EUR", "SAS")){
  #   print(plot_re_cc_autosig(df_res, df_cc, poppy, win_size = 10, sig_frac = 8 / 10))
  # }
  
  df_resid %>%
    pivot_wider(id_cols = rp:status, names_from = pop, values_from = res) %>%
    select(ALL:SAS) %>%
    correlate(method = "pearson", diagonal = 1) %>%
    shave() %>%
    fashion(decimals = 3) %>%
    print()
  
  df_resid %>%
    pivot_wider(id_cols = rp:status, names_from = pop, values_from = res) %>%
    select(ALL:SAS) %>%
    correlate(method = "spearman", diagonal = 1) %>%
    shave() %>%
    fashion(decimals = 3) %>%
    print()
  
  resid_plots <- residual_analysis(df_resid, subtype)

  print(resid_plots$p1)
  print(resid_plots$p2)
  print(resid_plots$p3)
  print(resid_plots$p4)
}

```


## All Single Position RE Correlations

```{r}
df_cor <- single_pos_resid_table(single_pos_dir)

df_cor |> 
  pivot_wider(names_from = res_type, values_from = value) |>
  tbl_summary(include = c(pearson, spearman), 
              by = range, 
              statistic = all_continuous() ~ "{min}-{max} ({mean})")

df_cor |> 
  pivot_wider(names_from = res_type, values_from = value) |>
  tbl_strata(strata = subtype,
             .tbl_fun = 
               ~ .x |>
               tbl_summary(by = range, include = c(pearson, spearman),
                           statistic = all_continuous() ~ "{min}-{max} ({mean})"))
```

## All Single Position Residual Correlations

```{r}
res_cor_pearson <- vector("list", length(subtypes))
names(res_cor_pearson) <- subtypes

res_cor_spearman <- vector("list", length(subtypes))
names(res_cor_spearman) <- subtypes

for(st in subtypes){
  if(str_starts(st, "cpg")) rps <- c(-10:-1,2:10)
  else rps = c(-10:-1,1:10)
  df_resid <- sp_resid_all_pop(single_pos_dir, st, rp = rps)
  res_cor_analysis <- residual_analysis(df_resid, st, no_plots = T)
  
  res_cor_pearson[[st]] <- res_cor_analysis$df_pearson
  res_cor_spearman[[st]] <- res_cor_analysis$df_spearman
}

res_cor_pearson %>%
  bind_rows() %>%
  select(-rp) %>%
  summary() %>%
  knitr::kable()

res_cor_pearson %>%
  bind_rows() %>%
  select(-rp) %>%
  unlist() %>%
  summary() %>% t() %>%
  knitr::kable()

res_cor_spearman %>%
  bind_rows() %>%
  select(-rp) %>%
  summary() %>%
  knitr::kable()

res_cor_spearman %>%
  bind_rows() %>%
  select(-rp) %>%
  unlist() %>%
  summary() %>% t() %>%
  knitr::kable()
```

### Tables by Subtype

```{r results='asis'}
for(st in subtypes){
  cat(paste0("#### Subtype: ", st, "\n"))
  
  res_cor_pearson[[st]] %>%
    select(-rp) %>%
    summary() %>%
    knitr::kable() %>%
    print()
  
  cat("\n")
  
  res_cor_pearson[[st]] %>%
    select(-rp) %>%
    unlist() %>%
    summary() %>% t() %>%
    knitr::kable() %>%
    print()
  
  cat("\n")
  
  res_cor_spearman[[st]] %>%
    select(-rp) %>%
    summary() %>%
    knitr::kable() %>%
    print()
  
  cat("\n")
  
  res_cor_spearman[[st]] %>%
    select(-rp) %>%
    unlist() %>%
    summary() %>% t() %>%
    knitr::kable() %>%
    print()
  
  cat("\n")
  
}
```

## Shared Directionality of Nucleotide Level Residuals

```{r results='asis'}
for(st in subtypes){
  if(str_starts(st, "cpg")) rps <- c(-10:-1,2:10)
  else rps = c(-10:-1,1:10)
  df_resid <- sp_resid_all_pop(single_pos_dir, st, rp = rps)
  
  p <- df_resid |>
    filter(status == "singletons", pop != "ALL") |>
    pivot_wider(names_from = pop, values_from = res) |>
    mutate(AFR_AMR = sign(AFR) == sign(AMR),
           AFR_EAS = sign(AFR) == sign(EAS),
           AFR_EUR = sign(AFR) == sign(EUR),
           AFR_SAS = sign(AFR) == sign(SAS),
           AMR_EAS = sign(AMR) == sign(EAS),
           AMR_EUR = sign(AMR) == sign(EUR),
           AMR_SAS = sign(AMR) == sign(SAS),
           EAS_EUR = sign(EAS) == sign(EUR),
           EAS_SAS = sign(EAS) == sign(SAS),
           EUR_SAS = sign(EUR) == sign(SAS)
           ) |>
    group_by(rp) |>
    summarize(AFR_AMR = sum(AFR_AMR),
              AFR_EAS = sum(AFR_EAS),
              AFR_EUR = sum(AFR_EUR),
              AFR_SAS = sum(AFR_SAS),
              AMR_EAS = sum(AMR_EAS),
              AMR_EUR = sum(AMR_EUR),
              AMR_SAS = sum(AMR_SAS),
              EAS_EUR = sum(EAS_EUR),
              EAS_SAS = sum(EAS_SAS),
              EUR_SAS = sum(EUR_SAS)) |>
    pivot_longer(-rp, names_to = "pops", values_to = "matches") |>
    separate_wider_delim(pops, delim = "_", names = c("pop1", "pop2")) |>
    ggplot(aes(x = rp, y = matches)) +
    geom_point() +
    ylim(c(0,4)) +
    facet_grid(rows = vars(pop2), cols = vars(pop1)) +
    xlab("Relative Position") +
    ylab("Nucleotides with Matching Sign") +
    ggtitle(st)
  print(p)
  
  ## For non-match, are these smaller in absolute value?
  test_nucleotides <- df_resid |>
    filter(status == "singletons", pop != "ALL") |>
    pivot_wider(names_from = pop, values_from = res) |>
    mutate(AFR_AMR = sign(AFR) == sign(AMR),
           AFR_EAS = sign(AFR) == sign(EAS),
           AFR_EUR = sign(AFR) == sign(EUR),
           AFR_SAS = sign(AFR) == sign(SAS),
           AMR_EAS = sign(AMR) == sign(EAS),
           AMR_EUR = sign(AMR) == sign(EUR),
           AMR_SAS = sign(AMR) == sign(SAS),
           EAS_EUR = sign(EAS) == sign(EUR),
           EAS_SAS = sign(EAS) == sign(SAS),
           EUR_SAS = sign(EUR) == sign(SAS)) %>%
    select(-AFR, -AMR, -EAS, -EUR, -SAS, -status) |>
    pivot_longer(AFR_AMR:EUR_SAS, names_to = "pops", values_to = "match") |>
    filter(!match) |>
    separate_wider_delim(pops, delim = "_", names = c("pop1", "pop2"))
  
  df_resid_rank <- df_resid |>
    filter(status == "singletons", pop != "ALL") |>
    group_by(rp, pop) %>%
    mutate(nuc_rank = rank(abs(res)))
  
  test_nucleotides <- test_nucleotides %>%
    left_join(df_resid_rank, by = join_by(rp == rp, nuc == nuc, pop1 == pop)) %>%
    rename(rank1 = nuc_rank) |>
    left_join(df_resid_rank, by = join_by(rp == rp, nuc == nuc, pop2 == pop)) %>%
    rename(rank2 = nuc_rank)
  
  test_nucleotides |>
    select(rp, rank1, rank2) %>%
    pivot_longer(-rp, values_to = "rank") |>
    tbl_summary(include = rank) |>
    print()
}
```


## Single Position Results

### A > C

```{r}
subtype <- "AT_CG"


single_pos_analysis(subtype, sig_frac = 5/10)
```

```{r}
df_res <- sp_re_all_pop(single_pos_dir, subtype)
df_resid <- sp_resid_all_pop(single_pos_dir, subtype, rp = c(-10:-1, 1:10))
df_cc <- get_pos_cc_pops(control_control, subtype)
res_cor_analysis <- residual_analysis(df_resid, subtype)

# resid_plots <- residual_analysis(df_resid, subtype)
# 
# print(resid_plots$p1)
# print(resid_plots$p2)
# print(resid_plots$p3)
# print(resid_plots$p4)

# for(poppy in c("ALL", "AFR", "AMR", "EAS", "EUR", "SAS")){
#   print(poppy)
#   print(plot_re_cc_autosig(df_res, df_cc, poppy))
# }

df_res %>%
  filter(abs(offset) <= 20) %>%
  group_by(pop) %>%
  mutate(rank_re = dense_rank(desc(re))) %>%
  mutate(pop = factor(pop, levels = c("ALL", "AFR", "AMR", "EAS", "EUR", "SAS"))) %>%
  filter(pop != "ALL") %>%
  ggplot(aes(x = rank_re, y = offset, colour = pop)) +
  geom_point(alpha = 0.5) +
  geom_line() +
  scale_colour_manual(values = friendly_pal("vibrant_seven")) +
  theme_classic() +
  labs(colour = "Population") +
  ylab("Relative Position") +
  xlab("Rank")
  

plot_re_cc_autosig(df_res, df_cc, "SAS")
```


For the A > C subtype, we observe a high correlation of relative entropy statistics at the +/- 20

### A > G

```{r}
subtype <- "AT_GC"
single_pos_analysis(subtype)
```

### A > T

```{r}
subtype <- "AT_TA"
single_pos_analysis(subtype)
```

### C > T

```{r}
subtype <- "GC_AT"
single_pos_analysis(subtype)
```

### C > A

```{r}
subtype <- "GC_TA"
single_pos_analysis(subtype)
```

### C > G

```{r}
subtype <- "GC_CG"
single_pos_analysis(subtype)
```

### CpG > TpG

```{r}
subtype <- "cpg_GC_AT"
single_pos_analysis(subtype, rp = c(-10:-1,2:10))
```

### CpG > ApG

```{r}
subtype <- "cpg_GC_TA"
single_pos_analysis(subtype, rp = c(-10:-1,2:10))
```

### CpG > GpG

```{r}
subtype <- "cpg_GC_CG"
single_pos_analysis(subtype, rp = c(-10:-1,2:10))
```

## Single Position Pop-Pop Model

In the following section, we utilize our log-linear model framework to assess consistency of single position model results by fitting models of the following form (in R's model notation):

```
glm(count ~ nuc*status + nuc*population + status*population, family = poisson())
```

Here, in contrast to our single population models, we allow for the distributions of nucleotides at flanking positions to differ between the singletons and the controls via the inclusion of a nucleotide-by-status interaction. The devinace statistic for this model evaluates the fit of this model to a fully-saturated model, in which in addition to the above terms is a three-way interaction between nucleotide, population, and status; under the null hypothesis that these parameters are 0, we assume that any nucleotide by status interaction is shared across the populations. Note that we fit these models for pairs of populations, so the degree of freedom for this test statistic is three (three two-way interactions between nucleotide and status, one population parameter to interact with).

```{r}
pop_combs <- combn(c("AFR", "AMR", "EAS", "EUR", "SAS"), 2) %>% t() 
```


### A > C

```{r}
subtype <- "AT_CG"

pop_pop_analysis <- function(pop_pop_dir, pop1, pop2, subtype, rps = c(-20:-1, 1:20)){
  df <- load_pop_pop_df(pop_pop_dir, pop1, pop2, subtype, rps)

  p <- df %>%
    ggplot(aes(x = rp, y = dev, colour = is_sig)) +
    geom_point() +
    scale_color_manual(values = friendly_pals$wong_eight[c(1:2)]) +
    theme_classic() +
    theme(text = element_text(family = "Arial"),
          legend.position = "inside",
          legend.position.inside = c(.85, .8)) +
    labs(colour = "Significant") +
    xlab("Relative Position") +
    ylab("Deviance") + 
    ggtitle(paste0(subtype_print_names(subtype), ": ", pop1, ", ", pop2))
  return(p)
}

for(i in 1:dim(pop_combs)[1]){
  pop1 <- pop_combs[i,1]
  pop2 <- pop_combs[i,2]
  p <- pop_pop_analysis(pop_pop_dir, pop1, pop2, subtype, rps = c(-10:-1, 1:10)) +
    geom_hline(yintercept = qchisq(0.95, 3), color = "red", linetype = 2) 
  print(p)
}
```


### A > G

```{r}
subtype <- "AT_GC"

for(i in 1:dim(pop_combs)[1]){
  pop1 <- pop_combs[i,1]
  pop2 <- pop_combs[i,2]
  p <- pop_pop_analysis(pop_pop_dir, pop1, pop2, subtype, rps = c(-10:-1, 1:10)) +
    geom_hline(yintercept = qchisq(0.95, 3), color = "red", linetype = 2) 
  print(p)
}
```

### A > T

```{r}
subtype <- "AT_TA"

for(i in 1:dim(pop_combs)[1]){
  pop1 <- pop_combs[i,1]
  pop2 <- pop_combs[i,2]
  p <- pop_pop_analysis(pop_pop_dir, pop1, pop2, subtype, rps = c(-10:-1, 1:10)) +
    geom_hline(yintercept = qchisq(0.95, 3), color = "red", linetype = 2) 
  print(p)
}
```

### C > T

```{r}
subtype <- "GC_AT"

for(i in 1:dim(pop_combs)[1]){
  pop1 <- pop_combs[i,1]
  pop2 <- pop_combs[i,2]
  p <- pop_pop_analysis(pop_pop_dir, pop1, pop2, subtype, rps = c(-10:-1, 1:10)) +
    geom_hline(yintercept = qchisq(0.95, 3), color = "red", linetype = 2) 
  print(p)
}
```

### C > A

```{r}
subtype <- "GC_TA"

for(i in 1:dim(pop_combs)[1]){
  pop1 <- pop_combs[i,1]
  pop2 <- pop_combs[i,2]
  p <- pop_pop_analysis(pop_pop_dir, pop1, pop2, subtype, rps = c(-10:-1, 1:10)) +
    geom_hline(yintercept = qchisq(0.95, 3), color = "red", linetype = 2) 
  print(p)
}
```

### C > G

```{r}
subtype <- "GC_CG"

for(i in 1:dim(pop_combs)[1]){
  pop1 <- pop_combs[i,1]
  pop2 <- pop_combs[i,2]
  p <- pop_pop_analysis(pop_pop_dir, pop1, pop2, subtype, rps = c(-10:-1, 1:10)) +
    geom_hline(yintercept = qchisq(0.95, 3), color = "red", linetype = 2) 
  print(p)
}
```

### CpG > TpG

```{r}
subtype <- "GC_AT"

for(i in 1:dim(pop_combs)[1]){
  pop1 <- pop_combs[i,1]
  pop2 <- pop_combs[i,2]
  p <- pop_pop_analysis(pop_pop_dir, pop1, pop2, subtype, rps = c(-10:-1, 2:10)) +
    geom_hline(yintercept = qchisq(0.95, 3), color = "red", linetype = 2) 
  print(p)
}
```

### CpG > ApG

```{r}
subtype <- "GC_TA"

for(i in 1:dim(pop_combs)[1]){
  pop1 <- pop_combs[i,1]
  pop2 <- pop_combs[i,2]
  p <- pop_pop_analysis(pop_pop_dir, pop1, pop2, subtype, rps = c(-10:-1, 2:10)) +
    geom_hline(yintercept = qchisq(0.95, 3), color = "red", linetype = 2) 
  print(p)
}
```

### CpG > GpG

```{r}
subtype <- "GC_CG"

for(i in 1:dim(pop_combs)[1]){
  pop1 <- pop_combs[i,1]
  pop2 <- pop_combs[i,2]
  p <- pop_pop_analysis(pop_pop_dir, pop1, pop2, subtype, rps = c(-10:-1, 2:10)) +
    geom_hline(yintercept = qchisq(0.95, 3), color = "red", linetype = 2) 
  print(p)
}
```

## Single Position Window Selection

### A > C

```{r}
subtype <- "AT_CG"
df_res <- sp_re_all_pop(single_pos_dir, subtype)
df_resid <- sp_resid_all_pop(single_pos_dir, subtype, rp = c(-10:-1, 1:10))
df_cc <- get_pos_cc_pops(control_control, subtype)
```


#### ALL

```{r}
population <- "ALL"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)
```

Using 0.7 as our threshold appears to have a good tradeoff between having interior 95%+ significant and less than 3% exterior significant:

```{r}
df_window_plots <- function(df_window, df_res, population, subtype, plot_cutoff = 0.8){
  #print("Hello!")
  for(i in 1:length(df_window$frac_thresh)){
    
    if(df_window$frac_in[i] < plot_cutoff) next
    
    thresh_cutoff <- df_window$frac_thresh[i]
    left_boundary <- df_window %>% filter(frac_thresh == thresh_cutoff) %>% pull(left_boundary)
    right_boundary <- df_window %>% filter(frac_thresh == thresh_cutoff) %>% pull(right_boundary)
    
    p <- df_res %>%
      filter(pop == population) %>%
      mutate(is_sig = as.character(is_sig)) %>%
      bind_rows({df_cc %>% filter(pop == population) %>% mutate(is_sig = "Control")}) %>%
      ggplot(aes(x = offset, y = re, colour = is_sig, alpha = is_sig)) +
      geom_point() +
      scale_alpha_manual(values = c(0.2, 1, 1)) + guides(alpha="none") + 
      scale_color_manual(values = friendly_pals$wong_eight[c(3,1:2)]) +
      theme_classic() +
      theme(text = element_text(family = "Arial"),
          legend.position = "inside",
          legend.position.inside = c(.85, .8)) +
      labs(colour = "Significant") +
      xlab("Relative Position") +
      ylab("Relative Entropy (log10)") +
      scale_y_log10() +
      geom_vline(xintercept = left_boundary, colour = friendly_pals$wong_eight[2]) +
      geom_vline(xintercept = right_boundary, colour = friendly_pals$wong_eight[2]) +
      annotate("text", x = left_boundary - 110, y = min(df_res$re), label = left_boundary) + 
      annotate("text", x = right_boundary + 100, y = min(df_res$re), label = right_boundary) +
      ggtitle(paste0("Population: ", population, "; Subtype: ", subtype_print_names(subtype)),
              paste0("Cutoff: ", thresh_cutoff,
                "; Frac Interior: ", 
                     round(df_window$frac_in[i],2), 
                     "; Frac Exterior: ", 
                     round(df_window$frac_out[i], 2) 
                     )
              )
    print(p)
  }
}

df_window_plots(df_window, df_res, population, subtype)
```

#### AFR

```{r}
population <- "AFR"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```

#### AMR

```{r}
population <- "AMR"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```

#### EAS

```{r}
population <- "EAS"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```

#### EUR

```{r}
population <- "EUR"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```

#### SAS

```{r}
population <- "SAS"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```

### A > G

```{r}
subtype <- "AT_GC"
df_res <- sp_re_all_pop(single_pos_dir, subtype)
df_resid <- sp_resid_all_pop(single_pos_dir, subtype, rp = c(-10:-1, 1:10))
df_cc <- get_pos_cc_pops(control_control, subtype)
```

#### ALL

```{r}
population <- "ALL"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```

#### AFR

```{r}
population <- "AFR"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```


#### AMR

```{r}
population <- "AMR"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```


#### EAS

```{r}
population <- "EAS"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```


#### EUR

```{r}
population <- "EUR"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```


#### SAS

```{r}
population <- "SAS"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```

### A > T

```{r}
subtype <- "AT_TA"
df_res <- sp_re_all_pop(single_pos_dir, subtype)
df_resid <- sp_resid_all_pop(single_pos_dir, subtype, rp = c(-10:-1, 1:10))
df_cc <- get_pos_cc_pops(control_control, subtype)
```

#### ALL

```{r}
population <- "ALL"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```

#### AFR

```{r}
population <- "AFR"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```


#### AMR

```{r}
population <- "AMR"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```


#### EAS

```{r}
population <- "EAS"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```


#### EUR

```{r}
population <- "EUR"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```


#### SAS

```{r}
population <- "SAS"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```

### C > T

```{r}
subtype <- "GC_AT"
df_res <- sp_re_all_pop(single_pos_dir, subtype)
df_resid <- sp_resid_all_pop(single_pos_dir, subtype, rp = c(-10:-1, 1:10))
df_cc <- get_pos_cc_pops(control_control, subtype)
```

#### ALL

```{r}
population <- "ALL"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```

#### AFR

```{r}
population <- "AFR"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```


#### AMR

```{r}
population <- "AMR"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```


#### EAS

```{r}
population <- "EAS"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```


#### EUR

```{r}
population <- "EUR"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```


#### SAS

```{r}
population <- "SAS"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```

### C > A

```{r}
subtype <- "GC_TA"
df_res <- sp_re_all_pop(single_pos_dir, subtype)
df_resid <- sp_resid_all_pop(single_pos_dir, subtype, rp = c(-10:-1, 1:10))
df_cc <- get_pos_cc_pops(control_control, subtype)
```

#### ALL

```{r}
population <- "ALL"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```

#### AFR

```{r}
population <- "AFR"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```


#### AMR

```{r}
population <- "AMR"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```


#### EAS

```{r}
population <- "EAS"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```


#### EUR

```{r}
population <- "EUR"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```


#### SAS

```{r}
population <- "SAS"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```

### C > G

```{r}
subtype <- "GC_CG"
df_res <- sp_re_all_pop(single_pos_dir, subtype)
df_resid <- sp_resid_all_pop(single_pos_dir, subtype, rp = c(-10:-1, 1:10))
df_cc <- get_pos_cc_pops(control_control, subtype)
```

#### ALL

```{r}
population <- "ALL"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```

#### AFR

```{r}
population <- "AFR"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```


#### AMR

```{r}
population <- "AMR"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```


#### EAS

```{r}
population <- "EAS"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```


#### EUR

```{r}
population <- "EUR"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```


#### SAS

```{r}
population <- "SAS"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```

### CpG > TpG

```{r}
subtype <- "cpg_GC_AT"
df_res <- sp_re_all_pop(single_pos_dir, subtype)
df_resid <- sp_resid_all_pop(single_pos_dir, subtype, rp = c(-10:-1, 1:10))
df_cc <- get_pos_cc_pops(control_control, subtype)
```

#### ALL

```{r}
population <- "ALL"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```

#### AFR

```{r}
population <- "AFR"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```


#### AMR

```{r}
population <- "AMR"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```


#### EAS

```{r}
population <- "EAS"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```


#### EUR

```{r}
population <- "EUR"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```


#### SAS

```{r}
population <- "SAS"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```

### CpG > ApG

```{r}
subtype <- "cpg_GC_TA"
df_res <- sp_re_all_pop(single_pos_dir, subtype)
df_resid <- sp_resid_all_pop(single_pos_dir, subtype, rp = c(-10:-1, 1:10))
df_cc <- get_pos_cc_pops(control_control, subtype)
```

#### ALL

```{r}
population <- "ALL"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```

#### AFR

```{r}
population <- "AFR"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```


#### AMR

```{r}
population <- "AMR"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```


#### EAS

```{r}
population <- "EAS"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```


#### EUR

```{r}
population <- "EUR"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```


#### SAS

```{r}
population <- "SAS"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```

### CpG > GpG

```{r}
subtype <- "cpg_GC_CG"
df_res <- sp_re_all_pop(single_pos_dir, subtype)
df_resid <- sp_resid_all_pop(single_pos_dir, subtype, rp = c(-10:-1, 1:10))
df_cc <- get_pos_cc_pops(control_control, subtype)
```

#### ALL

```{r}
population <- "ALL"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```

#### AFR

```{r}
population <- "AFR"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```


#### AMR

```{r}
population <- "AMR"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```


#### EAS

```{r}
population <- "EAS"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```


#### EUR

```{r}
population <- "EUR"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```


#### SAS

```{r}
population <- "SAS"

df_window <- get_df_windows(df_res, df_cc, population, thresh_vals = (1:10)/10)

df_window %>% reactable(pagination = FALSE)

df_window_plots(df_window, df_res, population, subtype)
```

## Two Position Models

### Functions

```{r}
tp_all_dev <- function(two_pos_dir, subtype){
  pops <- c("ALL", "AFR", "AMR", "EAS", "EUR", "SAS")
  results <- tp_deviance_df_load(two_pos_dir, pops[1], subtype) %>%
    select(rp1, rp2, dev, re) %>%
    mutate(pop = pops[1])
  for(i in 2:length(pops)){
    results <- results %>%
      bind_rows({
        tp_deviance_df_load(two_pos_dir, pops[i], subtype) %>%
          select(rp1, rp2, dev, re) %>%
          mutate(pop = pops[i])
      })
  }
  results <- results %>% distinct(rp1, rp2, pop, .keep_all = TRUE) %>%
    mutate(pop = factor(pop, levels = c("ALL", "AFR", "AMR", "EAS", "EUR", "SAS")))
  return(results)
}

tp_re_cor_plot <- function(df_tp, subtype, method="pearson"){
  cor_obj <- df_tp %>%
    select(-dev) %>%
    pivot_wider(names_from = pop, values_from = re, id_cols = c(rp1, rp2)) %>%
    select(ALL:SAS) %>%
    corrr::correlate(method = method) %>%
    corrr::shave()
  
  p <- cor_obj %>% reshape2::melt() %>%
    mutate(term = factor(term, levels = c("ALL", "AFR", "AMR", "EAS", "EUR", "SAS"))) %>%
    filter(variable != term) %>%
    filter(term != "ALL", variable != "SAS") %>%
    mutate(round_val = round(value, 2)) %>%
    ggplot(aes(x=term, y=variable, fill = value)) +
    geom_tile() +
    scale_fill_distiller(palette = "Reds", na.value = "white", direction = 1, limits = c(0,1)) +
    theme_classic() +
    xlab("") +
    ylab("") +
    geom_text(aes(label=round_val)) +
    ggtitle(subtype_print_names(subtype))

  return(p)
}
```

### Get table for all pairs of populations

```{r}
two_pos_resid_table <- function(two_pos_dir){
  subtypes <- c("AT_CG", "AT_GC", "AT_TA",
                "GC_AT", "GC_TA", "GC_CG",
                "cpg_GC_AT", "cpg_GC_TA", "cpg_GC_CG")
  
  cor_types <- c("pearson", "spearman")
  
  pops <- c("AFR", "AMR", "EAS", "EUR", "SAS")
  pop_combs <- combn(pops,2)
  pops_df <- data.frame(pop1 = pop_combs[1,],
                        pop2 = pop_combs[2,])
  
  final <- data.frame(subtype = character(0), 
                      res_type = character(0),
                      pop1 = character(0),
                      pop2 = character(0),
                      value = numeric(0))
  
  for(st in subtypes){
    df_res <- tp_all_dev(two_pos_dir, st)
    
    for(i in 1:length(pops_df$pop1)){
      pop1 <- pops_df$pop1[i]
      pop2 <- pops_df$pop2[i]
      df_res2 <- df_res %>%
        filter(pop == pop1 | pop == pop2) %>%
        select(rp1, rp2, re, pop)
      
      for(ct in cor_types){
        # All positions
        value <- df_res2 %>%
          pivot_wider(id_cols = c("rp1", "rp2"), names_from = "pop", values_from = re) %>%
          {cor(.[,3], .[,4], method = ct)} %>%
          as.numeric()
        final <- bind_rows(final, 
                           data.frame(subtype = st, 
                                      res_type = ct, 
                                      pop1 = pop1, 
                                      pop2=pop2, 
                                      value = value))
      }
    }
  }
  return(final)
}

df_tp_cor <- two_pos_resid_table(two_pos_dir)

write_csv(df_tp_cor, "output/two_pos_pop_cor.csv")

df_tp_cor |> 
  tbl_summary(include = value,
              by = res_type,
              statistic = all_continuous() ~ "{min}-{max} ({mean})")

df_tp_cor |> 
  pivot_wider(names_from = res_type, values_from = value) |>
  tbl_summary(include = c(pearson, spearman), by = subtype,
              statistic = all_continuous() ~ "{min}-{max} ({mean})")

```


### A > C

```{r, warning=FALSE, message=FALSE}
subtype <- "AT_CG"

df_tp <- tp_all_dev(two_pos_dir, subtype)
tp_re_cor_plot(df_tp, subtype)
tp_re_cor_plot(df_tp, subtype, "spearman")
```

### A > G

```{r, warning=FALSE, message=FALSE}
subtype <- "AT_GC"

df_tp <- tp_all_dev(two_pos_dir, subtype)
tp_re_cor_plot(df_tp, subtype)
tp_re_cor_plot(df_tp, subtype, "spearman")
```

### A > T

```{r, warning=FALSE, message=FALSE}
subtype <- "AT_TA"

df_tp <- tp_all_dev(two_pos_dir, subtype)
tp_re_cor_plot(df_tp, subtype)
tp_re_cor_plot(df_tp, subtype, "spearman")
```

### C > T

```{r, warning=FALSE, message=FALSE}
subtype <- "GC_AT"

df_tp <- tp_all_dev(two_pos_dir, subtype)
tp_re_cor_plot(df_tp, subtype)
tp_re_cor_plot(df_tp, subtype, "spearman")
```

### C > A

```{r, warning=FALSE, message=FALSE}
subtype <- "GC_TA"

df_tp <- tp_all_dev(two_pos_dir, subtype)
tp_re_cor_plot(df_tp, subtype)
tp_re_cor_plot(df_tp, subtype, "spearman")
```

### C > G

```{r, warning=FALSE, message=FALSE}
subtype <- "GC_CG"

df_tp <- tp_all_dev(two_pos_dir, subtype)
tp_re_cor_plot(df_tp, subtype)
tp_re_cor_plot(df_tp, subtype, "spearman")
```

### CpG > TpG

```{r, warning=FALSE, message=FALSE}
subtype <- "cpg_GC_AT"

df_tp <- tp_all_dev(two_pos_dir, subtype)
tp_re_cor_plot(df_tp, subtype)
tp_re_cor_plot(df_tp, subtype, "spearman")
```

### CpG > ApG

```{r, warning=FALSE, message=FALSE}
subtype <- "cpg_GC_TA"

df_tp <- tp_all_dev(two_pos_dir, subtype)
tp_re_cor_plot(df_tp, subtype)
tp_re_cor_plot(df_tp, subtype, "spearman")
```

### CpG > GpG

```{r, warning=FALSE, message=FALSE}
subtype <- "cpg_GC_CG"

df_tp <- tp_all_dev(two_pos_dir, subtype)
tp_re_cor_plot(df_tp, subtype)
tp_re_cor_plot(df_tp, subtype, "spearman")
```
