---
title: "single_position_results"
author: "Andy Beck"
date: "2025-01-06"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Notes

* Save figures using something along the lines of `ggsave("file.png", dpi=300)`

## Setup

### Libraries

```{r}
library(tidyverse)
library(ggpubfigs) # functions for color blind friendly palettes
library(reactable) # table output for html
library(knitr) # PDF table output
```

### Paths

```{r}
base_dir = "output"
single_pos_dir = paste0(base_dir, "/single_pos")
control_control <- paste0(base_dir, "/single_pos_cc")

figure_output = paste0(base_dir, "/figs/single_pos")
```


### Base Functions

Load position-level results:

```{r}
subtypes <- c("AT_CG", "AT_GC", "AT_TA",
            "GC_TA", "GC_CG", "GC_AT",
            "cpg_GC_TA", "cpg_GC_CG", "cpg_GC_AT")

get_pos_results <- function(res_dir, pop, subtype){
  f_name <- paste0(res_dir, "/", pop, "/", subtype, ".csv")
  df <- read_csv(f_name, show_col_types = FALSE) %>%
    rowwise() %>%
    mutate(re = dev / (2 * (singletons+controls)))
  return(df)
}

get_pos_cc <- function(control_control, pop, subtype){
  f_name <- paste0(control_control, "/", pop, "/", subtype, "_v2.csv")
  df <- read_csv(f_name, show_col_types = FALSE) %>%
           mutate(re = dev / (2 * (singletons + controls) ))
  return(df)
}

get_pos_results_all <- function(res_dir, pop, subtypes){
  results <- vector(mode = "list", length = length(subtypes))
  
  for(i in 1:length(subtypes)){
    results[[i]] <- get_pos_results(res_dir, pop, subtypes[[i]])
    results[[i]]$subtype <- subtypes[[i]]
  }
  
  return(bind_rows(results))
}

subtype_print_names <- function(st){
  if(str_starts(st, "AT")){
    return(paste0("A → ", str_sub(st, 4, 4)))
  } else if(str_starts(st, "GC")){
    return(paste0("C → ", str_sub(st, 5, 5)))
  } else{
    return(paste0("CpG → ", str_sub(st, 9, 9), "pG"))
  }
}
```

Load position-level residuals:

```{r}
get_pos_resid <- function(res_dir, pop, subtype, rp){
  f_name <- paste0(res_dir, "/resid/", pop, "/", subtype, "_rp_", rp, ".csv")
  df <- read_csv(f_name, show_col_types = FALSE)
  df$rp <- rp
  return(df)
}

get_pos_resid_all <- function(res_dir, pop, subtype, rp = c(-500:-1, 1:500)){
  results <- vector(mode = "list", length = length(rp))
  for(i in 1:length(rp)){
    if(rp[i] == 1 & str_starts(subtype, "cpg")) next
    results[[i]] <- get_pos_resid(res_dir, pop, subtype, rp[i])
  }
  df <- bind_rows(results)
  return(df)
}
```

### Figure Functions

```{r}
pos_level_line <- function(res_df, rp_lim = 20, show_line = T){
  p <- res_df %>%
    filter(abs(offset) <= rp_lim) %>%
    ggplot(aes(x = offset, y = re)) +
    geom_point() +
    theme_classic() +
    theme(text = element_text(family = "Arial"),
          legend.position = "inside",
          legend.position.inside = c(.85, .8)) +
    xlab(paste0("Relative Position: ", subtype_print_names(subtype))) +
    ylab("Relative Entropy")
  if(show_line){
    p <- p + geom_line()
  }
  return(p)
}

pos_level_bar <- function(res_df, dev=FALSE, rp_lim = 10){
  if(dev){
    var_plot <- "dev"
    y_text <- "Deviance"
  } else{
    var_plot <- "re"
    y_text <- "Relative Entropy"
  }
  p <- res_df %>%
    filter(abs(offset) <= rp_lim) %>%
    ggplot(aes(x = offset, y = !!sym(var_plot))) +
    geom_col() +
    theme_classic() +
    theme(text = element_text(family = "Arial"),
          legend.position = "inside",
          legend.position.inside = c(.85, .8)) +
    xlab(paste0("Relative Position: ", subtype_print_names(subtype))) +
    ylab(y_text)
  return(p)
}

pos_level_dot <- function(res_df, palette_name = friendly_pal("contrast_three")){
  p <- res_df %>%
    mutate(sig = dev > qchisq(0.95, 3)) %>%
    ggplot(aes(x = offset, y = re, color = sig)) +
    scale_y_log10() +
    geom_point() +
    scale_color_manual(values = palette_name) +
    theme_classic() +
    theme(text = element_text(family = "Arial"),
          legend.position = "inside",
          legend.position.inside = c(.85, .8)) +
    xlab(paste0("Relative Position: ", subtype_print_names(subtype))) +
    ylab("Log10 Relative Entropy") +
    labs(color = "Significant")
  return(p)
}

pos_level_dot_cc <- function(res_df, color_pal = friendly_pals$wong_eight[c(3,1:2)]){
  p <- res_df %>%
    ggplot(aes(x = offset, y = re, color = sig, alpha = sig)) +
    scale_y_log10() +
    geom_point() +
    scale_alpha_manual(values = c(0.2, 1, 1)) + guides(alpha="none") +
    scale_color_manual(values = color_pal) +
    theme_classic() +
    theme(text = element_text(family = "Arial"),
          legend.position = "inside",
          legend.position.inside = c(.85, .8)) +
    xlab(paste0("Relative Position: ", subtype_print_names(subtype))) +
    ylab("Log10 Relative Entropy") +
    labs(color = "Significant")
  return(p)
}

resid_plot <- function(res_dir, pop, subtype, rp = c(-10:-1, 1:10), palette_name = friendly_pal("nickel_five")){
  res_df <- get_pos_resid_all(res_dir, pop, subtype, rp)
  
  res_df$ID <- as.character(res_df$rp)
  xlab <- as.character(min(res_df$rp):max(res_df$rp))
  res_df$ID <- factor(res_df$ID, levels = xlab)
  
  p <- res_df %>%
    select(ID, nuc, status, res) %>%
    pivot_wider(names_from = status, values_from = res) %>%
    rowwise() %>%
    mutate(dev_cont = sign(singletons) * (singletons ^ 2 + controls ^ 2)) %>%
    ggplot(aes(x = ID, y = dev_cont, fill = nuc)) +
    scale_x_discrete(labels= xlab, drop = FALSE) +
    geom_bar(position="stack", stat="identity") +
    scale_fill_manual(values = palette_name) +
    theme_classic() +
    xlab(paste0("Relative Position: ", subtype_print_names(subtype))) +
    ylab("Singleton Enrichment") +
    labs(fill = "Nucleotide") +
    theme(legend.position = "inside", legend.position.inside = c(0.9, 0.8),
          legend.box.background = element_rect(color="black", linewidth =1)) +
    geom_hline(yintercept = 0, colour = "black", linetype = 3)
  return(p)
}
```

## Summary Across All Subtypes

First let's get a look at the single position model results for all subtypes using the singletons from all 5 1kGP super-populations. 

```{r}
load_all_pos_results <- function(subtypes, pop = "ALL"){
  final <- get_pos_results(single_pos_dir, pop, subtypes[1]) %>%
    mutate(subtype = subtypes[1])
  
  for(i in 2:length(subtypes)){
    final <- final %>%
      bind_rows({
        get_pos_results(single_pos_dir, pop, subtypes[i]) %>%
          mutate(subtype = subtypes[i])
      })
  }
  return(final)
}

df_all <- load_all_pos_results(subtypes)

df_all <- df_all %>% 
  mutate(from_n = ifelse(str_starts(subtype, "A"), "A", ifelse(str_starts(subtype, "cpg"), "CpG", "C")),
         #to_n = ifelse(str_starts(subtype, "A"), str_sub(subtype, 4, 4), str_sub(subtype, -1))
         to_n = ifelse(str_starts(subtype, "A"), str_sub(subtype, 4,4), 
                       ifelse(str_starts(subtype, "cpg"), 
                              paste0(str_sub(subtype, -1), "pG"), 
                              str_sub(subtype, -1) )))
```

### Figures

#### Original Dot-Line

```{r}
p1 <- df_all %>%
  filter(from_n == "A", abs(offset) < 11) %>%
  ggplot(aes(x = offset, y = re)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  facet_grid(vars(to_n), vars(from_n)) +
  xlab("Relative Position") +
  ylab("Relative Entropy")

y_lim_all <- ggplot_build(p1)$layout$panel_scales_y[[1]]$range$range
y_lim_all[1] <- 0

p1 <- p1 + ylim(y_lim_all)

p2 <- df_all %>%
  filter(from_n == "C", abs(offset) < 11) %>%
  ggplot(aes(x = offset, y = re)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  facet_grid(vars(to_n), vars(from_n)) +
  xlab("Relative Position") +
  ylab("") +
  ylim(y_lim_all)

p3 <- df_all %>%
  filter(from_n == "CpG", abs(offset) < 11) %>%
  ggplot(aes(x = offset, y = re)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  facet_grid(vars(to_n), vars(from_n)) +
  xlab("Relative Position") +
  ylab("") +
  ylim(y_lim_all)

gridExtra::grid.arrange(p1, p2, p3, nrow = 1)
```

#### Dots Only

```{r}
p1 <- df_all %>%
  filter(from_n == "A", abs(offset) < 11) %>%
  ggplot(aes(x = offset, y = re)) +
  geom_point() +
  theme_bw() +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())+
  facet_grid(vars(to_n), vars(from_n)) +
  xlab("Relative Position") +
  ylab("Relative Entropy")

y_lim_all <- ggplot_build(p1)$layout$panel_scales_y[[1]]$range$range
y_lim_all[1] <- 0

p1 <- p1 + ylim(y_lim_all)

p2 <- df_all %>%
  filter(from_n == "C", abs(offset) < 11) %>%
  ggplot(aes(x = offset, y = re)) +
  geom_point() +
  theme_bw() +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())+
  facet_grid(vars(to_n), vars(from_n)) +
  xlab("Relative Position") +
  ylab("") +
  ylim(y_lim_all)

p3 <- df_all %>%
  filter(from_n == "CpG", abs(offset) < 11) %>%
  ggplot(aes(x = offset, y = re)) +
  geom_point() +
  theme_bw() +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())+
  facet_grid(vars(to_n), vars(from_n)) +
  xlab("Relative Position") +
  ylab("") +
  ylim(y_lim_all)

gridExtra::grid.arrange(p1, p2, p3, nrow = 1) 
```

#### Bars

```{r}
df_all$c_val <- factor(abs(df_all$offset))

p1 <- df_all %>%
  filter(from_n == "A", abs(offset) < 11) %>%
  ggplot(aes(x = offset, y = re*100, fill = c_val)) +
  geom_col() +
  theme_bw() +
  theme(panel.grid.major.x  = element_blank(),
        panel.grid.minor.x = element_blank())+
  facet_grid(vars(to_n), vars(from_n)) +
  xlab("Relative Position") +
  ylab("Relative Entropy") +
  scale_fill_manual(values = rev(friendly_pal("bright_seven", n = 10, type = "continuous"))) +
  guides(fill = "none")

y_lim_all <- ggplot_build(p1)$layout$panel_scales_y[[1]]$range$range
y_lim_all[1] <- 0

p1 <- p1 + ylim(y_lim_all)

p2 <- df_all %>%
  filter(from_n == "C", abs(offset) < 11) %>%
  ggplot(aes(x = offset, y = re*100, fill = c_val)) +
  geom_col() +
  theme_bw() +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())+
  facet_grid(vars(to_n), vars(from_n)) +
  xlab("Relative Position") +
  ylab("") +
  ylim(y_lim_all) +
  scale_fill_manual(values = rev(friendly_pal("bright_seven", n = 10, type = "continuous"))) +
  guides(fill = "none")

p3 <- df_all %>%
  filter(from_n == "CpG", abs(offset) < 11) %>%
  ggplot(aes(x = offset, y = re*100, fill = c_val)) +
  geom_col() +
  theme_bw() +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())+
  facet_grid(vars(to_n), vars(from_n)) +
  xlab("Relative Position") +
  ylab("") +
  ylim(y_lim_all) +
  scale_fill_manual(values = rev(friendly_pal("bright_seven", n = 10, type = "continuous"))) +
  guides(fill = "none")

gridExtra::grid.arrange(p1, p2, p3, nrow = 1)
```

#### Bars: 1 subtype at a time

```{r}
plot_list <- vector(mode = "list", length = 9)
names(plot_list) <- subtypes
pal_vec <- rev(friendly_pal("bright_seven", n = 10, type = "continuous"))

find_max_y_range <- function(plot_list) {
  y_ranges <- lapply(plot_list, function(plot) {
    y_range <- ggplot_build(plot)$layout$panel_scales_y[[1]]$range$range
    return(y_range)
  })

  max_y_min <- min(unlist(lapply(y_ranges, function(range) range[1])))
  max_y_max <- max(unlist(lapply(y_ranges, function(range) range[2])))

  return(c(max_y_min, max_y_max))
}

for(st in subtypes){
  # Get top three
  df_sub <- df_all %>%
    filter(subtype == st, abs(offset) < 11)
  
  top_three <- df_sub %>%
    arrange(desc(re)) %>% pull(offset) %>% .[1:3]
  
  plot_list[[st]] <- df_sub %>%
    ggplot(aes(x = offset, y = re*100, fill = c_val)) +
    geom_col() +
    theme_bw() +
    theme(panel.grid.major.x = element_blank(),
          panel.grid.minor.x = element_blank())+
    xlab("") +
    ylab("") +
    scale_fill_manual(values = pal_vec) +
    guides(fill = "none") +
    geom_text(
      label = "1",
      x = {df_sub %>% filter(offset == top_three[1]) %>% pull(offset) %>% as.numeric()},
      y = ({df_sub %>% filter(offset == top_three[1]) %>% pull(re) %>% as.numeric()}*100) + 0.05,
      color = pal_vec[abs({df_sub %>% filter(offset == top_three[1]) %>% pull(offset)})]
    ) +
    geom_text(
      label = "2",
      x = {df_sub %>% filter(offset == top_three[2]) %>% pull(offset) %>% as.numeric()},
      y = ({df_sub %>% filter(offset == top_three[2]) %>% pull(re) %>% as.numeric()}*100) + 0.05,
      color = pal_vec[abs({df_sub %>% filter(offset == top_three[2]) %>% pull(offset)})]
    ) +
    geom_text(
      label = "3",
      x = {df_sub %>% filter(offset == top_three[3]) %>% pull(offset) %>% as.numeric()},
      y = ({df_sub %>% filter(offset == top_three[3]) %>% pull(re) %>% as.numeric()}*100) + 0.05,
      color = pal_vec[abs({df_sub %>% filter(offset == top_three[3]) %>% pull(offset)})]
    ) 
    
}

y_lims <- find_max_y_range(plot_list)
y_lims[2] <- y_lims[2] + 0.05

for(st in subtypes){
  plot_list[[st]] <- plot_list[[st]] + 
    ylim(y_lims) +
    geom_text(label = subtype_print_names(st),
              x = -9.5,
              y = {y_lims[2]}-0.1, 
              hjust = "left")
}
```

```{r fig.height = 8, fig.width = 8, fig.align = "center"}
gridExtra::grid.arrange(grobs = plot_list, ncol = 3)
```

#### Resid: all st in one

```{r}
plot_list <- vector(mode = "list", length = 9)
names(plot_list) <- subtypes
pal_vec <- rev(friendly_pal("bright_seven", n = 10, type = "continuous"))

for(st in subtypes){
  plot_list[[st]] <- resid_plot(single_pos_dir, "ALL", st, rp = c(-5:-1,1:5)) +
    theme(legend.position="top", 
          legend.title=element_blank(), 
          legend.text = element_text(size=10),
          legend.key.size = unit(0.35, 'cm'), 
          legend.box.background = element_blank())
}
# y_lims <- find_max_y_range(plot_list)
# 
# for(st in subtypes){
#   plot_list[[st]] <- plot_list[[st]] + 
#     ylim(y_lims) 
# }
```

```{r fig.height = 8, fig.width = 8, fig.align = "center"}
gridExtra::grid.arrange(grobs = plot_list, ncol = 3)
```

### +/- 10 BP

```{r}
df_all %>%
  filter(abs(offset) <= 10) %>%
  select(offset, re, subtype) %>%
  pivot_wider(id_cols = offset, names_from = subtype, values_from = re) %>%
  arrange(offset)
```

Across all subtypes, what range of values for RE do we see within the +/- 10 bp window?

```{r}
df_all %>%
  filter(abs(offset) <= 10) %>%
  select(re, mean_var, max_var, simple_re) %>%
  summary() %>%
  knitr::kable()
```


## Position Level Result Plots

### A > G

```{r}
subtype <- "AT_GC"
```


#### ALL

```{r}
pop <- "ALL"
df_results <- get_pos_results(single_pos_dir, pop = pop, subtype = subtype)
df_cc <- get_pos_cc(control_control, pop, subtype)

df_results$sig  <- ifelse(df_results$dev > qchisq(0.95,3), "Significant", "Non-significant")
df_cc$sig <- "Control"
df <- bind_rows(df_results, df_cc)

pos_level_dot(df_results, friendly_pals$wong_eight)
pos_level_dot_cc(df) 

pos_level_line(df_results) 
pos_level_line(df_results) + scale_y_log10() 

pos_level_line(df_results, show_line = F)

pos_level_bar(df_results)
pos_level_bar(df_results) + scale_y_log10()

```

For the A > G mutation subtype, we observe the strongest signal of influence (relative entropy) at the +1 relative position (re = 0.0096). We observed that, in general, the signal decreases as positions further from the focal (0) position are considered, but note that this pattern is not monotonic in either direction, with the most obvious example of this being the stronger signal observed at -2 relative to the signal observed at -1. When compared to the critical value for a $\chi^2$ distribution with 3 degrees of freedom at the $\alpha=0.05$ level of significance, we find that the underlying deviance statistics remain significant up to approximately +/- 250bp; from here, we see values hover in a range that includes values which are not significant. The above results are based on the distributions of nucleotides flanking singletons and their matched controls across all five 1kGP super-populations.

Next, we plot the nucleotide contribution to each position's deviance statistic:

```{r}
df %>%
  filter(abs(offset) <= 10) %>%
  mutate(highlighted = offset==1) %>%
  ggplot(aes(x = offset, y = re, fill = highlighted)) +
  geom_col() +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  scale_fill_manual(values = c("#595959", "darksalmon")) +
  guides(fill = "none") +
  xlab("Relative Entropy") + 
  ylab("Relative Position")
```


```{r}
resid_plot(single_pos_dir, pop, subtype, rp = c(-5:-1,1:5)) +
  theme(legend.position="top", 
        legend.title=element_blank(), 
        legend.text = element_text(size=10),
        legend.key.size = unit(0.35, 'cm'), 
        legend.box.background = element_blank())

resid_plot(single_pos_dir, pop, subtype, rp = c(1)) +
  theme(legend.position="top", 
        legend.title=element_blank())
```

Here we are interested in the result at the +1 position, since this is the largest statistic observed overall subtypes for the single-position models (and as we'll later see, the largest statistic we observe in any of our analyses). While we can easily discuss the magnitude of this statistic relative to other relative entropy statistics, it isn't necessarily as simple to describe the practical interpretation of this statistic. For that, let's first examine the distribution of counts for the singletons and the matched controls at this position:

```{r}
df_resid <- get_pos_resid(single_pos_dir, pop, subtype, rp = 1)

df_resid %>%
  select(nuc, status, n) %>%
  pivot_wider(id_cols = nuc, names_from = status, values_from = n) %>%
  mutate(p_sing = singletons / sum(singletons),
         p_con = controls / sum(controls)) %>%
  knitr::kable()
```

How does this compare to what we observe for the C>T subtype at +1 when we do not control for CpG status?

```{r}
df_cpg_resid <- get_pos_resid(single_pos_dir, pop, "all_GC_AT", rp = 1)

df_cpg_resid %>%
  select(nuc, status, n) %>%
  pivot_wider(id_cols = nuc, names_from = status, values_from = n) %>%
  mutate(p_sing = singletons / sum(singletons),
         p_con = controls / sum(controls)) %>%
  mutate(abs_diff = abs(p_sing - p_con)) %>%
  knitr::kable()
  
resid_plot(single_pos_dir, pop, "all_GC_AT", rp = c(-5:-1,1:5))
```

Finally, I would like to look at the "significant" results between +/- 10 or so to +/- 250 (so the significant results not most central to the focal position). For these less extreme relative entropy values, what kinds of shifts in distribution do these correspond to?

```{r}
max_max_re <- max(df_results$max_var)
max_tot_var <- 0.5 * max(df_results$tot_var)

df_results %>%
  ggplot(aes(x = re, y = simple_re)) +
  geom_point()

df_results %>%
  filter(abs(offset) > 26, abs(offset) < 40) %>%
  ggplot(aes(x = re, y = simple_re)) +
  geom_point()

df_results %>%
  filter(re > 1e-6, re < 6e-6) %>%
  ggplot(aes(x = re, y = simple_re)) +
  geom_point()

df_results %>%
  filter(re > 1e-6, re < 6e-6) %>%
  ggplot(aes(x = re, y = max_var * 14402284)) +
  geom_point()

df_results %>%
  filter(re > 1e-6, re < 6e-6) %>%
  ggplot(aes(x = re, y = max_var / max_max_re)) +
  geom_point()

df_results %>%
  filter(re > 1e-6, re < 6e-6) %>%
  ggplot(aes(x = offset, y = max_var * 14402284)) +
  geom_point()

df_results %>%
  ggplot(aes(x = offset, y = max_var)) +
  geom_point()

df_results %>%
  filter(re > 1e-6, re < 6e-6) %>%
  ggplot(aes(x = offset, y = max_var / max_max_re)) +
  geom_point()

df_results %>%
  filter(re > 1e-6, re < 6e-6) %>%
  ggplot(aes(x = offset, y = (max_max_re - max_var) / max_max_re)) +
  geom_point()

df_results %>%
  filter(re > 1e-6, re < 6e-6) %>%
  ggplot(aes(x = offset, y = (max_tot_var - 0.5*tot_var) / max_tot_var)) +
  geom_point()

df_results %>%
  filter(re > 1e-6, re < 6e-6) %>%
  ggplot(aes(x = offset, y = 0.5 * tot_var)) +
  geom_point()

df_results %>%
  filter(re > 1e-6, re < 6e-6) %>%
  ggplot(aes(x = offset, y = (0.5 * tot_var) / max_tot_var)) +
  geom_point()
```

Ok, let's be more rigorous here. We first subset to the positions with a significant chi square statistic outside of the +/- 10 bp window.

```{r}
df_results_sub <- df_results %>%
  filter(abs(offset) > 10,
         sig == "Significant")
```

Using only this criterion as our cutoff, we remain with `r length(df_results_sub$sig)` out of 980 possible flanking sites. If instead of using alpha=0.05 cutoff for deviance statistics we instead found the 95th-percentile of our control-control relative entropy statistics, we'd instead use a cutoff of $\textrm{re} > 1.41 \times 10^{-7}$

```{r}
quantile_cutoff <- quantile(df_cc$re, 0.95)
df_results_sub <- df_results %>%
  filter(abs(offset) > 10,
         re > quantile_cutoff)
```

Now we have limited ourselves to `r length(df_results_sub$re)` of the 980 possible flanking positions. Where are these positions located?

```{r}
df_results_sub %>%
  ggplot(aes(x = offset, y = re)) +
  geom_point() +
  scale_y_log10() +
  theme_bw()

print(max(abs(df_results_sub$offset)))
```

The furthest flanking position considered is 426 bases from the focal position, while 75% of included flanking positions are within 231 bases of the focal position. So then, at these moderately influential positions, what does the shift in distribution look like?

```{r}
df_results_sub$max_var %>% summary()
df_results_sub$mean_var %>% summary()

n_singletons <- max(df_results$singletons)
(df_results_sub$max_var * n_singletons) %>% summary()
(df_results_sub$mean_var * n_singletons) %>% summary()
```


#### AFR

```{r}
pop <- "AFR"

all_results <- function(single_pos_dir, pop, subtype, control_control){
  df_results <- get_pos_results(single_pos_dir, pop = pop, subtype = subtype)
  df_cc <- get_pos_cc(control_control, pop, subtype)
  
  df_results$sig  <- ifelse(df_results$dev > qchisq(0.95,3), "Significant", "Non-significant")
  df_cc$sig <- "Control"
  df <- bind_rows(df_results, df_cc)
  
  p1 <- pos_level_line(df_results) 
  p2 <- pos_level_dot(df_results, friendly_pals$wong_eight) 
  p3 <- pos_level_line(df_results) + scale_y_log10() + ylab("Relative Entropy (log10)")

  p4 <- resid_plot(single_pos_dir, pop, subtype, rp = c(-5:-1,1:5))
  p5 <- pos_level_dot_cc(df) 
  
  print(p1)
  print(p2)
  print(p3)
  print(p4)
  print(p5)
}

all_results(single_pos_dir, pop, subtype, control_control)
```

#### AMR

```{r}
pop <- "AMR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### EAS

```{r}
pop <- "EAS"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### EUR

```{r}
pop <- "EUR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### SAS

```{r}
pop <- "EUR"
all_results(single_pos_dir, pop, subtype, control_control)
```

### A > C

```{r}
subtype <- "AT_CG"
```

#### ALL

```{r}
pop <- "ALL"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### AFR

```{r}
pop <- "AFR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### AMR

```{r}
pop <- "AMR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### EAS

```{r}
pop <- "EAS"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### EUR

```{r}
pop <- "EUR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### SAS

```{r}
pop <- "EUR"
all_results(single_pos_dir, pop, subtype, control_control)
```

### A > T

```{r}
subtype <- "AT_TA"
```

#### ALL

```{r}
pop <- "ALL"
all_results(single_pos_dir, pop, subtype, control_control)
```

```{r}
df_resid <- get_pos_resid(single_pos_dir, pop, subtype, rp = 1)

df_resid %>%
  select(nuc, status, n) %>%
  pivot_wider(id_cols = nuc, names_from = status, values_from = n) %>%
  mutate(p_sing = singletons / sum(singletons),
         p_con = controls / sum(controls)) %>%
  knitr::kable()
```

#### AFR

```{r}
pop <- "AFR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### AMR

```{r}
pop <- "AMR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### EAS

```{r}
pop <- "EAS"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### EUR

```{r}
pop <- "EUR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### SAS

```{r}
pop <- "EUR"
all_results(single_pos_dir, pop, subtype, control_control)
```

### C > T

```{r}
subtype <- "GC_AT"
```

#### ALL

```{r}
pop <- "ALL"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### AFR

```{r}
pop <- "AFR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### AMR

```{r}
pop <- "AMR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### EAS

```{r}
pop <- "EAS"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### EUR

```{r}
pop <- "EUR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### SAS

```{r}
pop <- "EUR"
all_results(single_pos_dir, pop, subtype, control_control)
```

### C > A

```{r}
subtype <- "GC_TA"
```

#### ALL

```{r}
pop <- "ALL"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### AFR

```{r}
pop <- "AFR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### AMR

```{r}
pop <- "AMR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### EAS

```{r}
pop <- "EAS"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### EUR

```{r}
pop <- "EUR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### SAS

```{r}
pop <- "EUR"
all_results(single_pos_dir, pop, subtype, control_control)
```

### C > G

```{r}
subtype <- "GC_CG"
```

#### ALL

```{r}
pop <- "ALL"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### AFR

```{r}
pop <- "AFR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### AMR

```{r}
pop <- "AMR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### EAS

```{r}
pop <- "EAS"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### EUR

```{r}
pop <- "EUR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### SAS

```{r}
pop <- "EUR"
all_results(single_pos_dir, pop, subtype, control_control)
```

### CpG > TpG

```{r}
subtype <- "cpg_GC_AT"
```

#### ALL

```{r}
pop <- "ALL"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### AFR

```{r}
pop <- "AFR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### AMR

```{r}
pop <- "AMR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### EAS

```{r}
pop <- "EAS"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### EUR

```{r}
pop <- "EUR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### SAS

```{r}
pop <- "EUR"
all_results(single_pos_dir, pop, subtype, control_control)
```

### CpG > ApG

```{r}
subtype <- "cpg_GC_TA"
```

#### ALL

```{r}
pop <- "ALL"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### AFR

```{r}
pop <- "AFR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### AMR

```{r}
pop <- "AMR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### EAS

```{r}
pop <- "EAS"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### EUR

```{r}
pop <- "EUR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### SAS

```{r}
pop <- "EUR"
all_results(single_pos_dir, pop, subtype, control_control)
```

### CpG > GpG

```{r}
subtype <- "cpg_GC_CG"
```

#### ALL

```{r}
pop <- "ALL"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### AFR

```{r}
pop <- "AFR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### AMR

```{r}
pop <- "AMR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### EAS

```{r}
pop <- "EAS"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### EUR

```{r}
pop <- "EUR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### SAS

```{r}
pop <- "EUR"
all_results(single_pos_dir, pop, subtype, control_control)
```

## All-in-one Figures

Here we explore options for displaying results across multiple mutation subtypes in one (or more, but less than 9) figure(s).


* Bars instead of lines and dots
* Add colours to ease interpretation
    * The nucleotide "boes" on top and right of figure
* grid lines are "noise"

```{r fig.dim = c(10, 6)}
df <- get_pos_results_all(single_pos_dir, "ALL", subtypes)

df <- df %>% mutate(from_n = ifelse(str_starts(subtype, "A"), "A", ifelse(str_starts(subtype, "cpg"), "CpG", "C")),
         to_n = ifelse(str_starts(subtype, "A"), str_sub(subtype, 4, 4), str_sub(subtype, -1)))

p1 <- df %>%
  filter(from_n == "A", abs(offset) < 11) %>%
  ggplot(aes(x = offset, y = re)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  facet_grid(vars(to_n), vars(from_n)) +
  xlab("Relative Position") +
  ylab("Relative Entropy")

y_lim_all <- ggplot_build(p1)$layout$panel_scales_y[[1]]$range$range
y_lim_all[1] <- 0

p1 <- p1 + ylim(y_lim_all)

p2 <- df %>%
  filter(from_n == "C", abs(offset) < 11) %>%
  ggplot(aes(x = offset, y = re)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  facet_grid(vars(to_n), vars(from_n)) +
  xlab("Relative Position") +
  ylab("") +
  ylim(y_lim_all)

p3 <- df %>%
  filter(from_n == "CpG", abs(offset) < 11) %>%
  ggplot(aes(x = offset, y = re)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  facet_grid(vars(to_n), vars(from_n)) +
  xlab("Relative Position") +
  ylab("") +
  ylim(y_lim_all)

gridExtra::grid.arrange(p1, p2, p3, nrow = 1)
```

