---
title: "single_position_results"
author: "Andy Beck"
date: "2025-01-06"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Notes

* Save figures using something along the lines of `ggsave("file.png", dpi=300)`

## Setup

### Libraries

```{r}
library(tidyverse)
library(ggpubfigs) # functions for color blind friendly palettes
library(reactable) # table output for html
library(knitr) # PDF table output
```

### Paths

```{r}
base_dir = "output"
single_pos_dir = paste0(base_dir, "/single_pos")
control_control <- paste0(base_dir, "/single_pos_cc")

figure_output = paste0(base_dir, "/figs/single_pos")
```


### Base Functions

Load position-level results:

```{r}
subtypes <- c("AT_CG", "AT_GC", "AT_TA",
            "GC_AT", "GC_TA", "GC_CG",
            "cpg_GC_AT", "cpg_GC_TA", "cpg_GC_CG")

get_pos_results <- function(res_dir, pop, subtype){
  f_name <- paste0(res_dir, "/", pop, "/", subtype, ".csv")
  df <- read_csv(f_name, show_col_types = FALSE) %>%
    rowwise() %>%
    mutate(re = dev / (2 * (singletons+controls)))
  return(df)
}

get_pos_cc <- function(control_control, pop, subtype){
  f_name <- paste0(control_control, "/", pop, "/", subtype, "_v2.csv")
  df <- read_csv(f_name, show_col_types = FALSE) %>%
           mutate(re = dev / (2 * (singletons + controls) ))
  return(df)
}

get_pos_results_all <- function(res_dir, pop, subtypes){
  results <- vector(mode = "list", length = length(subtypes))
  
  for(i in 1:length(subtypes)){
    results[[i]] <- get_pos_results(res_dir, pop, subtypes[[i]])
    results[[i]]$subtype <- subtypes[[i]]
  }
  
  return(bind_rows(results))
}

subtype_print_names <- function(st){
  if(str_starts(st, "AT")){
    return(paste0("A → ", str_sub(st, 4, 4)))
  } else if(str_starts(st, "GC")){
    return(paste0("C → ", str_sub(st, 5, 5)))
  } else{
    return(paste0("CpG → ", str_sub(st, 9, 9), "pG"))
  }
}
```

Load position-level residuals:

```{r}
get_pos_resid <- function(res_dir, pop, subtype, rp){
  f_name <- paste0(res_dir, "/resid/", pop, "/", subtype, "_rp_", rp, ".csv")
  df <- read_csv(f_name, show_col_types = FALSE)
  df$rp <- rp
  return(df)
}

get_pos_resid_all <- function(res_dir, pop, subtype, rp = c(-500:-1, 1:500)){
  results <- vector(mode = "list", length = length(rp))
  for(i in 1:length(rp)){
    if(rp[i] == 1 & str_starts(subtype, "cpg")) next
    results[[i]] <- get_pos_resid(res_dir, pop, subtype, rp[i])
  }
  df <- bind_rows(results)
  return(df)
}
```

### Figure Functions

```{r}
pos_level_line <- function(res_df, rp_lim = 20){
  p <- res_df %>%
    filter(abs(offset) <= rp_lim) %>%
    ggplot(aes(x = offset, y = re)) +
    geom_point() +
    geom_line() +
    theme_classic() +
    theme(text = element_text(family = "Arial"),
          legend.position = "inside",
          legend.position.inside = c(.85, .8)) +
    xlab(paste0("Relative Position: ", subtype_print_names(subtype))) +
    ylab("Relative Entropy")
  return(p)
}

pos_level_dot <- function(res_df, palette_name = friendly_pal("contrast_three")){
  p <- res_df %>%
    mutate(sig = dev > qchisq(0.95, 3)) %>%
    ggplot(aes(x = offset, y = re, color = sig)) +
    scale_y_log10() +
    geom_point() +
    scale_color_manual(values = palette_name) +
    theme_classic() +
    theme(text = element_text(family = "Arial"),
          legend.position = "inside",
          legend.position.inside = c(.85, .8)) +
    xlab(paste0("Relative Position: ", subtype_print_names(subtype))) +
    ylab("Log10 Relative Entropy") +
    labs(color = "Significant")
  return(p)
}

pos_level_dot_cc <- function(res_df, color_pal = friendly_pals$wong_eight[c(3,1:2)]){
  p <- res_df %>%
    ggplot(aes(x = offset, y = re, color = sig, alpha = sig)) +
    scale_y_log10() +
    geom_point() +
    scale_alpha_manual(values = c(0.2, 1, 1)) + guides(alpha="none") +
    scale_color_manual(values = color_pal) +
    theme_classic() +
    theme(text = element_text(family = "Arial"),
          legend.position = "inside",
          legend.position.inside = c(.85, .8)) +
    xlab(paste0("Relative Position: ", subtype_print_names(subtype))) +
    ylab("Log10 Relative Entropy") +
    labs(color = "Significant")
  return(p)
}

resid_plot <- function(res_dir, pop, subtype, rp = c(-10:-1, 1:10), palette_name = friendly_pal("nickel_five")){
  res_df <- get_pos_resid_all(res_dir, pop, subtype, rp)
  p <- res_df %>%
    select(rp, nuc, status, res) %>%
    pivot_wider(names_from = status, values_from = res) %>%
    rowwise() %>%
    mutate(dev_cont = sign(singletons) * (singletons ^ 2 + controls ^ 2)) %>%
    ggplot(aes(x = rp, y = dev_cont, fill = nuc)) +
    geom_bar(position="stack", stat="identity") +
    scale_fill_manual(values = palette_name) +
    theme_classic() +
    xlab(paste0("Relative Position: ", subtype_print_names(subtype))) +
    ylab("Singleton Enrichment") +
    labs(fill = "Nucleotide") +
    theme(legend.position = "inside", legend.position.inside = c(0.9, 0.8),
          legend.box.background = element_rect(color="black", linewidth =1)) +
    geom_hline(yintercept = 0, colour = "black", linetype = 3)
  return(p)
}
```


## Position Level Result Plots

### A > G

```{r}
subtype <- "AT_GC"
```


#### ALL

```{r}
pop <- "ALL"
df_results <- get_pos_results(single_pos_dir, pop = pop, subtype = subtype)
df_cc <- get_pos_cc(control_control, pop, subtype)

df_results$sig  <- ifelse(df_results$dev > qchisq(0.95,3), "Significant", "Non-significant")
df_cc$sig <- "Control"
df <- bind_rows(df_results, df_cc)

pos_level_line(df_results) 
pos_level_dot(df_results, friendly_pals$wong_eight)
pos_level_dot_cc(df) 
pos_level_line(df_results) + scale_y_log10() + ylab("Relative Entropy (log10)")
```

For the A > G mutation subtype, we observe the strongest signal of influence (relative entropy) at the +1 relative position. We observed that, in general, the signal decreases as positions further from the focal (0) position are considered, but note that this pattern is not monotonic in either direction, with the most obvious example of this being the stronger signal observed at -2 relative to the signal observed at -1. When compared to the critical value for a $\chi^2$ distribution with 3 degrees of freedom at the $\alpha=0.05$ level of significance, we find that the underlying deviance statistics remain significant up to approximately +/- 250bp; from here, we see values hover in a range that includes values which are not significant. The above results are based on the distributions of nucleotides flanking singletons and their matched controls across all five 1kGP super-populations.

Next, we plot the nucleotide contribution to each position's deviance statistic:

```{r}
resid_plot(single_pos_dir, pop, subtype, rp = c(-5:-1,1:5))
```

Here we are interested in the result at the +1 position, since this is the largest statistic observed overall subtypes for the single-position models (and as we'll later see, the largest statistic we observe in any of our analyses). While we can easily discuss the magnitude of this statistic relative to other relative entropy statistics, it isn't necessarily as simple to describe the practical interpretation of this statistic. For that, let's first examine the distribution of counts for the singletons and the matched controls at this position:

```{r}
df_resid <- get_pos_resid(single_pos_dir, pop, subtype, rp = 1)

df_resid %>%
  select(nuc, status, n) %>%
  pivot_wider(id_cols = nuc, names_from = status, values_from = n) %>%
  mutate(p_sing = singletons / sum(singletons),
         p_con = controls / sum(controls)) %>%
  knitr::kable()
```

How does this compare to what we observe for the C>T subtype at +1 when we do not control for CpG status?

```{r}
df_cpg_resid <- get_pos_resid(single_pos_dir, pop, "all_GC_AT", rp = 1)

df_cpg_resid %>%
  select(nuc, status, n) %>%
  pivot_wider(id_cols = nuc, names_from = status, values_from = n) %>%
  mutate(p_sing = singletons / sum(singletons),
         p_con = controls / sum(controls)) %>%
  mutate(abs_diff = abs(p_sing - p_con)) %>%
  knitr::kable()
  
resid_plot(single_pos_dir, pop, "all_GC_AT", rp = c(-5:-1,1:5))
```

Finally, I would like to look at the "significant" results between +/- 10 or so to +/- 250 (so the significant results not most central to the focal position). For these less extreme relative entropy values, what kinds of shifts in distribution do these correspond to?

```{r}
max_max_re <- max(df_results$max_var)
max_tot_var <- 0.5 * max(df_results$tot_var)

df_results %>%
  ggplot(aes(x = re, y = simple_re)) +
  geom_point()

df_results %>%
  filter(abs(offset) > 26, abs(offset) < 40) %>%
  ggplot(aes(x = re, y = simple_re)) +
  geom_point()

df_results %>%
  filter(re > 1e-6, re < 6e-6) %>%
  ggplot(aes(x = re, y = simple_re)) +
  geom_point()

df_results %>%
  filter(re > 1e-6, re < 6e-6) %>%
  ggplot(aes(x = re, y = max_var * 14402284)) +
  geom_point()

df_results %>%
  filter(re > 1e-6, re < 6e-6) %>%
  ggplot(aes(x = re, y = max_var / max_max_re)) +
  geom_point()

df_results %>%
  filter(re > 1e-6, re < 6e-6) %>%
  ggplot(aes(x = offset, y = max_var * 14402284)) +
  geom_point()

df_results %>%
  ggplot(aes(x = offset, y = max_var)) +
  geom_point()

df_results %>%
  filter(re > 1e-6, re < 6e-6) %>%
  ggplot(aes(x = offset, y = max_var / max_max_re)) +
  geom_point()

df_results %>%
  filter(re > 1e-6, re < 6e-6) %>%
  ggplot(aes(x = offset, y = (max_max_re - max_var) / max_max_re)) +
  geom_point()

df_results %>%
  filter(re > 1e-6, re < 6e-6) %>%
  ggplot(aes(x = offset, y = (max_tot_var - 0.5*tot_var) / max_tot_var)) +
  geom_point()

df_results %>%
  filter(re > 1e-6, re < 6e-6) %>%
  ggplot(aes(x = offset, y = 0.5 * tot_var)) +
  geom_point()

df_results %>%
  filter(re > 1e-6, re < 6e-6) %>%
  ggplot(aes(x = offset, y = (0.5 * tot_var) / max_tot_var)) +
  geom_point()
```


#### AFR

```{r}
pop <- "AFR"

all_results <- function(single_pos_dir, pop, subtype, control_control){
  df_results <- get_pos_results(single_pos_dir, pop = pop, subtype = subtype)
  df_cc <- get_pos_cc(control_control, pop, subtype)
  
  df_results$sig  <- ifelse(df_results$dev > qchisq(0.95,3), "Significant", "Non-significant")
  df_cc$sig <- "Control"
  df <- bind_rows(df_results, df_cc)
  
  p1 <- pos_level_line(df_results) 
  p2 <- pos_level_dot(df_results, friendly_pals$wong_eight) 
  p3 <- pos_level_line(df_results) + scale_y_log10() + ylab("Relative Entropy (log10)")

  p4 <- resid_plot(single_pos_dir, pop, subtype, rp = c(-5:-1,1:5))
  p5 <- pos_level_dot_cc(df) 
  
  print(p1)
  print(p2)
  print(p3)
  print(p4)
  print(p5)
}

all_results(single_pos_dir, pop, subtype, control_control)
```

#### AMR

```{r}
pop <- "AMR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### EAS

```{r}
pop <- "EAS"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### EUR

```{r}
pop <- "EUR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### SAS

```{r}
pop <- "EUR"
all_results(single_pos_dir, pop, subtype, control_control)
```

### A > C

```{r}
subtype <- "AT_CG"
```

#### ALL

```{r}
pop <- "ALL"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### AFR

```{r}
pop <- "AFR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### AMR

```{r}
pop <- "AMR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### EAS

```{r}
pop <- "EAS"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### EUR

```{r}
pop <- "EUR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### SAS

```{r}
pop <- "EUR"
all_results(single_pos_dir, pop, subtype, control_control)
```

### A > T

```{r}
subtype <- "AT_TA"
```

#### ALL

```{r}
pop <- "ALL"
all_results(single_pos_dir, pop, subtype, control_control)
```

```{r}
df_resid <- get_pos_resid(single_pos_dir, pop, subtype, rp = 1)

df_resid %>%
  select(nuc, status, n) %>%
  pivot_wider(id_cols = nuc, names_from = status, values_from = n) %>%
  mutate(p_sing = singletons / sum(singletons),
         p_con = controls / sum(controls)) %>%
  knitr::kable()
```

#### AFR

```{r}
pop <- "AFR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### AMR

```{r}
pop <- "AMR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### EAS

```{r}
pop <- "EAS"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### EUR

```{r}
pop <- "EUR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### SAS

```{r}
pop <- "EUR"
all_results(single_pos_dir, pop, subtype, control_control)
```

### C > T

```{r}
subtype <- "GC_AT"
```

#### ALL

```{r}
pop <- "ALL"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### AFR

```{r}
pop <- "AFR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### AMR

```{r}
pop <- "AMR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### EAS

```{r}
pop <- "EAS"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### EUR

```{r}
pop <- "EUR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### SAS

```{r}
pop <- "EUR"
all_results(single_pos_dir, pop, subtype, control_control)
```

### C > A

```{r}
subtype <- "GC_TA"
```

#### ALL

```{r}
pop <- "ALL"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### AFR

```{r}
pop <- "AFR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### AMR

```{r}
pop <- "AMR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### EAS

```{r}
pop <- "EAS"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### EUR

```{r}
pop <- "EUR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### SAS

```{r}
pop <- "EUR"
all_results(single_pos_dir, pop, subtype, control_control)
```

### C > G

```{r}
subtype <- "GC_CG"
```

#### ALL

```{r}
pop <- "ALL"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### AFR

```{r}
pop <- "AFR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### AMR

```{r}
pop <- "AMR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### EAS

```{r}
pop <- "EAS"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### EUR

```{r}
pop <- "EUR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### SAS

```{r}
pop <- "EUR"
all_results(single_pos_dir, pop, subtype, control_control)
```

### CpG > TpG

```{r}
subtype <- "cpg_GC_AT"
```

#### ALL

```{r}
pop <- "ALL"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### AFR

```{r}
pop <- "AFR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### AMR

```{r}
pop <- "AMR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### EAS

```{r}
pop <- "EAS"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### EUR

```{r}
pop <- "EUR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### SAS

```{r}
pop <- "EUR"
all_results(single_pos_dir, pop, subtype, control_control)
```

### CpG > ApG

```{r}
subtype <- "cpg_GC_TA"
```

#### ALL

```{r}
pop <- "ALL"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### AFR

```{r}
pop <- "AFR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### AMR

```{r}
pop <- "AMR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### EAS

```{r}
pop <- "EAS"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### EUR

```{r}
pop <- "EUR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### SAS

```{r}
pop <- "EUR"
all_results(single_pos_dir, pop, subtype, control_control)
```

### CpG > GpG

```{r}
subtype <- "cpg_GC_CG"
```

#### ALL

```{r}
pop <- "ALL"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### AFR

```{r}
pop <- "AFR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### AMR

```{r}
pop <- "AMR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### EAS

```{r}
pop <- "EAS"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### EUR

```{r}
pop <- "EUR"
all_results(single_pos_dir, pop, subtype, control_control)
```

#### SAS

```{r}
pop <- "EUR"
all_results(single_pos_dir, pop, subtype, control_control)
```

## All-in-one Figures

Here we explore options for displaying results across multiple mutation subtypes in one (or more, but less than 9) figure(s).

```{r fig.dim = c(10, 6)}
df <- get_pos_results_all(single_pos_dir, "ALL", subtypes)

df <- df %>% mutate(from_n = ifelse(str_starts(subtype, "A"), "A", ifelse(str_starts(subtype, "cpg"), "CpG", "C")),
         to_n = ifelse(str_starts(subtype, "A"), str_sub(subtype, 4, 4), str_sub(subtype, -1)))

p1 <- df %>%
  filter(from_n == "A", abs(offset) < 11) %>%
  ggplot(aes(x = offset, y = re)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  facet_grid(vars(to_n), vars(from_n)) +
  xlab("Relative Position") +
  ylab("Relative Entropy")

y_lim_all <- ggplot_build(p1)$layout$panel_scales_y[[1]]$range$range

p2 <- df %>%
  filter(from_n == "C", abs(offset) < 11) %>%
  ggplot(aes(x = offset, y = re)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  facet_grid(vars(to_n), vars(from_n)) +
  xlab("Relative Position") +
  ylab("Relative Entropy") +
  ylim(y_lim_all)

p3 <- df %>%
  filter(from_n == "CpG", abs(offset) < 11) %>%
  ggplot(aes(x = offset, y = re)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  facet_grid(vars(to_n), vars(from_n)) +
  xlab("Relative Position") +
  ylab("Relative Entropy") +
  ylim(y_lim_all)

gridExtra::grid.arrange(p1, p2, p3, nrow = 1)
```

