---
title: "sliding_window"
author: "theandyb"
date: "2026-09-25"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

```{r}
library(tidyverse)
library(ggpubfigs) # functions for color blind friendly palettes
library(reactable) # table output for html
library(knitr) # PDF table output
library(conflicted)
library(kableExtra)

conflict_prefer("filter", "dplyr")
conflict_prefer("lag", "dplyr")
conflict_prefer("theme_grey", "ggpubfigs")

cb_palette <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", 
                "#0072B2", "#D55E00", "#CC79A7", "#000000",
                "#490092", "#999999", "#59260B")
```

### Paths

```{r}
base_dir = "output"
single_pos_dir = paste0(base_dir, "/single_pos")
control_control <- paste0(base_dir, "/single_pos_cc")
```

```{r}
theme_publication <- function(base_size = 12, base_family = "sans") {
  theme_bw(base_size = base_size, base_family = base_family) %+replace% 
    theme(
      # Remove background elements
      panel.border     = element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      
      # Style the axes
      axis.line        = element_line(color = "black", linewidth = 0.5),
      axis.ticks       = element_line(color = "black", linewidth = 0.5),
      
      # Refine text and labels
      strip.background = element_rect(fill = "white", color = "white"),
      strip.text       = element_text(face = "bold", size = rel(1)),
      plot.title       = element_text(face = "bold", size = rel(1.2), hjust = 0),
      axis.title = element_text(size = rel(1.0)),
      axis.text = element_text(size = rel(0.8)),
      
      # Legend positioning
      legend.key       = element_blank(),
      legend.direction = "horizontal",
      legend.position  = "bottom"
    )
}

cb_palette <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", 
                "#0072B2", "#D55E00", "#CC79A7", "#000000")
```

### Base Functions

Load position-level results:

```{r}
subtypes <- c("AT_CG", "AT_GC", "AT_TA",
            "GC_TA", "GC_CG", "GC_AT",
            "cpg_GC_TA", "cpg_GC_CG", "cpg_GC_AT")

get_pos_results <- function(res_dir, pop, subtype){
  f_name <- paste0(res_dir, "/", pop, "/", subtype, ".csv")
  df <- read_csv(f_name, show_col_types = FALSE) %>%
    rowwise() %>%
    mutate(re = dev / (2 * (singletons+controls)))
  return(df)
}

get_pos_cc <- function(control_control, pop, subtype){
  f_name <- paste0(control_control, "/", pop, "/", subtype, "_v2.csv")
  df <- read_csv(f_name, show_col_types = FALSE) %>%
           mutate(re = dev / (2 * (singletons + controls) ))
  return(df)
}

get_pos_results_all <- function(res_dir, pop, subtypes){
  results <- vector(mode = "list", length = length(subtypes))
  
  for(i in 1:length(subtypes)){
    results[[i]] <- get_pos_results(res_dir, pop, subtypes[[i]])
    results[[i]]$subtype <- subtypes[[i]]
  }
  
  return(bind_rows(results))
}

subtype_print_names <- function(st){
  if(str_starts(st, "AT")){
    return(paste0("A → ", str_sub(st, 4, 4)))
  } else if(str_starts(st, "GC")){
    return(paste0("C → ", str_sub(st, 5, 5)))
  } else if(str_starts(st, "all")){
    return(paste0("C* → ", str_sub(st, 9, 9)))
  } else{
    return(paste0("CpG → ", str_sub(st, 9, 9), "pG"))
  }
}
```

Load position-level residuals:

```{r}
get_pos_resid <- function(res_dir, pop, subtype, rp){
  f_name <- paste0(res_dir, "/resid/", pop, "/", subtype, "_rp_", rp, ".csv")
  df <- read_csv(f_name, show_col_types = FALSE)
  df$rp <- rp
  
  n_tot <- sum(df$n)
  df$re_res <- (df$res^2) / (2*n_tot)
  return(df)
}

get_pos_resid_all <- function(res_dir, pop, subtype, rp = c(-500:-1, 1:500)){
  results <- vector(mode = "list", length = length(rp))
  for(i in 1:length(rp)){
    if(rp[i] == 1 & str_starts(subtype, "cpg")) next
    results[[i]] <- get_pos_resid(res_dir, pop, subtype, rp[i])
  }
  df <- bind_rows(results)
  return(df)
}
```

## Summary Across All Subtypes

First let's get a look at the single position model results for all subtypes using the singletons from all 5 1kGP super-populations. 

```{r}
load_all_pos_results <- function(subtypes, pop = "ALL"){
  final <- get_pos_results(single_pos_dir, pop, subtypes[1]) %>%
    mutate(subtype = subtypes[1])
  
  for(i in 2:length(subtypes)){
    final <- final %>%
      bind_rows({
        get_pos_results(single_pos_dir, pop, subtypes[i]) %>%
          mutate(subtype = subtypes[i])
      })
  }
  final <- final %>%
    mutate(from_n = ifelse(str_starts(subtype, "A"), "A", 
                           ifelse(str_starts(subtype, "cpg"), "CpG", 
                                  ifelse(str_starts(subtype, "all"), "C*", "C"))),
         #to_n = ifelse(str_starts(subtype, "A"), str_sub(subtype, 4, 4), str_sub(subtype, -1))
         to_n = ifelse(str_starts(subtype, "A"), str_sub(subtype, 4,4), 
                       ifelse(str_starts(subtype, "cpg"), 
                              paste0(str_sub(subtype, -1), "pG"), 
                              str_sub(subtype, -1) )))
  return(final)
}

df_all <- load_all_pos_results(c(subtypes, "all_GC_AT"))
df_all$cramer <- sqrt(df_all$dev / (df_all$singletons + df_all$controls))

df_all$side <- ifelse(df_all$offset > 0, 1, -1)
```


## A > C

```{r}
df <- df_all %>% filter(subtype == "AT_CG")
```

Get the significant value for chi square df = 3, alpha = 95%:

```{r}
crit_val <- qchisq(0.95, 3)
```


Sliding window approach: for each window, starting at +/- 5 and increasing in size by 5, cacluate the proportion of significant statistics:

```{r}
get_window_prop <- function(df, window, crit_val = qchisq(0.95, 3)){
  sig_vec <- df %>%
    filter(abs(offset) <= window) %>%
    mutate(sig = dev > crit_val) %>%
    pull(sig)
  return(sum(sig_vec) / length(sig_vec))
}

get_window_df <- function(df, step_size = 5){
  max_window <- max(abs(df$offset))
  window_vec <- seq(step_size, max_window, by = step_size)
  prop_vec <- sapply(window_vec, function(w) get_window_prop(df, w))
  return(tibble(window = window_vec, prop_sig = prop_vec))
}

window_a_c <- get_window_df(df)

# Find smallest window with prop < 0.95
boundary_point <- window_a_c %>%
  filter(prop_sig < 0.95) %>%
  slice(1)

# plot window_a_c
window_a_c %>%
  ggplot(aes(x = window, y = prop_sig)) +
  geom_line() +
  geom_point() +
  theme_publication() +
  xlab("Window Size") +
  ylab("Proportion Significant") +
  ggtitle("A > C: Proportion of Significant Positions by Window Size")
```

## Boundary Point DF

We now want to get the boundary point for all subtypes

```{r}
result_df <- data.frame(subtype = character(), boundary_point = numeric())
for(st in subtypes){
  print(st)
  df <- df_all %>% filter(subtype == st)
  df_window <- get_window_df(df)
  boundary_point <- df_window %>%
    filter(prop_sig < 0.95) %>%
    slice(1)
  result_df <- rbind(result_df, data.frame(subtype = st, boundary_point = boundary_point$window))
}
```

Print out table for knitr html document:

```{r}
result_df %>%
  rename(Subtype = subtype, `Boundary Point` = boundary_point) %>%
  reactable::reactable()
```


Format the result_df as a LaTeX table for publication:

```{r}
# result_df %>%
#   rename(Subtype = subtype, `Boundary Point` = boundary_point) %>%
#   kbl(booktabs = T, caption = "Boundary Points for Each Subtype", format = "latex")
```


