---
title: "statistic_calibration"
author: "Andy Beck"
date: "2024-12-06"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r}
library(tidyverse)
```


## Introduction

In this document, we will use simulation to evaluate the calibration of our relative entropy statistic for the influence of local sequence context on mutation rates.

### Parameters

The main parameter for each mutation subtype is the number of singletons for that subtype. Here we'll collect these counts for each subtype across the five 1kGP super populations.

```{r}
df_singleton <- data.frame(pop = rep(c("AFR", "AMR", "EAS", "EUR", "SAS"), each = 9),
                           subtype = rep(c("AT_CG", "AT_GC", "AT_TA", "GC_AT", "GC_TA", "GC_CG", "cpg_GC_AT", "cpg_GC_TA", "cpg_GC_CG"), times = 5),
                           n_s = c(1262888, 4178952, 1138869, 3821143, 1330996, 1452209, 1921101, 124142, 162111,
                                   537361, 1834264, 495037, 1681832, 563594, 623455,  827258,  55814, 67466,
                                   924484, 3190894, 888364, 2849828, 981610, 1084701, 1451677, 94483, 121589,
                                   710178, 2336815, 650119, 2124485, 731704, 821459,  1064404, 70878, 88486,
                                   897754, 2861359, 791460, 2686394, 890932, 1000478, 1346096, 89577, 113843))

df_singleton <- df_singleton %>%
  bind_rows({ df_singleton %>%
    group_by(subtype) %>%
    summarize(n_s = sum(n_s)) %>%
    ungroup() %>%
    mutate(pop = "ALL") %>%
    select(pop, subtype, n_s)})
```

#### Nucleotide frequencies in chr22

For some of our simulations, we will want proportions of nucleotides. While we could fiddle around with this, for a starting point let's get the frequency of each nucleotide in chromosome 22:

```{r}
df_22 <- read_tsv("data/reference/gc100_22.bed",
                  col_names = c("chr", "start", "end", "pct_at", "pct_gc", "n_A", "n_C", "n_G", "n_T", "n_N", "n_O", "len")) %>%
  filter(n_N < 50)

n_nuc_22 <- df_22 %>%
  ungroup() %>%
  summarize(n_A = sum(n_A),
            n_C = sum(n_C),
            n_G = sum(n_G),
            n_T = sum(n_T)
            ) %>%
  as.numeric()

prop_nuc_22 <- n_nuc_22 / sum(n_nuc_22)
```



## Single Position Models

In our first analysis, we will sample observed tables of singletons and matched controls stratified by the nucleotide at a single flanking position. We fill the following table:

| Nucleotide | Singletons | Controls  |
|------------|------------|-----------|
| A          | $n_{A,s}$  | $n_{A,c}$ |
| C          | $n_{C,s}$  | $n_{C,c}$ |
| G          | $n_{G,s}$  | $n_{G,c}$ |
| T          | $n_{T,s}$  | $n_{T,c}$ |

using the following model:

$$
\log(n_{i,s}) = \lambda_0 + \sum_{j \in \{C,G,T\}} \lambda_j 1(i==j) + \lambda_{\textrm{singleton}} 1(s==\textrm{singleton})
$$

We'll look at both the deviance statistic (which has a theoretical $\chi^2$ distribution with 3 degrees of freedom) and the relative entropy statistic.

```{r}
simulate_1_pos_null <- function(n_s, nuc_prop, c_s_ratio = 5){
  singletons <- rmultinom(1, n_s, nuc_prop)
  controls <- rmultinom(1, n_s * c_s_ratio, nuc_prop)
  
  test_df <- data.frame(nuc = c("A", "C", "G", "T"),
                        singletons = singletons,
                        controls = controls) %>%
    pivot_longer(singletons:controls, names_to = "status", values_to = "n")
  
  mod_obj <- glm(n ~ nuc + status, test_df, family = poisson())
  
  dev_stat <- deviance(mod_obj)
  re_val <- dev_stat / (2 * n_s * (1 + c_s_ratio))
  return(data.frame(deviance = dev_stat, re = re_val))
}
```

```{r}
n_s <- df_singleton %>%
  filter(pop == "ALL", subtype == "AT_GC") %>%
  pull(n_s)

n_sim <- 2000

set.seed(1848)
df_sim <- vector(mode = "list", length = n_sim)
for(i in 1:n_sim){
  df_sim[[i]] <- simulate_1_pos_null(n_s, prop_nuc_22)
}

df_sim <- df_sim %>% bind_rows()
```

Out of our 2000 simulations, how many false positive statistics do we observe?

```{r}
sum(df_sim$deviance > qchisq(0.95, 3))
```

This is incredibly close to 0.05. Now, among the statistics which were significant, what does their distribution look like?

```{r}
df_sim %>%
  filter(deviance > qchisq(0.95, 3)) %>%
  pull(deviance) %>%
  {plot(density(.))}
```

How does the overall distribution of statistics look relative to the theoretical $\chi^2_{\textrm{df}=3}$?

```{r}
ggplot() +
  aes(sample = df_sim$deviance) +
  stat_qq(distribution = qchisq, dparams = 3) +
  stat_qq_line(col = "blue", distribution = qchisq, dparams = 3) +
  xlab("Quantile") + ylab("Deviance")
```

And finally, what does our density look like?

```{r}
df_sim %>%
  ggplot(aes(x = deviance)) +
  geom_density() +
  stat_function(fun = dchisq, args = c(3), n = 1000, colour = "red") + 
  geom_vline(xintercept = 1, colour = "red", linetype = 2)
```

```{r}
MASS::fitdistr(df_sim$deviance, "chi-squared", start = list("df" = 4))
```


## Alternative simulation strategy

Here we'll simulate directly from the poisson distribution. We'll place normal priors on the nucleotide parameters, and fix the status parameter at $log(\frac{1}{5})$


```{r}
simulate_1_pos_null_poisson <- function(n_s, nuc_prop, c_s_ratio = 5){
  
  beta_s = log(1/c_s_ratio)
  beta_0 = log(n_s * c_s_ratio * nuc_prop[1])
  beta_c = log(nuc_prop[2] / nuc_prop[1])
  beta_g = log(nuc_prop[3] / nuc_prop[1])
  beta_t = log(nuc_prop[4] / nuc_prop[1])
  
  # theta_0 <- rnorm(1, beta_0, 1/10)
  # theta_c <- rnorm(1, beta_c, 1/(10 ^ (ceiling(-1*log10(-1 * beta_c)) + 2) ))
  # theta_g <- rnorm(1, beta_g, 1/(10 ^ (ceiling(-1*log10(-1 * beta_g)) + 2) ))
  # theta_t <- rnorm(1, beta_t, 1/(10 ^ (ceiling(-1*log10(-1 * beta_t)) + 2) ))
  
  controls <- c(rpois(1, exp(beta_0)), rpois(1, exp(beta_0 + beta_c)), rpois(1, exp(beta_0+ beta_g)),  rpois(1, exp(beta_0+ beta_t)) )
  #controls <- c(controls, n_s*c_s_ratio - sum(controls))
  controls[controls<0] <- 0
  
  singletons <- c(rpois(1, exp(beta_0 + beta_s)), rpois(1, exp(beta_0 + beta_c + beta_s)), rpois(1, exp(beta_0+ beta_g + beta_s)), rpois(1, exp(beta_0+ beta_t + beta_s)))
  #singletons <- c(singletons, n_s - sum(singletons))
  singletons[singletons<0] <- 0
  
  test_df <- data.frame(nuc = c("A", "C", "G", "T"),
                        singletons = singletons,
                        controls = controls) %>%
    pivot_longer(singletons:controls, names_to = "status", values_to = "n")
  
  mod_obj <- glm(n ~ nuc + status, test_df, family = poisson())
  
  dev_stat <- deviance(mod_obj)
  re_val <- dev_stat / (2 * n_s * (1 + c_s_ratio))
  return(data.frame(deviance = dev_stat, re = re_val))
}
```

```{r}
n_s <- df_singleton %>%
  filter(pop == "ALL", subtype == "AT_GC") %>%
  pull(n_s)

n_sim <- 2000

set.seed(1848)
df_sim2 <- vector(mode = "list", length = n_sim)
for(i in 1:n_sim){
  df_sim2[[i]] <- simulate_1_pos_null_poisson(n_s, prop_nuc_22)
}

df_sim2 <- df_sim2 %>% bind_rows()
```

```{r}
sum(df_sim2$deviance > qchisq(0.95, 3))
```

```{r}
df_sim2 %>%
  ggplot(aes(x = deviance)) +
  geom_density() +
  stat_function(fun = dchisq, args = c(3), n = 1000, colour = "red") + 
  geom_vline(xintercept = 1, colour = "red", linetype = 2)
```


```{r}
ggplot() +
  aes(sample = df_sim2$deviance) +
  stat_qq(distribution = qchisq, dparams = 3) +
  stat_qq_line(col = "blue", distribution = qchisq, dparams = 3) +
  xlab("Quantile") + ylab("Deviance")
```

```{r}
MASS::fitdistr(df_sim2$deviance, "chi-squared", start = list("df" = 3))
```

## Poisson Simulation with Fixed Margins

```{r}
simulate_1_pos_null_fixM <- function(n_s, nuc_prop, c_s_ratio = 5){
  
  beta_s = log(1/c_s_ratio)
  beta_0 = log(n_s * c_s_ratio * nuc_prop[1])
  beta_c = log(nuc_prop[2] / nuc_prop[1])
  beta_g = log(nuc_prop[3] / nuc_prop[1])
  beta_t = log(nuc_prop[4] / nuc_prop[1])
  
  controls <- c(rpois(1, exp(beta_0)), rpois(1, exp(beta_0 + beta_c)), rpois(1, exp(beta_0+ beta_g)))
  controls <- c(controls, n_s*c_s_ratio - sum(controls))
  #controls[controls<0] <- 0
  
  singletons <- c(rpois(1, exp(beta_0 + beta_s)), rpois(1, exp(beta_0 + beta_c + beta_s)), rpois(1, exp(beta_0+ beta_g + beta_s)))
  singletons <- c(singletons, n_s - sum(singletons))
  #singletons[singletons<0] <- 0
  
  test_df <- data.frame(nuc = c("A", "C", "G", "T"),
                        singletons = singletons,
                        controls = controls) %>%
    pivot_longer(singletons:controls, names_to = "status", values_to = "n")
  
  mod_obj <- glm(n ~ nuc + status, test_df, family = poisson())
  
  dev_stat <- deviance(mod_obj)
  re_val <- dev_stat / (2 * n_s * (1 + c_s_ratio))
  return(data.frame(deviance = dev_stat, re = re_val))
}

simulate_1_pos_null_fixM_altM <- function(n_s, nuc_prop, c_s_ratio = 5){
  
  beta_s = log(1/c_s_ratio)
  beta_0 = log(n_s * c_s_ratio * nuc_prop[1])
  beta_c = log(nuc_prop[2] / nuc_prop[1])
  beta_g = log(nuc_prop[3] / nuc_prop[1])
  beta_t = log(nuc_prop[4] / nuc_prop[1])
  
  controls <- c(rpois(1, exp(beta_0)), rpois(1, exp(beta_0 + beta_c)), rpois(1, exp(beta_0+ beta_g)))
  controls <- c(controls, n_s*c_s_ratio - sum(controls))
  #controls[controls<0] <- 0
  
  singletons <- c(rpois(1, exp(beta_0 + beta_s)), rpois(1, exp(beta_0 + beta_c + beta_s)), rpois(1, exp(beta_0+ beta_g + beta_s)))
  singletons <- c(singletons, n_s - sum(singletons))
  #singletons[singletons<0] <- 0
  
  test_df <- data.frame(nuc = c("A", "C", "G", "T"),
                        singletons = singletons,
                        controls = controls) %>%
    pivot_longer(singletons:controls, names_to = "status", values_to = "n") %>%
    left_join(data.frame(status = c("singletons", "controls"), col_sum = c(n_s, n_s*5)), by = "status")
  
  mod_obj <- glm(n ~ nuc + offset(log(col_sum)), test_df, family = poisson())
  
  dev_stat <- deviance(mod_obj)
  re_val <- dev_stat / (2 * n_s * (1 + c_s_ratio))
  return(data.frame(deviance = dev_stat, re = re_val))
}
```

```{r}
n_s <- df_singleton %>%
  filter(pop == "ALL", subtype == "AT_GC") %>%
  pull(n_s)

n_sim <- 2000

set.seed(1848)
df_sim3 <- vector(mode = "list", length = n_sim)
for(i in 1:n_sim){
  df_sim3[[i]] <- simulate_1_pos_null_fixM(n_s, prop_nuc_22)
}

df_sim3 <- df_sim3 %>% bind_rows()

set.seed(1848)
df_sim4 <- vector(mode = "list", length = n_sim)
for(i in 1:n_sim){
  df_sim4[[i]] <- simulate_1_pos_null_fixM_altM(n_s, prop_nuc_22)
}

df_sim4 <- df_sim4 %>% bind_rows()
```

```{r}
sum(df_sim3$deviance > qchisq(0.95, 3))

sum(df_sim4$deviance > qchisq(0.95, 4))
```

```{r}
df_sim3 %>%
  ggplot(aes(x = deviance)) +
  geom_density() +
  stat_function(fun = dchisq, args = c(3), n = 1000, colour = "red") +
  geom_vline(xintercept = 1, colour = "red", linetype = 2)
# 
df_sim4 %>%
  ggplot(aes(x = deviance)) +
  geom_density() +
  stat_function(fun = dchisq, args = list("df" = 4, "ncp" = 0), n = 1000, colour = "red") +
  geom_vline(xintercept = 2, colour = "red", linetype = 2)
```


```{r}
ggplot() +
  aes(sample = df_sim3$deviance) +
  stat_qq(distribution = qchisq, dparams = 3) +
  stat_qq_line(col = "blue", distribution = qchisq, dparams = list("df" = 4, "ncp" = 0)) +
  xlab("Quantile") + ylab("Deviance")

ggplot() +
  aes(sample = df_sim3$deviance) +
  stat_qq(distribution = qchisq, dparams = 4) +
  stat_qq_line(col = "blue", distribution = qchisq, dparams = 4) +
  xlab("Quantile") + ylab("Deviance")
```

```{r}
opt_1 <- MASS::fitdistr(df_sim3$deviance, "chi-squared", start = list("df" = 4, "ncp" = 3))
opt_1$estimate
```

```{r}
opt_2 <- MASS::fitdistr(df_sim3$deviance, "chi-squared", start = list("ncp" = 1), df=4)
opt_2$estimate
```

Below we plot the density of the deviances along with the non-central $\chi^2$ distribution with both df and ncp parameters optimized:

```{r}
df_sim3 %>%
  ggplot(aes(x = deviance)) +
  geom_density() +
  stat_function(fun = dchisq, args = list("df" = opt_1$estimate["df"], "ncp" = opt_1$estimate["ncp"]), n = 1000, colour = "red")
```

Now plot the estimated distribution where we've fixed the df to be 4:

```{r}
df_sim3 %>%
  ggplot(aes(x = deviance)) +
  geom_density() +
  stat_function(fun = dchisq, args = list("df" = 4, "ncp" = opt_2$estimate["ncp"]), n = 1000, colour = "red")
```

And finally the distribution where we use a quasi-method-of-moments estimator of the ncp assuming df = 3

```{r}
ncp_est <- mean(df_sim3$deviance) - 4
df_sim3 %>%
  ggplot(aes(x = deviance)) +
  geom_density() +
  stat_function(fun = dchisq, args = list("df" = 4, "ncp" = ncp_est), n = 1000, colour = "red")
```

Lastly, to satisfy my curiosity, if we fix df = 4 and find ncp based on variance formula for the nc-chisq:

```{r}
ncp_var_est <- (var(df_sim3$deviance) - 8) / 4
df_sim3 %>%
  ggplot(aes(x = deviance)) +
  geom_density() +
  stat_function(fun = dchisq, args = list("df" = 4, "ncp" = ncp_var_est), n = 1000, colour = "red")
```


```{r}
df_sim3 %>%
  ggplot(aes(x = deviance)) +
  geom_density() +
  stat_function(fun = dchisq, args = c(3, ncp_est), n = 1000, colour = "red")
```


```{r}
df_sim3 %>%
  ggplot(aes(x = deviance)) +
  geom_density() +
  stat_function(fun = dchisq, args = c(3, ncp_est), n = 1000, colour = "red")
```

## Fit poisson model with offset

Here I want to see what happens with our single position test if we use an offset to condition on fix totals for singleton/control categories:

```{r}
singletons <- rmultinom(1, n_s, prop_nuc_22) #n_s * prop_nuc_22
controls <- rmultinom(1, n_s*5, prop_nuc_22 + c(-0.1, 0.1, -0.02, 0.02))

test_df <- data.frame(nuc = c("A", "C", "G", "T"),
                      singletons = singletons,
                      controls = controls) %>%
  pivot_longer(singletons:controls, names_to = "status", values_to = "n")

test_df <- test_df %>%
  left_join(data.frame(status = c("singletons", "controls"), col_sum = c(n_s, n_s*5)), by = "status")

mod_obj <- glm(n ~ nuc + status, test_df, family = poisson)
mod_obj2 <- glm(n ~ nuc + offset(log(col_sum)), test_df, family = poisson)
mod_obj3 <- glm(n ~ nuc * status + offset(log(col_sum)), test_df, family=poisson)
```



### Impact of sample size on distribution

Let's look at the distribution of the statistics in a simulation using a smaller sample size:

```{r}
n_s <- df_singleton %>%
  pull(n_s) %>%
  min()

n_sim <- 2000

set.seed(1848)
df_sim <- vector(mode = "list", length = n_sim)
for(i in 1:n_sim){
  df_sim[[i]] <- simulate_1_pos_null_fixM(n_s, prop_nuc_22)
}
```

```{r}
sum(df_sim$deviance > qchisq(0.95, 3))
```

```{r}
ggplot() +
  aes(sample = df_sim$deviance) +
  stat_qq(distribution = qchisq, dparams = 4) +
  stat_qq_line(col = "blue", distribution = qchisq, dparams = 3) +
  xlab("Quantile") + ylab("Deviance")
```

And finally, what does our density look like?

```{r}
df_sim %>%
  ggplot(aes(x = deviance)) +
  geom_density() +
  stat_function(fun = dchisq, args = list("df" = 4), n = 1000, colour = "red") + 
  geom_vline(xintercept = 2, colour = "red", linetype = 2)
```

```{r}
df_sim %>%
  mutate(type = "small") %>%
  bind_rows({df_sim3 %>% mutate(type = "big")}) %>%
  ggplot(aes(x = deviance, color = type)) +
  geom_density()

df_sim %>%
  mutate(type = "small") %>%
  bind_rows({df_sim3 %>% mutate(type = "big")}) %>%
  ggplot() + 
  aes(sample = deviance, color = type) + 
  stat_qq(distribution = qchisq, dparams = 4) +
  stat_qq_line(col = "blue", distribution = qchisq, dparams = 3) +
  xlab("Quantile") + ylab("Deviance")
```

### Varyinig nucleotide distributions

```{r}
nuc_prop_high_gh <- c(0.15,0.34,0.36,0.15)

n_s <- df_singleton %>%
  pull(n_s) %>%
  min()

n_sim <- 2000

set.seed(1848)
df_sim_gc <- vector(mode = "list", length = n_sim)
for(i in 1:n_sim){
  df_sim_gc[[i]] <- simulate_1_pos_null_fixM(n_s, nuc_prop_high_gh)
}

df_sim_gc <- df_sim_gc %>% bind_rows()
```

```{r}
sum(df_sim_gc$deviance > qchisq(0.95, 4))
```

```{r}
ggplot() +
  aes(sample = df_sim_gc$deviance) +
  stat_qq(distribution = qchisq, dparams = 4) +
  stat_qq_line(col = "blue", distribution = qchisq, dparams = 4) +
  xlab("Quantile") + ylab("Deviance")
```

And finally, what does our density look like?

```{r}
df_sim_gc %>%
  ggplot(aes(x = deviance)) +
  geom_density() +
  stat_function(fun = dchisq, args = list("df" = 4), n = 1000, colour = "red") + 
  geom_vline(xintercept = 2, colour = "red", linetype = 2)
```

