---
title: "three_pos_results"
author: "Andy Beck"
date: "2025-02-04"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

In this document we will evaluate the results from three-way interaction models. Our primary interest is contrasting their strength to single-position and two-way interactions.

## Setup

### Libraries

```{r}
library(tidyverse)
library(ggpubfigs) # functions for color blind friendly palettes
library(reactable)
```

### Paths

```{r}
base_dir <- "output" 
three_dir <- paste0(base_dir, "/three_pos") # template: base_dir/three_pos/{population}/{subtype}.csv
two_dir <- paste0(base_dir, "/two_pos") # template: base_dir/two_pos/{population}/{subtype}.csv
single_dir <- paste0(base_dir, "/single_pos") # template: base_dir/single_pos/{population}/{subtype}.csv
```

### Data Reading

```{r}
subtypes <- c("AT_CG", "AT_GC", "AT_TA",
            "GC_AT", "GC_TA", "GC_CG",
            "cpg_GC_AT", "cpg_GC_TA", "cpg_GC_CG")

read_single <- function(res_dir, pop, subtype, tp=FALSE){
  
  f_name <- paste0(res_dir, "/", pop, "/", subtype, ".csv")
  if(tp){
    return(read_csv(f_name, show_col_types = F, col_names = c("dev","singletons","controls","rp1","rp2")))
  }
  return(read_csv(f_name, show_col_types = FALSE))
}

read_all <- function(single_dir, two_dir, three_dir, pop, subtype){
  s_df <- read_single(single_dir, pop, subtype)
  two_df <- read_single(two_dir, pop, subtype, tp=T)
  three_df <- read_single(three_dir, pop, subtype)
  
  # compute re 
  three_df <- three_df %>%
    mutate(re = dev / (2 * (singletons + controls)))
  two_df <- two_df %>%
    mutate(re = dev / (2 * (singletons + controls)))
  s_df <- s_df %>%
    mutate(re = dev / (2 * (singletons + controls)))
  
  # rank by re
  three_df <- three_df %>%
    mutate(class_rank = dense_rank(desc(re)))
  two_df <- two_df %>%
    mutate(class_rank = dense_rank(desc(re)))
  s_df <- s_df %>%
    mutate(class_rank = dense_rank(desc(re)))
  
  # unify structure: dev, offsets, re, class, class_rank
  three_df <- three_df %>%
    rowwise() %>%
    mutate(offset = paste0(rp1, "_", rp2, "_", rp3)) %>%
    ungroup() %>%
    mutate(class = "3-pos") %>%
    select(dev, offset, re, class ,class_rank) 
  
  two_df <- two_df %>%
    mutate(class = "2-pos") %>%
    rowwise() %>%
    mutate(offset = paste0(rp1, "_", rp2)) %>%
    ungroup() %>%
    select(dev, offset, re, class, class_rank)
  
  s_df <- s_df %>%
    mutate(class = "1-pos",
           offset = as.character(offset)) %>%
    select(dev, offset, re, class, class_rank)
  
  final_df <- bind_rows(s_df, two_df, three_df)
  return(final_df)
}
```

### Rank Plots

```{r}
subtype_print_names <- function(st){
  if(str_starts(st, "AT")){
    return(paste0("A → ", str_sub(st, 4, 4)))
  } else if(str_starts(st, "GC")){
    return(paste0("C → ", str_sub(st, 5, 5)))
  } else{
    return(paste0("CpG → ", str_sub(st, 9, 9), "pG"))
  }
}

# function will print both table and plot
plot_ranked_re <- function(single_dir, two_dir, three_dir, pop, subtype, rank_lim = 20){
  df <- read_all(single_dir, two_dir, three_dir, pop, subtype)
  
  max_re <- max(df$re)
  df$pct_max <- df$re / max_re
  
  p <- df %>%
    filter(class_rank <= rank_lim) %>%
    mutate(class2 = recode(class, "1-pos" = "Single", "2-pos" = "Two-Way", "3-pos" = "Three-Way")) %>%
    ggplot(aes(x = class_rank, y = pct_max, colour = class2, shape = class2)) +
    geom_point(size = 2.5) +
    scale_colour_manual(values=friendly_pal("contrast_three"), name = "Type", breaks=c("Single","Two-Way","Three-Way")) +
    scale_shape_discrete(name = "Type", breaks=c("Single","Two-Way","Three-Way")) +
    theme_classic(base_size = 12, base_family = "Helvetica") +
    theme(legend.position = "inside",
          legend.position.inside = c(.8, .8)) +
    ylab("% of Max RE") +
    xlab(paste0("Within-class Rank: ", subtype_print_names(subtype)))
  return(list(df=df, fig=p))
}
```

## Figures and Tables

Let's look at mtcars instead of local sequence context non-sense...

```{r eval=TRUE}
reactable(mtcars[1:4, 1:5])
```

And now that the reactable stuff has been attached to our document, let's look at local sequence context results :)

### ALL

```{r results='asis', echo=FALSE}
subtype <- "ALL"
for(st in subtypes){
  res <- plot_ranked_re(single_dir, two_dir, three_dir, "ALL", st)
  cat(paste0('#### ', st, "\n"))
  print(res$fig)
  print(htmltools::tagList(reactable(res$df)))
}
```

### AFR

```{r results='asis', echo=FALSE}
subtype <- "AFR"
for(st in subtypes){
  res <- plot_ranked_re(single_dir, two_dir, three_dir, "ALL", st)
  cat(paste0('#### ', st, "\n"))
  print(res$fig)
  print(htmltools::tagList(reactable(res$df)))
}
```

### AMR

```{r results='asis', echo=FALSE}
subtype <- "AMR"
for(st in subtypes){
  res <- plot_ranked_re(single_dir, two_dir, three_dir, "ALL", st)
  cat(paste0('#### ', st, "\n"))
  print(res$fig)
  print(htmltools::tagList(reactable(res$df)))
}
```

### EAS

```{r results='asis', echo=FALSE}
subtype <- "EAS"
for(st in subtypes){
  res <- plot_ranked_re(single_dir, two_dir, three_dir, "ALL", st)
  cat(paste0('#### ', st, "\n"))
  print(res$fig)
  print(htmltools::tagList(reactable(res$df)))
}
```

### EUR

```{r results='asis', echo=FALSE}
subtype <- "EUR"
for(st in subtypes){
  res <- plot_ranked_re(single_dir, two_dir, three_dir, "ALL", st)
  cat(paste0('#### ', st, "\n"))
  print(res$fig)
  print(htmltools::tagList(reactable(res$df)))
}
```

### SAS

```{r results='asis', echo=FALSE}
subtype <- "SAS"
for(st in subtypes){
  res <- plot_ranked_re(single_dir, two_dir, three_dir, "ALL", st)
  cat(paste0('#### ', st, "\n"))
  print(res$fig)
  print(htmltools::tagList(reactable(res$df)))
}
```
