---
title: "two_position_results"
author: "Andy Beck"
date: "2025-01-14"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Setup

```{r}
library(tidyverse)
library(ggpubfigs) # functions for color blind friendly palettes
library(reactable) # table output for html
library(knitr) # PDF table output
library(gridExtra)
```

### Paths

```{r}
base_dir <- "output/two_pos"
```

### Load Data

```{r}
deviance_df_load <- function(base_dir, pop, subtype){
  f_name <- paste0(base_dir, "/", pop, "/", subtype, ".csv")
  df <- read_csv(f_name, show_col_types = F, 
                 col_names = c("dev","singletons","controls","rp1","rp2")) %>%
    rowwise() %>%
    mutate(re = dev / (2 * (singletons + controls))) %>%
    ungroup()
  return(df)
}

resid_df_load <- function(base_dir, pop, subtype, p, q){
  f_name <- paste0(base_dir, "/resid/", pop, "/", subtype, "_p", p, "_q", q, ".csv")
  df <- read_csv(f_name, show_col_types = F)
  return(df)
}
```

### Plot data

```{r}
deviance_heatmap_1 <- function(df, max_rp = 20, palette = "Purples"){
  p <- df %>%
    filter(abs(rp1) <= max_rp, abs(rp2) <= max_rp) %>%
    ggplot(aes(x = rp2, y = rp1, fill = re)) +
    geom_tile() +
    xlab("Relative Position 2") +
    ylab("Relative Position 1") +
    scale_fill_distiller(palette = palette, direction = 1) +
    theme_classic() +
    labs(fill = "Relative Entropy") +
    theme(legend.position = "inside", legend.position.inside = c(0.3, 0.8),
          legend.background = element_rect(colour = 1))
  return(p)
}

deviance_heatmap_2 <- function(df, max_rp = 20, palette = "Purples"){
  p <- df %>%
    filter(abs(rp1) <= max_rp, abs(rp2) <= max_rp) %>%
    ggplot(aes(x = rp2, y = rp1, fill = re)) +
    geom_tile() +
    xlab("Relative Position 2") +
    ylab("Relative Position 1") +
    scale_fill_distiller(palette = palette, direction = 1) +
    theme_classic() +
    labs(fill = "Rel. Ent.")
  return(p)
}
```

### Main analysis function

```{r}
main_analysis <- function(base_dir, pop, subtype, max_rp = 20){
  df <- deviance_df_load(base_dir, pop, subtype)
  deviance_heatmap_1(df, max_rp) %>% print()
  reactable(df)
}
```

### RE plot style figure

```{r}
data <- deviance_df_load(base_dir, "ALL", "AT_GC")

# Function to extract legend
g_legend <- function(a.gplot){
  tmp <- ggplot_gtable(ggplot_build(a.gplot))
  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
  if (length(leg) >= 1) {
    return(tmp$grobs[[leg]])
  } else {
    return(NULL) # No legend
  }
}

get_sp_fig <- function(pop, subtype, rp_lim=10){
  base_dir = "output"
  single_pos_dir = paste0(base_dir, "/single_pos")
  control_control <- paste0(base_dir, "/single_pos_cc")
  
  f_name <- paste0(single_pos_dir, "/", pop, "/", subtype, ".csv")
  res_df <- read_csv(f_name, show_col_types = FALSE) %>%
    rowwise() %>%
    mutate(re = dev / (2 * (singletons+controls)))
  
  p <- res_df %>%
    filter(abs(offset) <= rp_lim) %>%
    ggplot(aes(x = offset, y = re)) +
    geom_point() +
    geom_line() +
    theme_classic() +
    theme(text = element_text(family = "Arial"),
          legend.position = "inside",
          legend.position.inside = c(.85, .8)) +
    xlab("") +
    ylab("")
  return(p)
}

re_ld_plot <- function(data, max_rp = 10, palette = "Purples"){
  data <- data %>%
    filter(abs(rp1) <= max_rp, abs(rp2) <= max_rp) %>%
    select(rp1, rp2, re) %>%
    as.data.frame()
  # add "0" values
  data <- data %>% 
    bind_rows(data.frame(rp1 = -(max_rp):-1, rp2 = 0, re = 0),
              data.frame(rp1 = 0, rp2 = 1:max_rp, re=0))
  # add "diagonal" values
  pos_names <- unique(c(data$rp1, data$rp2))
  data <- data %>%
    bind_rows(data.frame(rp1 = pos_names, rp2 = pos_names, re = 0)) %>%
    arrange(rp1, rp2)
  upper <- data %>%
    select(rp1, rp2, re) %>%
    pivot_wider(names_from = rp2, values_from = re)
  
  upper <- upper %>%
    select(-rp1) %>%
    as.matrix()
  
  rownames(upper) <- 1:dim(upper)[1]
  colnames(upper) <- 1:dim(upper)[2]
  
  lower <- t(upper)
  lower[is.na(lower)] <- 0
  upper[is.na(upper)] <- 0
  full <- upper + lower
  
  data <- tidyr::as_tibble(full)
  colnames(data) <- c(1:ncol(data))
  n <- length(data)
  
  # Tidy data, only taking unique pairs of data
  values <- data %>%
    dplyr::mutate(idx1 = c(1:nrow(data))) %>%
    tidyr::pivot_longer(!idx1, names_to = "idx2", values_to = "re") %>%
    dplyr::mutate(dplyr::across(idx2, as.double)) %>%
    dplyr::filter(!duplicated(paste(pmax(idx1, idx2), pmin(idx1, idx2), sep = "_"))) %>%
    tidyr::unite("id", idx1:idx2, remove = FALSE) %>%
    dplyr::mutate(diff = abs(idx2 - idx1))
  
  # Calculate coordinates for geom_polygon
  positions <- dplyr::bind_rows(values, values, values, values) %>%
    dplyr::group_by(diff, idx1) %>%
    dplyr::mutate(add_index1 = c(0, 1, 0, 1),
                  add_index2 = c(0, -1, 0, 1),
                  minus1_index = c(1, 1, 0, 1)) %>%
    dplyr::mutate(x = diff * 5 / n + 10 / n * (idx1 - minus1_index) + 5 / n * add_index1,
                  y = 5 - diff * 5 / n + 5 / n * add_index2) %>%
    dplyr::ungroup()
  
  positions$re[positions$re == 0] <- NA
  
  n <- dim(data)[1]
  
  # ggplot2
  p1 <- positions %>%
    ggplot2::ggplot(ggplot2::aes(x = x, y = y)) +
    ggplot2::geom_polygon(ggplot2::aes(fill = re, group = id)) +
    ggplot2::theme_void() +
    ggplot2::scale_fill_distiller(type = "seq", palette = palette, direction = 1, na.value = "grey90") +
    theme( 
        legend.position="bottom",
        legend.title = element_blank(),
        legend.direction = "horizontal",
        legend.key.size = unit(1, "cm"),  # Adjust key size (width)
        legend.spacing.x = unit(0.5, "cm")
    ) 
  legend_obj <- g_legend(p1)
  p2 <- p1 +
    theme(legend.position = "none") + 
    theme(aspect.ratio = (1 / n  + (n - 1) / (2 * n)))
  # grid.arrange(p2, legend_obj, ncol=1)
  # 
  # cowplot::plot_grid(
  #   p2,
  #   legend_obj,
  #   ncol = 1, rel_heights = c(6, 1)
  # )
  return(p2)
}

p1 <- re_ld_plot(data, max_rp = 10, palette = "Reds")
p2 <- get_sp_fig("ALL", "AT_GC") +
  theme(axis.line.y = element_blank(),     # Remove the y-axis line
        axis.text.y = element_blank(),     # Remove y-axis text labels
        axis.ticks.y = element_blank())

cowplot::plot_grid(
  p2,
  p1,
  ncol=1,
  align="v"
)
```


## Results

```{r}
max_rp <- 10
```


### A > C

```{r}
subtype <- "AT_CG"
```

#### ALL

```{r}
pop <- "ALL"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### AFR

```{r}
pop <- "AFR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### AMR

```{r}
pop <- "AMR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### EAS

```{r}
pop <- "EAS"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### EUR

```{r}
pop <- "EUR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### SAS

```{r}
pop <- "SAS"
main_analysis(base_dir, pop, subtype, max_rp)
```

### A > G

```{r}
subtype <- "AT_GC"
```

#### ALL

```{r}
pop <- "ALL"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### AFR

```{r}
pop <- "AFR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### AMR

```{r}
pop <- "AMR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### EAS

```{r}
pop <- "EAS"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### EUR

```{r}
pop <- "EUR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### SAS

```{r}
pop <- "SAS"
main_analysis(base_dir, pop, subtype, max_rp)
```

### A > T

```{r}
subtype <- "AT_TA"
```

#### ALL

```{r}
pop <- "ALL"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### AFR

```{r}
pop <- "AFR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### AMR

```{r}
pop <- "AMR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### EAS

```{r}
pop <- "EAS"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### EUR

```{r}
pop <- "EUR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### SAS

```{r}
pop <- "SAS"
main_analysis(base_dir, pop, subtype, max_rp)
```

### C > T

```{r}
subtype <- "GC_AT"
```

#### ALL

```{r}
pop <- "ALL"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### AFR

```{r}
pop <- "AFR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### AMR

```{r}
pop <- "AMR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### EAS

```{r}
pop <- "EAS"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### EUR

```{r}
pop <- "EUR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### SAS

```{r}
pop <- "SAS"
main_analysis(base_dir, pop, subtype, max_rp)
```

### C > A

```{r}
subtype <- "GC_TA"
```

#### ALL

```{r}
pop <- "ALL"
main_analysis(base_dir, pop, subtype, max_rp)
```

C to A has the largest overall RE observed across all subtypes for two position models. Let's take a look at the residuals for this position:

```{r}
df_resid <- resid_df_load(base_dir, pop, subtype, -1, 1)

df_resid %>%
  ggplot(aes(y = p1, x = p2, fill = res)) +
  geom_tile() +
  facet_grid(cols = vars(status)) +
  scale_fill_distiller(palette = "YlGn", direction = 1)

df_resid %>%
  select(p1, p2, status, n) %>%
  pivot_wider(id_cols = c("p1", "p2"), names_from = status, values_from = n) %>%
  mutate(p_s = singletons / sum(singletons),
         p_c = controls / sum(controls)) %>%
  mutate(p_diff = p_s - p_c) %>%
  ggplot(aes(y = p1, x = p2, fill = p_diff)) +
  geom_tile() +
  scale_fill_distiller(palette = "YlGn", direction = 1)
```


#### AFR

```{r}
pop <- "AFR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### AMR

```{r}
pop <- "AMR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### EAS

```{r}
pop <- "EAS"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### EUR

```{r}
pop <- "EUR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### SAS

```{r}
pop <- "SAS"
main_analysis(base_dir, pop, subtype, max_rp)
```

### C > G

```{r}
subtype <- "GC_CG"
```

#### ALL

```{r}
pop <- "ALL"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### AFR

```{r}
pop <- "AFR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### AMR

```{r}
pop <- "AMR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### EAS

```{r}
pop <- "EAS"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### EUR

```{r}
pop <- "EUR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### SAS

```{r}
pop <- "SAS"
main_analysis(base_dir, pop, subtype, max_rp)
```

### CpG > TpG

```{r}
subtype <- "cpg_GC_AT"
```

#### ALL

```{r}
pop <- "ALL"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### AFR

```{r}
pop <- "AFR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### AMR

```{r}
pop <- "AMR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### EAS

```{r}
pop <- "EAS"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### EUR

```{r}
pop <- "EUR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### SAS

```{r}
pop <- "SAS"
main_analysis(base_dir, pop, subtype, max_rp)
```


### CpG > ApG

```{r}
subtype <- "cpg_GC_TA"
```

#### ALL

```{r}
pop <- "ALL"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### AFR

```{r}
pop <- "AFR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### AMR

```{r}
pop <- "AMR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### EAS

```{r}
pop <- "EAS"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### EUR

```{r}
pop <- "EUR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### SAS

```{r}
pop <- "SAS"
main_analysis(base_dir, pop, subtype, max_rp)
```


### CpG > GpG

```{r}
subtype <- "cpg_GC_CG"
```

#### ALL

```{r}
pop <- "ALL"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### AFR

```{r}
pop <- "AFR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### AMR

```{r}
pop <- "AMR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### EAS

```{r}
pop <- "EAS"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### EUR

```{r}
pop <- "EUR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### SAS

```{r}
pop <- "SAS"
main_analysis(base_dir, pop, subtype, max_rp)
```

