---
title: "two_position_results"
author: "Andy Beck"
date: "2025-01-14"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Setup

```{r}
library(tidyverse)
library(ggpubfigs) # functions for color blind friendly palettes
library(reactable) # table output for html
library(knitr) # PDF table output
```

### Paths

```{r}
base_dir <- "output/two_pos"
```

### Load Data

```{r}
deviance_df_load <- function(base_dir, pop, subtype){
  f_name <- paste0(base_dir, "/", pop, "/", subtype, ".csv")
  df <- read_csv(f_name, show_col_types = F, 
                 col_names = c("dev","singletons","controls","rp1","rp2")) %>%
    rowwise() %>%
    mutate(re = dev / (2 * (singletons + controls))) %>%
    ungroup()
  return(df)
}

resid_df_load <- function(base_dir, pop, subtype, p, q){
  f_name <- paste0(base_dir, "/resid/", pop, "/", subtype, "_p", p, "_q", q, ".csv")
  df <- read_csv(f_name, show_col_types = F)
  return(df)
}
```

### Plot data

```{r}
deviance_heatmap_1 <- function(df, max_rp = 20){
  p <- df %>%
    filter(abs(rp1) <= max_rp, abs(rp2) <= max_rp) %>%
    ggplot(aes(x = rp2, y = rp1, fill = re)) +
    geom_tile() +
    xlab("Relative Position 2") +
    ylab("Relative Position 1") +
    scale_fill_distiller(palette = "Purples", direction = 1) +
    theme_classic() +
    labs(fill = "Relative Entropy") +
    theme(legend.position = "inside", legend.position.inside = c(0.3, 0.8),
          legend.background = element_rect(colour = 1))
  return(p)
}

deviance_heatmap_2 <- function(df, max_rp = 20){
  p <- df %>%
    filter(abs(rp1) <= max_rp, abs(rp2) <= max_rp) %>%
    ggplot(aes(x = rp2, y = rp1, fill = re)) +
    geom_tile() +
    xlab("Relative Position 2") +
    ylab("Relative Position 1") +
    scale_fill_distiller(palette = "Purples", direction = 1) +
    theme_classic() +
    labs(fill = "Rel. Ent.")
  return(p)
}
```

### Main analysis function

```{r}
main_analysis <- function(base_dir, pop, subtype, max_rp = 20){
  df <- deviance_df_load(base_dir, pop, subtype)
  deviance_heatmap_1(df, max_rp) %>% print()
  reactable(df)
}
```

## Results

```{r}
max_rp <- 20
```


### A > C

```{r}
subtype <- "AT_CG"
```

#### ALL

```{r}
pop <- "ALL"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### AFR

```{r}
pop <- "AFR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### AMR

```{r}
pop <- "AMR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### EAS

```{r}
pop <- "EAS"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### EUR

```{r}
pop <- "EUR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### SAS

```{r}
pop <- "SAS"
main_analysis(base_dir, pop, subtype, max_rp)
```

### A > G

```{r}
subtype <- "AT_GC"
```

#### ALL

```{r}
pop <- "ALL"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### AFR

```{r}
pop <- "AFR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### AMR

```{r}
pop <- "AMR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### EAS

```{r}
pop <- "EAS"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### EUR

```{r}
pop <- "EUR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### SAS

```{r}
pop <- "SAS"
main_analysis(base_dir, pop, subtype, max_rp)
```

### A > T

```{r}
subtype <- "AT_TA"
```

#### ALL

```{r}
pop <- "ALL"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### AFR

```{r}
pop <- "AFR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### AMR

```{r}
pop <- "AMR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### EAS

```{r}
pop <- "EAS"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### EUR

```{r}
pop <- "EUR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### SAS

```{r}
pop <- "SAS"
main_analysis(base_dir, pop, subtype, max_rp)
```

### C > T

```{r}
subtype <- "GC_AT"
```

#### ALL

```{r}
pop <- "ALL"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### AFR

```{r}
pop <- "AFR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### AMR

```{r}
pop <- "AMR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### EAS

```{r}
pop <- "EAS"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### EUR

```{r}
pop <- "EUR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### SAS

```{r}
pop <- "SAS"
main_analysis(base_dir, pop, subtype, max_rp)
```

### C > A

```{r}
subtype <- "GC_TA"
```

#### ALL

```{r}
pop <- "ALL"
main_analysis(base_dir, pop, subtype, max_rp)
```

C to A has the largest overall RE observed across all subtypes for two position models. Let's take a look at the residuals for this position:

```{r}
df_resid <- resid_df_load(base_dir, pop, subtype, -1, 1)

df_resid %>%
  ggplot(aes(y = p1, x = p2, fill = res)) +
  geom_tile() +
  facet_grid(cols = vars(status)) +
  scale_fill_distiller(palette = "YlGn", direction = 1)

df_resid %>%
  select(p1, p2, status, n) %>%
  pivot_wider(id_cols = c("p1", "p2"), names_from = status, values_from = n) %>%
  mutate(p_s = singletons / sum(singletons),
         p_c = controls / sum(controls)) %>%
  mutate(p_diff = p_s - p_c) %>%
  ggplot(aes(y = p1, x = p2, fill = p_diff)) +
  geom_tile() +
  scale_fill_distiller(palette = "YlGn", direction = 1)
```


#### AFR

```{r}
pop <- "AFR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### AMR

```{r}
pop <- "AMR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### EAS

```{r}
pop <- "EAS"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### EUR

```{r}
pop <- "EUR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### SAS

```{r}
pop <- "SAS"
main_analysis(base_dir, pop, subtype, max_rp)
```

### C > G

```{r}
subtype <- "GC_CG"
```

#### ALL

```{r}
pop <- "ALL"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### AFR

```{r}
pop <- "AFR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### AMR

```{r}
pop <- "AMR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### EAS

```{r}
pop <- "EAS"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### EUR

```{r}
pop <- "EUR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### SAS

```{r}
pop <- "SAS"
main_analysis(base_dir, pop, subtype, max_rp)
```

### CpG > TpG

```{r}
subtype <- "cpg_GC_AT"
```

#### ALL

```{r}
pop <- "ALL"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### AFR

```{r}
pop <- "AFR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### AMR

```{r}
pop <- "AMR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### EAS

```{r}
pop <- "EAS"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### EUR

```{r}
pop <- "EUR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### SAS

```{r}
pop <- "SAS"
main_analysis(base_dir, pop, subtype, max_rp)
```


### CpG > ApG

```{r}
subtype <- "cpg_GC_TA"
```

#### ALL

```{r}
pop <- "ALL"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### AFR

```{r}
pop <- "AFR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### AMR

```{r}
pop <- "AMR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### EAS

```{r}
pop <- "EAS"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### EUR

```{r}
pop <- "EUR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### SAS

```{r}
pop <- "SAS"
main_analysis(base_dir, pop, subtype, max_rp)
```


### CpG > GpG

```{r}
subtype <- "cpg_GC_CG"
```

#### ALL

```{r}
pop <- "ALL"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### AFR

```{r}
pop <- "AFR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### AMR

```{r}
pop <- "AMR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### EAS

```{r}
pop <- "EAS"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### EUR

```{r}
pop <- "EUR"
main_analysis(base_dir, pop, subtype, max_rp)
```

#### SAS

```{r}
pop <- "SAS"
main_analysis(base_dir, pop, subtype, max_rp)
```

